### 2021-03-18 submission/manuscript.rmd
transfer text from previous submission

---
title: "A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases"
output:
    bookdown::pdf_document2: default
    bookdown::word_document2:
        reference_docx: reference.docx
geometry: margin=1.0in
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left, pagewise]{lineno}
editor_options: 
  chunk_output_type: console
csl: jamia.csl
---

\newpage
# Abstract (247 words)
\singlespacing
**Aim:** One of the aims of the Observation Health Data Sciences and Informatics (OHDSI) initiative is population-level treatment effect estimation in large observational databases. Since treatment effects are well-known to vary across groups of patients with different baseline risk, we aimed to extend the OHDSI methods library with a framework for risk-based assessment of treatment effect heterogeneity.

**Materials and Methods:** The proposed framework consists of five steps: 1) definition of the problem, i.e. the population, the treatment, the comparator and the outcome(s) of interest; 2) identification of relevant databases; 3) development of a prediction model for the outcome(s) of interest; 4) estimation of propensity scores within strata of predicted risk and estimation of relative and absolute treatment effect within strata of predicted risk; 5) evaluation and presentation of results.

**Results:** We demonstrate our framework by evaluating heterogeneity of the effect of angiotensin-converting enzyme (ACE) inhibitors versus beta blockers on a set of 9 outcomes of interest across three observational databases. With increasing risk of acute myocardial infarction we observed increasing absolute benefits, i.e. from -0.03% to 0.54% in the lowest to highest risk groups. Cough-related absolute harms decreased from 4.1% to 2.6%.

**Conclusions:** The proposed framework may be useful for the evaluation of heterogeneity of treatment effect on observational data that are mapped to the OMOP Common Data Model. The proof of concept study demonstrates its feasibility in large observational data. Further insights may arise by application to safety and effectiveness questions across the global data network.

\newpage
\doublespacing
\linenumbers
# Introduction
Interest in understanding how a treatment’s effect varies across patients—a concept described as heterogeneity of treatment effects (HTE)—has been growing. This concept is central to the agenda for both personalized (or precision) medicine and comparative effectiveness research. More formally, HTE has been defined as non-random variability in the direction or magnitude of a treatment effect, in which the effect is measured using clinical outcomes. Usually, analyses focus on the relative scale, where treatment effects are assessed one at a time in patient subgroups defined from single covariates, an approach that suffers from low power and multiplicity issues. However, even with well-established constant relative effects, treatment benefit (or harm) may vary substantially on the absolute scale.

More recently, “predictive” HTE analyses have been described (and contrasted with “one-variable-at-a-time” subgroup analysis) as approaches that provide predictions of potential outcomes in a particular patient with one intervention versus an alternative, taking into account multiple relevant patient characteristics. One promising approach is “risk modeling”, in which treatment effects are estimated  in strata of predicted risk. Such a risk-based approach first stratifies patients according to baseline risk predictions, using either an existing or an internally developed risk prediction model. Then, relative and absolute treatment effects are estimated within risk strata.

While these approaches have generally been recommended for application to clinical trials, observational databases are also an appealing substrate. Observational healthcare databases, such as administrative claims and electronic health records, are already highly available for the analysis of pharmacoepidemiologic research questions. They are also often larger than many typical trials, providing excellent power for HTE analysis, and include heterogeneous populations. However, unlike trials, treatment effects are subject to confounding and the unique structure of different databases calls for database-specific analysis plans that are often not easily transportable.

The Observational Health Data Sciences and Informatics (OHDSI) collaborative has established an international network of data partners and researchers that aim to bring out the value of health data through large-scale analytics by mapping all available databases to the Observational Medical Outcomes Partnership (OMOP) Common Data Model (CDM). The common data structure enables analyses at a very large scale. For example, in a recent study, a large set of first-line treatments for hypertension was compared with respect to 55 outcomes in a network of databases, including 4.9 million patients from around the world.

We aimed to develop a framework for risk-based assessment of treatment effect heterogeneity in high-dimensional observational data. We implemented the framework using existing OHDSI methods for use in the OMOP-CDM, including the patient-level prediction framework and the population-level effect estimation framework based on new-user cohort design. As a proof-of-concept we analyzed heterogeneity of the effects of first-line hypertension treatment: we compared the effect of angiotensin converting enzyme (ACE) inhibitors to beta blockers on 9 outcomes across three different US claims databases.

# Methods
The proposed framework defines 5 distinct steps that enable a standardized approach for risk-based assessment of treatment effect heterogeneity for databases mapped to the OMOP-CDM. These are: 1) general definition of the research aim; 2) identification of the database within which the analyses will be performed; 3) a prediction step where internal or external prediction models are used to assign patient-level risk predictions; 4) an estimation step where absolute and relative treatment effects are estimated within risk strata; 5) presentation and evaluation of the results. A simple overview of the procedure can be seen in Figure XXXX.

## Step 1: General definition of the problem
The typical research aim is: "to compare the effect of treatment $T$ to a comparator treatment $C$ in patients with disease $D$ with respect to outcomes $O_1,\dots,O_n$". At least three cohorts are defined:

* A single treatment cohort ($T$) which includes patients with disease $D$ receiving the target treatment of interest. For example, a set of hypertension patients within a database that receive angiotensin-converting enzyme inhibitors, followed from the time of initiation until the time of censoring.
* A single comparator cohort ($C$) which includes patients with disease $D$ receiving the comparator (control) treatment. For example, a set of patients in a database that receive beta blockers, followed from the time of initiation until the time of censoring.
* One or more outcome cohorts ($O_1,\dots,O_n$) that contain patients developing the outcomes of interest. For example, the set of patients in a database that have at least one occurrence of acute myocardial infarction in their record.

## Step 2: Identification of the database
The aim of this step is the inclusion of databases that represent the patient population of interest. The inclusion of multiple databases potentially increases the generalizability of results. Furthermore, the cohorts should preferably have adequate sample size to ensure precise effect estimation, even within smaller risk strata.

## Step 3: Prediction
We adopt the standardized framework for the generation of patient-level prediction models using observational data that ensures adherence to existing guidelines. This prediction framework requires the definition of two essential cohorts: a target cohort and an outcome cohort.

To generate the target cohort we pool the already defined treatment cohort  and comparator cohort . However, for risk-based analysis of treatment effects it is necessary to avoid deferentially fitting the prediction model to patients across the treatment arm in order to avoid inducing spurious interactions. To do this, we developed the patient-level prediction model in the propensity score-matched subset of the population (1:1), where treatment assignment is well-balanced. The propensity scores are based on LASSO logistic regression for modeling the association between treatment assignment and all available demographics, drug exposures, diagnoses, measurements and medical procedures. Finally, we need to define the time horizon for which we aim to make predictions and we need to select the machine-learning algorithm we want to use to generate patient-level predictions. Currently, the available options are regularized logistic regression, random forest, gradient boosting machines, decision tree, naive Bayes, K-nearest neighbors, neural network and deep learning (convolutional neural networks, recurrent neural network and deep nets).

## Step 4: Estimation
We use the patient-level prediction model to divide the target population into a set of equally-sized risk strata, typically 4 risk quarters. Then, we estimate propensity scores within risk strata. These propensity scores are used when estimating treatment effects, either by matching of patients from different treatment cohorts, by stratification of patients into groups with similar propensity scores, or by weighing patients’ contribution to the estimation process. Within risk strata we estimate treatment effect both on the relative and the absolute scale. It is important to evaluate treatment effects in both scales, as effect cannot remain constant on both the relative and the absolute scale at the same time, assuming a non-zero treatment effect—treatment. Any appropriate method for the evaluation of relative and absolute treatment effects can be considered, as long as the this is done consistently in all risk strata.

## Step 5: Result presentation and evaluation
Our framework provides standardized output for each step of the analysis. The number of patients and person years by treatment arm along with the number of outcomes. A performance overview of the derived prediction models, including discrimination and calibration both in the propensity score matched subset, the entire population and separately for treated and comparator patients. This is rather relevant as the performance of the prediction models is directly related to our ability to single out patient subgroups where treatment may be highly beneficial or unsafe. Kent et al  demonstrated that the event rate and the discriminative ability of the prediction model can predict very well the distribution of predicted risk. The lower the event rate and the higher the c-statistic (given good calibration) result in high risk heterogeneity, thus making estimated average treatment effects uninformative. In this case, risk stratified analysis of HTE can be more effective in singling out patient subgroups that stand to benefit (or be harmed) most by treatment in question.

Propensity score distributions by treatment group and covariate balance plots for each risk stratum. Event rates, hazard ratios and absolute risk differences in risk strata for a selected outcome, both in tables and in graphs. Hazard ratios and absolute risk differences for all analyzed outcomes by risk stratum. Finally, shiny application can be generated to enable easy sharing of the results.

# Results
As a proof of concept, we focus on the comparison of angiotensin converting enzyme (ACE) inhibitors to beta blockers. ACE inhibitors are among the most common treatment classes for hypertension, with a well-established effectiveness. Beta blockers, even though initially widely used for the treatment of hypertension, more recent trials and meta-analyses have cast doubt on their relative effectiveness. As a result, newer US guidelines do not consider them for initial treatment for hypertension while in the EU guidelines combination with other antihypertensive treatments is recommended. However, another meta-analysis suggested that the efficacy profile of beta blockers is similar to other major treatment classes in younger hypertensive patients and, thus, countries like Canada still include them as a first-line candidate for the treatment.

## Step 1: General definition of the problem
We demonstrate the framework with the following research aim: "to compare the effect of ACE-inhibitors ($T$) to the effect of beta blockers ($C$) in patients with established hypertension ($D$) with respect to 9 outcomes ($O_1,\dots,O_9$)". The cohorts are:

* Treatment cohort: Patients receiving any drug within the ACE-inhibitor class with at least one year of follow-up before treatment start and a recorded hypertension diagnosis within that year.
* Comparator cohort: Patients receiving any drug within the beta blocker class with at least one year of follow-up before treatment start and a recorded hypertension diagnosis within that year.
* Outcome cohorts: We consider 3 main and 6 safety outcome cohorts. These are patients in the database with a diagnosis of: acute myocardial infarction (MI); hospitalization with heart failure; ischemic or hemorrhagic stroke (efficacy outcomes); hypokalemia; hyperkalemia; hypotension; angioedema; cough; abnormal weight gain (safety outcomes).

All cohort definitions can be found in the supplementary material.

## Step 2: Identification of the databases
We used the following databases:

* IBM MarketScan Medicare Supplemental Beneficiaries (MDCR): Represents health services of retirees (aged 65 or older) in the United States with primary or Medicare supplemental coverage through privately insured fee-for-service, point-of-service or capitated health plans. These data include adjudicated health insurance claims (e.g. inpatient, outpatient and outpatient pharmacy). Additionally, it captures laboratory tests for a subset of the covered lives.
* IBM MarketScan Medicaid (MDCD): Adjudicated US health insurance claims for Medicaid enrollees from multiple states. It includes hospital discharge diagnoses, outpatient diagnoses and procedures and outpatient pharmacy claims as well as ethnicity and Medicare eligibility.
* IBM MarketScan Commercial Claims and Encounters (CCAE): Data from individuals enrolled in US employer-sponsored   insurance   health   plans. The data includes adjudicated health insurance   claims (e.g. inpatient, outpatient, and outpatient pharmacy) as well as enrollment data from large employers and health plans who provide private healthcare coverage to employees, their spouses and dependents. Additionally, it captures laboratory tests for a subset of the covered lives.

## Step 3: Prediction
To obtain a target cohort for developing patient-level predictions we first merged the ACE-inhibitors cohort with the beta blockers cohort. We then matched patients in the ACE-inhibitor cohort to patients in the beta blockers cohort on the propensity score. We focused on the efficacy outcomes (acute MI, hospitalization with heart failure and hemorrhagic or ischemic stroke) for risk stratification of the patient population. In each database, for each main outcome, we developed a prediction model. We chose a time horizon of 2 years after inclusion into the target cohort. We developed the prediction models using LASSO logistic regression with 3-fold cross validation for hyper-parameter selection.

## Step 4: Estimation
We used patient-level predictions to stratify the patient population into 4 risk quarters. We estimated relative and absolute treatment effects for all 9 outcomes of interest. In order to estimate risk quarter-specific treatment effects we first estimated propensity scores within risk quarters. We then used the propensity scores to stratify patients into 5 strata.

## Step 5: Result presentation and evaluation
In the main manuscript we present the results of the analysis in the CCAE database with stratification based on risk predictions of acute MI. Results of analyses in the other databases and with other risk stratifications are included in the supplementary material.

For each outcome and in each risk stratum there were adequate numbers of patients (Table XXXX). The discriminative ability of the prediction models was moderate in the matched development subset (c-index 0.76 for acute MI ; 0.79 for hospitalization with heart failure;  0.74 for stroke; ), in the general population (c-index 0.74 for acute MI; 0.77 for hospitalization with heart failure; 0.73 for stroke), in the treatment cohort (c-index for acute MI it was 0.71, for hospitalization with heart failure was 0.76 and for stroke it was 0.72) and in the comparator cohort (c-index for acute MI it was 0.79 for hospitalization with heart failure was 0.79 and for stroke it was 0.75).

Relative treatment effects of ACE-inhibitors vs beta blockers increased (hazard ratios decreased) with increasing acute MI risk, resulting in more pronounced increases of absolute treatment effects (ARD) with increasing acute MI risk (Figure XXXX). Patients in the low risk quarter did not receive absolute treatment benefit (ARD -0.03%) while absolute risk was 0.54% lower (95% confidence interval 0.36%—0.71%) for patients in the high-risk quarter. In contrast, the absolute and relative effects of ACE-inhibitors on safety outcomes (e.g. cough and angioedema) are approximately constant or even slightly decreasing with increasing acute MI risk (Figure XXXX and XXXX). Similar results were observed in the other two databases (see supplementary material).

This example nicely illustrates heterogeneity of absolute treatment effects, i.e. differences in absolute benefits and harms of ACE-inhibitors vs beta blockers for patients with different baseline risk. The results suggest that treatment with ACE-inhibitors, compared to treatment with beta blockers, may be focused on the higher risk patients, in whom the benefits outweigh the harms. However, treatment with beta blockers may be a viable option in lower risk patients, in whom the benefit-harm tradeoff is in favor of beta blockers. This is in accordance with earlier findings that beta blockers should be considered as first-line treatment for younger hypertensive patients. More thorough evaluation of these results is required in future research.

The results of the analyses performed can be accessed and assessed through a publicly available web application (https://data.ohdsi.org/AceBeta9Outcomes).

# Discussion
We developed a framework for the assessment of heterogeneity of treatment effect in large observational databases using a risk modeling approach. The framework is implemented in an open source R-package in the OHDSI methods library (https://github.com/OHDSI/RiskStratifiedEstimation). As a proof-of-concept, we used our framework to evaluate heterogeneity of the effect of treatment with ACE-inhibitors compared to beta blockers on 3 efficacy and 6 safety outcomes.

In recent years several methods for the evaluation of treatment effect heterogeneity have been developed in the setting of RCTs. However, low power and restricted prior knowledge on the mechanisms of variation in treatment effect are often inherent in RCTs, which are often adequately powered only for the analysis of the primary outcome. Observational databases contain a large amount of information on treatment assignment and outcomes of interest, while also capturing key patient characteristics. Our framework provides a standardized approach that can be used to leverage available information from these data sources, allowing for large-scale assessment of treatment effect heterogeneity. Multiple outcomes can be evaluated in patient subgroups of similar baseline outcome risk, where different outcome risk stratification schemes can be considered. The standardized nature of the framework enables transportability to multiple databases, provided that they are mapped to the OMOP-CDM.

Recently, guidelines on the application of risk modeling approaches for the assessment of heterogeneity of treatment effect in RCT settings have been proposed. Our framework aims to translate these guidelines to the observational setting while also providing a toolset for its implementation. Several considerations need to be made. First, estimates may be biased due to the observational nature of the data. We attempt to account for potential confounding by estimating propensity scores within strata of predicted risk. These scores are estimated using regularized logistic regression on a large set of pre-defined covariates. This specific approach gave accurate results in extensive simulation studies. However, such approaches do not account for unobserved confounding. Several sensitivity analyses have been proposed in the literature for measuring the robustness of results in the presence of unobserved confounding. Another approach is to calibrate estimates and confidence intervals based on a large set of negative controls. Negative controls are treatment-outcome pairs for which a null effect has been established. Estimating these effects within available data provides an approximation of the null distribution that can be used to empirically recalibrate effect estimates. Future work may extend our framework with this type of analyses.

Our method provides a risk-stratified assessment of treatment effect heterogeneity. However, even though stratification can provide a rough guide for clinical interpretation, it is not appropriate to guide clinical practice, where decisions need to be made at the individual level. Presentation of treatment effects as a continuous function of risk would be more helpful, but is methodologically challenging. Future research is necessary for the development of methods for continuous risk-based assessment of HTE.

Externally derived prediction models are preferred for analyzing treatment effect heterogeneity. Such external models should be well transportable. In the absence of such prediction models, simulations of RCTs have shown that internally derived models can be used to provide unbiased estimates of treatment effect across the spectrum of baseline risk. However, in observational databases treatment arms may significantly differ in sample size. Because the prediction model will possibly better fit to the larger treatment arm, this may introduce spurious treatment-covariate interactions in the prediction model, leading to sub-optimal risk stratification. As a remedy, we first match the patients in the treatment and the comparator cohorts on the basis of propensity scores. Additionally, we propose to assess model performance in the separate treatment arms to evaluate its aptness for risk stratification.

Our contribution is a translation of the PATH statement principles to the OHDSI methods library. Our methods encourage open science as it requires clear definition of the research questions translated into clear and reproducible cohort definitions that can easily be shared among researchers. Our R-package provides a standardized stepwise procedure for the assessment of HTE. This enables source code to be easily shared and evaluated. The results of these analyses can be reproduced in a straightforward manner. Researchers with access to different databases mapped to OMOP-CDM can also very easily extend their overall analyses with risk-based assessment of treatment effect heterogeneity. This enables collaboration among multiple sites with access to different patient populations. We propose that the framework is implemented any time treatment effect estimation in high-dimensional observational data is undertaken.

Recently, disease risk scores have been explored as an alternative to propensity scores for balancing covariates. In our method, the objective of risk stratification is not balancing, but assessing the variation of treatment effects on multiple outcomes across patients with different levels of baseline risk. Although using the same risk model for balancing and risk-based HTE analysis may sound attractive, we note that our method only uses one risk model for stratification and one propensity score model for balancing, while separate disease risk score models would be required to analyze treatment effects for each of the multiple outcomes.

In conclusion, the proof-of-concept study demonstrates the feasibility of our framework for risk-based assessment of treatment effect heterogeneity in large observational data. The standardized framework is easily applicable and highly informative whenever treatment effect estimation in high-dimensional observational data is of interest. Our framework is a supplement to the population-level effect estimation framework developed within OHDSI and, in the presence of an adequately discriminating prediction model, can be used to make the overall results more actionable for medical decision making.

### 2021-03-18 submission/manuscript.rmd
set line width for readability

@@
 \singlespacing 
-**Aim:** One of the aims of the Observation Health Data Sciences and Informatics (OHDSI) initiative is population-level treatment effect estimation in large observational databases.
-Since treatment effects are well-known to vary across groups of patients with different baseline risk, we aimed to extend the OHDSI methods library with a framework for risk-based assessment of treatment effect heterogeneity.
+**Aim:** One of the aims of the Observation Health Data Sciences
+and Informatics (OHDSI) initiative is population-level treatment effect
+estimation in large observational databases. Since treatment effects are
+well-known to vary across groups of patients with different baseline risk, we
+aimed to extend the OHDSI methods library with a framework for risk-based
+assessment of treatment effect heterogeneity.
 
-**Materials and Methods:** The proposed framework consists of five steps: 1) definition of the problem, i.e.
-the population, the treatment, the comparator and the outcome(s) of interest; 2) identification of relevant databases; 3) development of a prediction model for the outcome(s) of interest; 4) estimation of propensity scores within strata of predicted risk and estimation of relative and absolute treatment effect within strata of predicted risk; 5) evaluation and presentation of results.
+**Materials and Methods:** The proposed framework consists of five steps: 1)
+definition of the problem, i.e. the population, the treatment, the comparator
+and the outcome(s) of interest; 2) identification of relevant databases; 3)
+development of a prediction model for the outcome(s) of interest; 4) estimation
+of propensity scores within strata of predicted risk and estimation of relative
+and absolute treatment effect within strata of predicted risk; 5) evaluation and
+presentation of results.
 
-**Results:** We demonstrate our framework by evaluating heterogeneity of the effect of angiotensin-converting enzyme (ACE) inhibitors versus beta blockers on a set of 9 outcomes of interest across three observational databases.
-With increasing risk of acute myocardial infarction we observed increasing absolute benefits, i.e.
-from -0.03% to 0.54% in the lowest to highest risk groups.
+**Results:** We demonstrate our framework by evaluating heterogeneity of the
+effect of angiotensin-converting enzyme (ACE) inhibitors versus beta blockers on
+a set of 9 outcomes of interest across three observational databases. With
+increasing risk of acute myocardial infarction we observed increasing absolute
+benefits, i.e. from -0.03% to 0.54% in the lowest to highest risk groups.
 Cough-related absolute harms decreased from 4.1% to 2.6%.
 
-**Conclusions:** The proposed framework may be useful for the evaluation of heterogeneity of treatment effect on observational data that are mapped to the OMOP Common Data Model.
-The proof of concept study demonstrates its feasibility in large observational data.
-Further insights may arise by application to safety and effectiveness questions across the global data network.
+**Conclusions:** The proposed framework may be useful for the evaluation of
+heterogeneity of treatment effect on observational data that are mapped to the
+OMOP Common Data Model. The proof of concept study demonstrates its feasibility
+in large observational data. Further insights may arise by application to safety
+and effectiveness questions across the global data network.
 
@@
 \linenumbers
+
 # Introduction
-Interest in understanding how a treatment’s effect varies across patients—a concept described as heterogeneity of treatment effects (HTE)—has been growing.
-This concept is central to the agenda for both personalized (or precision) medicine and comparative effectiveness research.
-More formally, HTE has been defined as non-random variability in the direction or magnitude of a treatment effect, in which the effect is measured using clinical outcomes.
-Usually, analyses focus on the relative scale, where treatment effects are assessed one at a time in patient subgroups defined from single covariates, an approach that suffers from low power and multiplicity issues.
-However, even with well-established constant relative effects, treatment benefit (or harm) may vary substantially on the absolute scale.
+Interest in understanding how a treatment’s effect varies across patients—a
+concept described as heterogeneity of treatment effects (HTE)—has been growing.
+This concept is central to the agenda for both personalized (or precision)
+medicine and comparative effectiveness research. More formally, HTE has been
+defined as non-random variability in the direction or magnitude of a treatment
+effect, in which the effect is measured using clinical outcomes. Usually,
+analyses focus on the relative scale, where treatment effects are assessed one
+at a time in patient subgroups defined from single covariates, an approach that
+suffers from low power and multiplicity issues. However, even with
+well-established constant relative effects, treatment benefit (or harm) may vary
+substantially on the absolute scale.
 
-More recently, “predictive” HTE analyses have been described (and contrasted with “one-variable-at-a-time” subgroup analysis) as approaches that provide predictions of potential outcomes in a particular patient with one intervention versus an alternative, taking into account multiple relevant patient characteristics.
-One promising approach is “risk modeling”, in which treatment effects are estimated  in strata of predicted risk.
-Such a risk-based approach first stratifies patients according to baseline risk predictions, using either an existing or an internally developed risk prediction model.
-Then, relative and absolute treatment effects are estimated within risk strata.
+More recently, “predictive” HTE analyses have been described (and contrasted
+with “one-variable-at-a-time” subgroup analysis) as approaches that provide
+predictions of potential outcomes in a particular patient with one intervention
+versus an alternative, taking into account multiple relevant patient
+characteristics. One promising approach is “risk modeling”, in which treatment
+effects are estimated  in strata of predicted risk. Such a risk-based approach
+first stratifies patients according to baseline risk predictions, using either
+an existing or an internally developed risk prediction model. Then, relative and
+absolute treatment effects are estimated within risk strata.
 
-While these approaches have generally been recommended for application to clinical trials, observational databases are also an appealing substrate.
-Observational healthcare databases, such as administrative claims and electronic health records, are already highly available for the analysis of pharmacoepidemiologic research questions.
-They are also often larger than many typical trials, providing excellent power for HTE analysis, and include heterogeneous populations.
-However, unlike trials, treatment effects are subject to confounding and the unique structure of different databases calls for database-specific analysis plans that are often not easily transportable.
+While these approaches have generally been recommended for application to
+clinical trials, observational databases are also an appealing substrate.
+Observational healthcare databases, such as administrative claims and electronic
+health records, are already highly available for the analysis of
+pharmacoepidemiologic research questions. They are also often larger than many
+typical trials, providing excellent power for HTE analysis, and include
+heterogeneous populations. However, unlike trials, treatment effects are subject
+to confounding and the unique structure of different databases calls for
+database-specific analysis plans that are often not easily transportable.
 
-The Observational Health Data Sciences and Informatics (OHDSI) collaborative has established an international network of data partners and researchers that aim to bring out the value of health data through large-scale analytics by mapping all available databases to the Observational Medical Outcomes Partnership (OMOP) Common Data Model (CDM).
-The common data structure enables analyses at a very large scale.
-For example, in a recent study, a large set of first-line treatments for hypertension was compared with respect to 55 outcomes in a network of databases, including 4.9 million patients from around the world.
+The Observational Health Data Sciences and Informatics (OHDSI) collaborative has
+established an international network of data partners and researchers that aim
+to bring out the value of health data through large-scale analytics by mapping
+all available databases to the Observational Medical Outcomes Partnership (OMOP)
+Common Data Model (CDM). The common data structure enables analyses at a very
+large scale. For example, in a recent study, a large set of first-line
+treatments for hypertension was compared with respect to 55 outcomes in a
+network of databases, including 4.9 million patients from around the world.
 
-We aimed to develop a framework for risk-based assessment of treatment effect heterogeneity in high-dimensional observational data.
-We implemented the framework using existing OHDSI methods for use in the OMOP-CDM, including the patient-level prediction framework and the population-level effect estimation framework based on new-user cohort design.
-As a proof-of-concept we analyzed heterogeneity of the effects of first-line hypertension treatment: we compared the effect of angiotensin converting enzyme (ACE) inhibitors to beta blockers on 9 outcomes across three different US claims databases.
+We aimed to develop a framework for risk-based assessment of treatment effect
+heterogeneity in high-dimensional observational data. We implemented the
+framework using existing OHDSI methods for use in the OMOP-CDM, including the
+patient-level prediction framework and the population-level effect estimation
+framework based on new-user cohort design. As a proof-of-concept we analyzed
+heterogeneity of the effects of first-line hypertension treatment: we compared
+the effect of angiotensin converting enzyme (ACE) inhibitors to beta blockers on
+9 outcomes across three different US claims databases.
 
 # Methods
-The proposed framework defines 5 distinct steps that enable a standardized approach for risk-based assessment of treatment effect heterogeneity for databases mapped to the OMOP-CDM.
-These are: 1) general definition of the research aim; 2) identification of the database within which the analyses will be performed; 3) a prediction step where internal or external prediction models are used to assign patient-level risk predictions; 4) an estimation step where absolute and relative treatment effects are estimated within risk strata; 5) presentation and evaluation of the results.
-A simple overview of the procedure can be seen in Figure XXXX.
+The proposed framework defines 5 distinct steps that enable a standardized
+approach for risk-based assessment of treatment effect heterogeneity for
+databases mapped to the OMOP-CDM. These are: 1) general definition of the
+research aim; 2) identification of the database within which the analyses will
+be performed; 3) a prediction step where internal or external prediction models
+are used to assign patient-level risk predictions; 4) an estimation step where
+absolute and relative treatment effects are estimated within risk strata; 5)
+presentation and evaluation of the results. A simple overview of the procedure
+can be seen in Figure XXXX.
 
-## Step 1: General definition of the problem
-The typical research aim is: "to compare the effect of treatment $T$ to a comparator treatment $C$ in patients with disease $D$ with respect to outcomes $O_1,\dots,O_n$".
-At least three cohorts are defined:
+## Step 1: General definition of the problem The typical research aim is: "to
+compare the effect of treatment $T$ to a comparator treatment $C$ in patients
+with disease $D$ with respect to outcomes $O_1,\dots,O_n$". At least three
+cohorts are defined:
 
-* A single treatment cohort ($T$) which includes patients with disease $D$ receiving the target treatment of interest.
-For example, a set of hypertension patients within a database that receive angiotensin-converting enzyme inhibitors, followed from the time of initiation until the time of censoring.
-* A single comparator cohort ($C$) which includes patients with disease $D$ receiving the comparator (control) treatment.
-For example, a set of patients in a database that receive beta blockers, followed from the time of initiation until the time of censoring.
-* One or more outcome cohorts ($O_1,\dots,O_n$) that contain patients developing the outcomes of interest.
-For example, the set of patients in a database that have at least one occurrence of acute myocardial infarction in their record.
+* A single treatment cohort ($T$) which includes patients with disease $D$
+receiving the target treatment of interest. For example, a set of hypertension
+patients within a database that receive angiotensin-converting enzyme
+inhibitors, followed from the time of initiation until the time of censoring.
+* A single comparator cohort ($C$) which includes patients with disease $D$
+receiving the comparator (control) treatment. For example, a set of patients
+in a database that receive beta blockers, followed from the time of initiation
+until the time of censoring.
+* One or more outcome cohorts ($O_1,\dots,O_n$) that contain patients
+developing the outcomes of interest. For example, the
+set of patients in a database that have at least one occurrence of acute
+myocardial infarction in their record.
 
-## Step 2: Identification of the database
-The aim of this step is the inclusion of databases that represent the patient population of interest.
-The inclusion of multiple databases potentially increases the generalizability of results.
-Furthermore, the cohorts should preferably have adequate sample size to ensure precise effect estimation, even within smaller risk strata.
+## Step 2: Identification of the database The aim of this step is the inclusion
+of databases that represent the patient population of interest. The inclusion of
+multiple databases potentially increases the generalizability of results.
+Furthermore, the cohorts should preferably have adequate sample size to ensure
+precise effect estimation, even within smaller risk strata.
 
-## Step 3: Prediction
-We adopt the standardized framework for the generation of patient-level prediction models using observational data that ensures adherence to existing guidelines.
-This prediction framework requires the definition of two essential cohorts: a target cohort and an outcome cohort.
+## Step 3: Prediction We adopt the standardized framework for the generation of
+patient-level prediction models using observational data that ensures adherence
+to existing guidelines. This prediction framework requires the definition of two
+essential cohorts: a target cohort and an outcome cohort.
 
-To generate the target cohort we pool the already defined treatment cohort  and comparator cohort .
-However, for risk-based analysis of treatment effects it is necessary to avoid deferentially fitting the prediction model to patients across the treatment arm in order to avoid inducing spurious interactions.
-To do this, we developed the patient-level prediction model in the propensity score-matched subset of the population (1:1), where treatment assignment is well-balanced.
-The propensity scores are based on LASSO logistic regression for modeling the association between treatment assignment and all available demographics, drug exposures, diagnoses, measurements and medical procedures.
-Finally, we need to define the time horizon for which we aim to make predictions and we need to select the machine-learning algorithm we want to use to generate patient-level predictions.
-Currently, the available options are regularized logistic regression, random forest, gradient boosting machines, decision tree, naive Bayes, K-nearest neighbors, neural network and deep learning (convolutional neural networks, recurrent neural network and deep nets).
+To generate the target cohort we pool the already defined treatment cohort  and
+comparator cohort . However, for risk-based analysis of treatment effects it is
+necessary to avoid deferentially fitting the prediction model to patients across
+the treatment arm in order to avoid inducing spurious interactions. To do this,
+we developed the patient-level prediction model in the propensity score-matched
+subset of the population (1:1), where treatment assignment is well-balanced. The
+propensity scores are based on LASSO logistic regression for modeling the
+association between treatment assignment and all available demographics, drug
+exposures, diagnoses, measurements and medical procedures. Finally, we need to
+define the time horizon for which we aim to make predictions and we need to
+select the machine-learning algorithm we want to use to generate patient-level
+predictions. Currently, the available options are regularized logistic
+regression, random forest, gradient boosting machines, decision tree, naive
+Bayes, K-nearest neighbors, neural network and deep learning (convolutional
+neural networks, recurrent neural network and deep nets).
 
-## Step 4: Estimation
-We use the patient-level prediction model to divide the target population into a set of equally-sized risk strata, typically 4 risk quarters.
-Then, we estimate propensity scores within risk strata.
-These propensity scores are used when estimating treatment effects, either by matching of patients from different treatment cohorts, by stratification of patients into groups with similar propensity scores, or by weighing patients’ contribution to the estimation process.
-Within risk strata we estimate treatment effect both on the relative and the absolute scale.
-It is important to evaluate treatment effects in both scales, as effect cannot remain constant on both the relative and the absolute scale at the same time, assuming a non-zero treatment effect—treatment.
-Any appropriate method for the evaluation of relative and absolute treatment effects can be considered, as long as the this is done consistently in all risk strata.
+## Step 4: Estimation We use the patient-level prediction model to divide the
+target population into a set of equally-sized risk strata, typically 4 risk
+quarters. Then, we estimate propensity scores within risk strata. These
+propensity scores are used when estimating treatment effects, either by matching
+of patients from different treatment cohorts, by stratification of patients into
+groups with similar propensity scores, or by weighing patients’ contribution to
+the estimation process. Within risk strata we estimate treatment effect both on
+the relative and the absolute scale. It is important to evaluate treatment
+effects in both scales, as effect cannot remain constant on both the relative
+and the absolute scale at the same time, assuming a non-zero treatment
+effect—treatment. Any appropriate method for the evaluation of relative and
+absolute treatment effects can be considered, as long as the this is done
+consistently in all risk strata.
 
-## Step 5: Result presentation and evaluation
-Our framework provides standardized output for each step of the analysis.
-The number of patients and person years by treatment arm along with the number of outcomes.
-A performance overview of the derived prediction models, including discrimination and calibration both in the propensity score matched subset, the entire population and separately for treated and comparator patients.
-This is rather relevant as the performance of the prediction models is directly related to our ability to single out patient subgroups where treatment may be highly beneficial or unsafe.
-Kent et al  demonstrated that the event rate and the discriminative ability of the prediction model can predict very well the distribution of predicted risk.
-The lower the event rate and the higher the c-statistic (given good calibration) result in high risk heterogeneity, thus making estimated average treatment effects uninformative.
-In this case, risk stratified analysis of HTE can be more effective in singling out patient subgroups that stand to benefit (or be harmed) most by treatment in question.
+## Step 5: Result presentation and evaluation Our framework provides
+standardized output for each step of the analysis. The number of patients and
+person years by treatment arm along with the number of outcomes. A performance
+overview of the derived prediction models, including discrimination and
+calibration both in the propensity score matched subset, the entire population
+and separately for treated and comparator patients. This is rather relevant as
+the performance of the prediction models is directly related to our ability to
+single out patient subgroups where treatment may be highly beneficial or unsafe.
+Kent et al  demonstrated that the event rate and the discriminative ability of
+the prediction model can predict very well the distribution of predicted risk.
+The lower the event rate and the higher the c-statistic (given good calibration)
+result in high risk heterogeneity, thus making estimated average treatment
+effects uninformative. In this case, risk stratified analysis of HTE can be more
+effective in singling out patient subgroups that stand to benefit (or be harmed)
+most by treatment in question.
 
-Propensity score distributions by treatment group and covariate balance plots for each risk stratum.
-Event rates, hazard ratios and absolute risk differences in risk strata for a selected outcome, both in tables and in graphs.
-Hazard ratios and absolute risk differences for all analyzed outcomes by risk stratum.
-Finally, shiny application can be generated to enable easy sharing of the results.
+Propensity score distributions by treatment group and covariate balance plots
+for each risk stratum. Event rates, hazard ratios and absolute risk differences
+in risk strata for a selected outcome, both in tables and in graphs. Hazard
+ratios and absolute risk differences for all analyzed outcomes by risk stratum.
+Finally, shiny application can be generated to enable easy sharing of the
+results.
 
 # Results
-As a proof of concept, we focus on the comparison of angiotensin converting enzyme (ACE) inhibitors to beta blockers.
-ACE inhibitors are among the most common treatment classes for hypertension, with a well-established effectiveness.
-Beta blockers, even though initially widely used for the treatment of hypertension, more recent trials and meta-analyses have cast doubt on their relative effectiveness.
-As a result, newer US guidelines do not consider them for initial treatment for hypertension while in the EU guidelines combination with other antihypertensive treatments is recommended.
-However, another meta-analysis suggested that the efficacy profile of beta blockers is similar to other major treatment classes in younger hypertensive patients and, thus, countries like Canada still include them as a first-line candidate for the treatment.
+As a proof of concept, we focus on the comparison of angiotensin converting
+enzyme (ACE) inhibitors to beta blockers. ACE inhibitors are among the most
+common treatment classes for hypertension, with a well-established
+effectiveness. Beta blockers, even though initially widely used for the
+treatment of hypertension, more recent trials and meta-analyses have cast doubt
+on their relative effectiveness. As a result, newer US guidelines do not
+consider them for initial treatment for hypertension while in the EU guidelines
+combination with other antihypertensive treatments is recommended. However,
+another meta-analysis suggested that the efficacy profile of beta blockers is
+similar to other major treatment classes in younger hypertensive patients and,
+thus, countries like Canada still include them as a first-line candidate for the
+treatment.
 
-## Step 1: General definition of the problem
-We demonstrate the framework with the following research aim: "to compare the effect of ACE-inhibitors ($T$) to the effect of beta blockers ($C$) in patients with established hypertension ($D$) with respect to 9 outcomes ($O_1,\dots,O_9$)".
-The cohorts are:
+## Step 1: General definition of the problem We demonstrate the framework with
+the following research aim: "to compare the effect of ACE-inhibitors ($T$) to
+the effect of beta blockers ($C$) in patients with established hypertension
+($D$) with respect to 9 outcomes ($O_1,\dots,O_9$)". The cohorts are:
 
-* Treatment cohort: Patients receiving any drug within the ACE-inhibitor class with at least one year of follow-up before treatment start and a recorded hypertension diagnosis within that year.
-* Comparator cohort: Patients receiving any drug within the beta blocker class with at least one year of follow-up before treatment start and a recorded hypertension diagnosis within that year.
+* Treatment cohort: Patients receiving any drug within the ACE-inhibitor class
+with at least one year of follow-up before treatment start and a recorded
+hypertension diagnosis within that year.
+* Comparator cohort: Patients
+receiving any drug within the beta blocker class with at least one year of
+follow-up before treatment start and a recorded hypertension diagnosis within
+that year.
 * Outcome cohorts: We consider 3 main and 6 safety outcome cohorts.
-These are patients in the database with a diagnosis of: acute myocardial infarction (MI); hospitalization with heart failure; ischemic or hemorrhagic stroke (efficacy outcomes); hypokalemia; hyperkalemia; hypotension; angioedema; cough; abnormal weight gain (safety outcomes).
+These are patients in the database with a diagnosis of: acute myocardial
+infarction (MI); hospitalization with heart failure; ischemic or hemorrhagic
+stroke (efficacy outcomes); hypokalemia; hyperkalemia; hypotension;
+angioedema; cough; abnormal weight gain (safety outcomes).
 
@@
 
-## Step 2: Identification of the databases
-We used the following databases:
+## Step 2: Identification of the databases We used the following databases:
 
-* IBM MarketScan Medicare Supplemental Beneficiaries (MDCR): Represents health services of retirees (aged 65 or older) in the United States with primary or Medicare supplemental coverage through privately insured fee-for-service, point-of-service or capitated health plans.
-These data include adjudicated health insurance claims (e.g.
-inpatient, outpatient and outpatient pharmacy).
+* IBM MarketScan Medicare Supplemental Beneficiaries (MDCR): Represents health
+services of retirees (aged 65 or older) in the United States with primary or
+Medicare supplemental coverage through privately insured fee-for-service,
+point-of-service or capitated health plans. These data include adjudicated
+health insurance claims (e.g. inpatient, outpatient and outpatient pharmacy).
 Additionally, it captures laboratory tests for a subset of the covered lives.
-* IBM MarketScan Medicaid (MDCD): Adjudicated US health insurance claims for Medicaid enrollees from multiple states.
-It includes hospital discharge diagnoses, outpatient diagnoses and procedures and outpatient pharmacy claims as well as ethnicity and Medicare eligibility.
-* IBM MarketScan Commercial Claims and Encounters (CCAE): Data from individuals enrolled in US employer-sponsored   insurance   health   plans.
-The data includes adjudicated health insurance   claims (e.g.
-inpatient, outpatient, and outpatient pharmacy) as well as enrollment data from large employers and health plans who provide private healthcare coverage to employees, their spouses and dependents.
-Additionally, it captures laboratory tests for a subset of the covered lives.
+* IBM MarketScan Medicaid (MDCD): Adjudicated US health insurance claims for
+Medicaid enrollees from multiple states. It includes hospital discharge
+diagnoses, outpatient diagnoses and procedures and outpatient pharmacy claims
+as well as ethnicity and Medicare eligibility.  IBM MarketScan Commercial
+* Claims and Encounters (CCAE): Data from individuals enrolled in US
+employer-sponsored   insurance   health   plans. The data includes adjudicated
+health insurance   claims (e.g. inpatient, outpatient, and outpatient
+pharmacy) as well as enrollment data from large employers and health plans who
+provide private healthcare coverage to employees, their spouses and
+dependents. Additionally, it captures laboratory tests for a subset of the
+covered lives.
 
-## Step 3: Prediction
-To obtain a target cohort for developing patient-level predictions we first merged the ACE-inhibitors cohort with the beta blockers cohort.
-We then matched patients in the ACE-inhibitor cohort to patients in the beta blockers cohort on the propensity score.
-We focused on the efficacy outcomes (acute MI, hospitalization with heart failure and hemorrhagic or ischemic stroke) for risk stratification of the patient population.
-In each database, for each main outcome, we developed a prediction model.
-We chose a time horizon of 2 years after inclusion into the target cohort.
-We developed the prediction models using LASSO logistic regression with 3-fold cross validation for hyper-parameter selection.
+## Step 3: Prediction To obtain a target cohort for developing patient-level
+predictions we first merged the ACE-inhibitors cohort with the beta blockers
+cohort. We then matched patients in the ACE-inhibitor cohort to patients in the
+beta blockers cohort on the propensity score. We focused on the efficacy
+outcomes (acute MI, hospitalization with heart failure and hemorrhagic or
+ischemic stroke) for risk stratification of the patient population. In each
+database, for each main outcome, we developed a prediction model. We chose a
+time horizon of 2 years after inclusion into the target cohort. We developed the
+prediction models using LASSO logistic regression with 3-fold cross validation
+for hyper-parameter selection.
 
-## Step 4: Estimation
-We used patient-level predictions to stratify the patient population into 4 risk quarters.
-We estimated relative and absolute treatment effects for all 9 outcomes of interest.
-In order to estimate risk quarter-specific treatment effects we first estimated propensity scores within risk quarters.
-We then used the propensity scores to stratify patients into 5 strata.
+## Step 4: Estimation We used patient-level predictions to stratify the patient
+population into 4 risk quarters. We estimated relative and absolute treatment
+effects for all 9 outcomes of interest. In order to estimate risk
+quarter-specific treatment effects we first estimated propensity scores within
+risk quarters. We then used the propensity scores to stratify patients into 5
+strata.
 
-## Step 5: Result presentation and evaluation
-In the main manuscript we present the results of the analysis in the CCAE database with stratification based on risk predictions of acute MI.
-Results of analyses in the other databases and with other risk stratifications are included in the supplementary material.
+## Step 5: Result presentation and evaluation In the main manuscript we present
+the results of the analysis in the CCAE database with stratification based on
+risk predictions of acute MI. Results of analyses in the other databases and
+with other risk stratifications are included in the supplementary material.
 
-For each outcome and in each risk stratum there were adequate numbers of patients (Table XXXX).
-The discriminative ability of the prediction models was moderate in the matched development subset (c-index 0.76 for acute MI ; 0.79 for hospitalization with heart failure;  0.74 for stroke; ), in the general population (c-index 0.74 for acute MI; 0.77 for hospitalization with heart failure; 0.73 for stroke), in the treatment cohort (c-index for acute MI it was 0.71, for hospitalization with heart failure was 0.76 and for stroke it was 0.72) and in the comparator cohort (c-index for acute MI it was 0.79 for hospitalization with heart failure was 0.79 and for stroke it was 0.75).
+For each outcome and in each risk stratum there were adequate numbers of
+patients (Table XXXX). The discriminative ability of the prediction models was
+moderate in the matched development subset (c-index 0.76 for acute MI ; 0.79 for
+hospitalization with heart failure;  0.74 for stroke; ), in the general
+population (c-index 0.74 for acute MI; 0.77 for hospitalization with heart
+failure; 0.73 for stroke), in the treatment cohort (c-index for acute MI it was
+0.71, for hospitalization with heart failure was 0.76 and for stroke it was
+0.72) and in the comparator cohort (c-index for acute MI it was 0.79 for
+hospitalization with heart failure was 0.79 and for stroke it was 0.75).
 
-Relative treatment effects of ACE-inhibitors vs beta blockers increased (hazard ratios decreased) with increasing acute MI risk, resulting in more pronounced increases of absolute treatment effects (ARD) with increasing acute MI risk (Figure XXXX).
-Patients in the low risk quarter did not receive absolute treatment benefit (ARD -0.03%) while absolute risk was 0.54% lower (95% confidence interval 0.36%—0.71%) for patients in the high-risk quarter.
-In contrast, the absolute and relative effects of ACE-inhibitors on safety outcomes (e.g.
-cough and angioedema) are approximately constant or even slightly decreasing with increasing acute MI risk (Figure XXXX and XXXX).
-Similar results were observed in the other two databases (see supplementary material).
+Relative treatment effects of ACE-inhibitors vs beta blockers increased (hazard
+ratios decreased) with increasing acute MI risk, resulting in more pronounced
+increases of absolute treatment effects (ARD) with increasing acute MI risk
+(Figure XXXX). Patients in the low risk quarter did not receive absolute
+treatment benefit (ARD -0.03%) while absolute risk was 0.54% lower (95%
+confidence interval 0.36%—0.71%) for patients in the high-risk quarter. In
+contrast, the absolute and relative effects of ACE-inhibitors on safety outcomes
+(e.g. cough and angioedema) are approximately constant or even slightly
+decreasing with increasing acute MI risk (Figure XXXX and XXXX). Similar results
+were observed in the other two databases (see supplementary material).
 
-This example nicely illustrates heterogeneity of absolute treatment effects, i.e.
-differences in absolute benefits and harms of ACE-inhibitors vs beta blockers for patients with different baseline risk.
-The results suggest that treatment with ACE-inhibitors, compared to treatment with beta blockers, may be focused on the higher risk patients, in whom the benefits outweigh the harms.
-However, treatment with beta blockers may be a viable option in lower risk patients, in whom the benefit-harm tradeoff is in favor of beta blockers.
-This is in accordance with earlier findings that beta blockers should be considered as first-line treatment for younger hypertensive patients.
-More thorough evaluation of these results is required in future research.
+This example nicely illustrates heterogeneity of absolute treatment effects,
+i.e. differences in absolute benefits and harms of ACE-inhibitors vs beta
+blockers for patients with different baseline risk. The results suggest that
+treatment with ACE-inhibitors, compared to treatment with beta blockers, may be
+focused on the higher risk patients, in whom the benefits outweigh the harms.
+However, treatment with beta blockers may be a viable option in lower risk
+patients, in whom the benefit-harm tradeoff is in favor of beta blockers. This
+is in accordance with earlier findings that beta blockers should be considered
+as first-line treatment for younger hypertensive patients. More thorough
+evaluation of these results is required in future research.
 
-The results of the analyses performed can be accessed and assessed through a publicly available web application (https://data.ohdsi.org/AceBeta9Outcomes).
+The results of the analyses performed can be accessed and assessed through a
+publicly available web application (https://data.ohdsi.org/AceBeta9Outcomes).
 
 # Discussion
-We developed a framework for the assessment of heterogeneity of treatment effect in large observational databases using a risk modeling approach.
-The framework is implemented in an open source R-package in the OHDSI methods library (https://github.com/OHDSI/RiskStratifiedEstimation).
-As a proof-of-concept, we used our framework to evaluate heterogeneity of the effect of treatment with ACE-inhibitors compared to beta blockers on 3 efficacy and 6 safety outcomes.
+We developed a framework for the assessment of heterogeneity of treatment effect
+in large observational databases using a risk modeling approach. The framework
+is implemented in an open source R-package in the OHDSI methods library
+(https://github.com/OHDSI/RiskStratifiedEstimation). As a proof-of-concept, we
+used our framework to evaluate heterogeneity of the effect of treatment with
+ACE-inhibitors compared to beta blockers on 3 efficacy and 6 safety outcomes.
 
-In recent years several methods for the evaluation of treatment effect heterogeneity have been developed in the setting of RCTs.
-However, low power and restricted prior knowledge on the mechanisms of variation in treatment effect are often inherent in RCTs, which are often adequately powered only for the analysis of the primary outcome.
-Observational databases contain a large amount of information on treatment assignment and outcomes of interest, while also capturing key patient characteristics.
-Our framework provides a standardized approach that can be used to leverage available information from these data sources, allowing for large-scale assessment of treatment effect heterogeneity.
-Multiple outcomes can be evaluated in patient subgroups of similar baseline outcome risk, where different outcome risk stratification schemes can be considered.
-The standardized nature of the framework enables transportability to multiple databases, provided that they are mapped to the OMOP-CDM.
+In recent years several methods for the evaluation of treatment effect
+heterogeneity have been developed in the setting of RCTs. However, low power and
+restricted prior knowledge on the mechanisms of variation in treatment effect
+are often inherent in RCTs, which are often adequately powered only for the
+analysis of the primary outcome. Observational databases contain a large amount
+of information on treatment assignment and outcomes of interest, while also
+capturing key patient characteristics. Our framework provides a standardized
+approach that can be used to leverage available information from these data
+sources, allowing for large-scale assessment of treatment effect heterogeneity.
+Multiple outcomes can be evaluated in patient subgroups of similar baseline
+outcome risk, where different outcome risk stratification schemes can be
+considered. The standardized nature of the framework enables transportability to
+multiple databases, provided that they are mapped to the OMOP-CDM.
 
-Recently, guidelines on the application of risk modeling approaches for the assessment of heterogeneity of treatment effect in RCT settings have been proposed.
-Our framework aims to translate these guidelines to the observational setting while also providing a toolset for its implementation.
-Several considerations need to be made.
-First, estimates may be biased due to the observational nature of the data.
-We attempt to account for potential confounding by estimating propensity scores within strata of predicted risk.
-These scores are estimated using regularized logistic regression on a large set of pre-defined covariates.
-This specific approach gave accurate results in extensive simulation studies.
-However, such approaches do not account for unobserved confounding.
-Several sensitivity analyses have been proposed in the literature for measuring the robustness of results in the presence of unobserved confounding.
-Another approach is to calibrate estimates and confidence intervals based on a large set of negative controls.
-Negative controls are treatment-outcome pairs for which a null effect has been established.
-Estimating these effects within available data provides an approximation of the null distribution that can be used to empirically recalibrate effect estimates.
+Recently, guidelines on the application of risk modeling approaches for the
+assessment of heterogeneity of treatment effect in RCT settings have been
+proposed. Our framework aims to translate these guidelines to the observational
+setting while also providing a toolset for its implementation. Several
+considerations need to be made. First, estimates may be biased due to the
+observational nature of the data. We attempt to account for potential
+confounding by estimating propensity scores within strata of predicted risk.
+These scores are estimated using regularized logistic regression on a large set
+of pre-defined covariates. This specific approach gave accurate results in
+extensive simulation studies. However, such approaches do not account for
+unobserved confounding. Several sensitivity analyses have been proposed in the
+literature for measuring the robustness of results in the presence of unobserved
+confounding. Another approach is to calibrate estimates and confidence intervals
+based on a large set of negative controls. Negative controls are
+treatment-outcome pairs for which a null effect has been established. Estimating
+these effects within available data provides an approximation of the null
+distribution that can be used to empirically recalibrate effect estimates.
 Future work may extend our framework with this type of analyses.
 
-Our method provides a risk-stratified assessment of treatment effect heterogeneity.
-However, even though stratification can provide a rough guide for clinical interpretation, it is not appropriate to guide clinical practice, where decisions need to be made at the individual level.
-Presentation of treatment effects as a continuous function of risk would be more helpful, but is methodologically challenging.
-Future research is necessary for the development of methods for continuous risk-based assessment of HTE.
+Our method provides a risk-stratified assessment of treatment effect
+heterogeneity. However, even though stratification can provide a rough guide for
+clinical interpretation, it is not appropriate to guide clinical practice, where
+decisions need to be made at the individual level. Presentation of treatment
+effects as a continuous function of risk would be more helpful, but is
+methodologically challenging. Future research is necessary for the development
+of methods for continuous risk-based assessment of HTE.
 
-Externally derived prediction models are preferred for analyzing treatment effect heterogeneity.
-Such external models should be well transportable.
-In the absence of such prediction models, simulations of RCTs have shown that internally derived models can be used to provide unbiased estimates of treatment effect across the spectrum of baseline risk.
-However, in observational databases treatment arms may significantly differ in sample size.
-Because the prediction model will possibly better fit to the larger treatment arm, this may introduce spurious treatment-covariate interactions in the prediction model, leading to sub-optimal risk stratification.
-As a remedy, we first match the patients in the treatment and the comparator cohorts on the basis of propensity scores.
-Additionally, we propose to assess model performance in the separate treatment arms to evaluate its aptness for risk stratification.
+Externally derived prediction models are preferred for analyzing treatment
+effect heterogeneity. Such external models should be well transportable. In the
+absence of such prediction models, simulations of RCTs have shown that
+internally derived models can be used to provide unbiased estimates of treatment
+effect across the spectrum of baseline risk. However, in observational databases
+treatment arms may significantly differ in sample size. Because the prediction
+model will possibly better fit to the larger treatment arm, this may introduce
+spurious treatment-covariate interactions in the prediction model, leading to
+sub-optimal risk stratification. As a remedy, we first match the patients in the
+treatment and the comparator cohorts on the basis of propensity scores.
+Additionally, we propose to assess model performance in the separate treatment
+arms to evaluate its aptness for risk stratification.
 
-Our contribution is a translation of the PATH statement principles to the OHDSI methods library.
-Our methods encourage open science as it requires clear definition of the research questions translated into clear and reproducible cohort definitions that can easily be shared among researchers.
-Our R-package provides a standardized stepwise procedure for the assessment of HTE.
-This enables source code to be easily shared and evaluated.
-The results of these analyses can be reproduced in a straightforward manner.
-Researchers with access to different databases mapped to OMOP-CDM can also very easily extend their overall analyses with risk-based assessment of treatment effect heterogeneity.
-This enables collaboration among multiple sites with access to different patient populations.
-We propose that the framework is implemented any time treatment effect estimation in high-dimensional observational data is undertaken.
+Our contribution is a translation of the PATH statement principles to the OHDSI
+methods library. Our methods encourage open science as it requires clear
+definition of the research questions translated into clear and reproducible
+cohort definitions that can easily be shared among researchers. Our R-package
+provides a standardized stepwise procedure for the assessment of HTE. This
+enables source code to be easily shared and evaluated. The results of these
+analyses can be reproduced in a straightforward manner. Researchers with access
+to different databases mapped to OMOP-CDM can also very easily extend their
+overall analyses with risk-based assessment of treatment effect heterogeneity.
+This enables collaboration among multiple sites with access to different patient
+populations. We propose that the framework is implemented any time treatment
+effect estimation in high-dimensional observational data is undertaken.
 
-Recently, disease risk scores have been explored as an alternative to propensity scores for balancing covariates.
-In our method, the objective of risk stratification is not balancing, but assessing the variation of treatment effects on multiple outcomes across patients with different levels of baseline risk.
-Although using the same risk model for balancing and risk-based HTE analysis may sound attractive, we note that our method only uses one risk model for stratification and one propensity score model for balancing, while separate disease risk score models would be required to analyze treatment effects for each of the multiple outcomes.
+Recently, disease risk scores have been explored as an alternative to propensity
+scores for balancing covariates. In our method, the objective of risk
+stratification is not balancing, but assessing the variation of treatment
+effects on multiple outcomes across patients with different levels of baseline
+risk. Although using the same risk model for balancing and risk-based HTE
+analysis may sound attractive, we note that our method only uses one risk model
+for stratification and one propensity score model for balancing, while separate
+disease risk score models would be required to analyze treatment effects for
+each of the multiple outcomes.
 
-In conclusion, the proof-of-concept study demonstrates the feasibility of our framework for risk-based assessment of treatment effect heterogeneity in large observational data.
-The standardized framework is easily applicable and highly informative whenever treatment effect estimation in high-dimensional observational data is of interest.
-Our framework is a supplement to the population-level effect estimation framework developed within OHDSI and, in the presence of an adequately discriminating prediction model, can be used to make the overall results more actionable for medical decision making.
+In conclusion, the proof-of-concept study demonstrates the feasibility of our
+framework for risk-based assessment of treatment effect heterogeneity in large
+observational data. The standardized framework is easily applicable and highly
+informative whenever treatment effect estimation in high-dimensional
+observational data is of interest. Our framework is a supplement to the
+population-level effect estimation framework developed within OHDSI and, in the
+presence of an adequately discriminating prediction model, can be used to make
+the overall results more actionable for medical decision making.

### 2021-04-16 submission/manuscript.rmd
fix titles
include references

@@
   chunk_output_type: console
+bibliography: references.bib
 csl: jamia.csl
@@
 defined as non-random variability in the direction or magnitude of a treatment
-effect, in which the effect is measured using clinical outcomes. Usually,
-analyses focus on the relative scale, where treatment effects are assessed one
-at a time in patient subgroups defined from single covariates, an approach that
-suffers from low power and multiplicity issues. However, even with
-well-established constant relative effects, treatment benefit (or harm) may vary
-substantially on the absolute scale.
+effect, in which the effect is measured using clinical outcomes
+[@KRAVITZ2004]. Usually, analyses focus on the relative scale, where treatment
+effects are assessed one at a time in patient subgroups defined from single
+covariates, an approach that suffers from low power and multiplicity issues
+[@Varadhan2013]. However, even with well-established constant relative
+effects, treatment benefit (or harm) may vary substantially on the absolute
+scale.
 
@@
 characteristics. One promising approach is “risk modeling”, in which treatment
-effects are estimated  in strata of predicted risk. Such a risk-based approach
-first stratifies patients according to baseline risk predictions, using either
-an existing or an internally developed risk prediction model. Then, relative and
-absolute treatment effects are estimated within risk strata.
+effects are estimated in strata of predicted risk [@Kent2010; @Kent2018]. Such
+a risk-based approach first stratifies patients according to baseline risk
+predictions, using either an existing or an internally developed risk prediction
+model [@Kent2010]. Then, relative and absolute treatment effects are estimated
+within risk strata.
 
@@
 health records, are already highly available for the analysis of
-pharmacoepidemiologic research questions. They are also often larger than many
-typical trials, providing excellent power for HTE analysis, and include
-heterogeneous populations. However, unlike trials, treatment effects are subject
-to confounding and the unique structure of different databases calls for
-database-specific analysis plans that are often not easily transportable.
+pharmacoepidemiologic research questions [@AdlerMilstein2017; @Dahabreh2014].
+They are also often larger than many typical trials, providing excellent power
+for HTE analysis, and include heterogeneous populations. However, unlike trials,
+treatment effects are subject to confounding and the unique structure of
+different databases calls for database-specific analysis plans that are often
+not easily transportable.
 
-The Observational Health Data Sciences and Informatics (OHDSI) collaborative has
-established an international network of data partners and researchers that aim
-to bring out the value of health data through large-scale analytics by mapping
-all available databases to the Observational Medical Outcomes Partnership (OMOP)
-Common Data Model (CDM). The common data structure enables analyses at a very
-large scale. For example, in a recent study, a large set of first-line
-treatments for hypertension was compared with respect to 55 outcomes in a
-network of databases, including 4.9 million patients from around the world.
+The Observational Health Data Sciences and Informatics (OHDSI) collaborative
+(!!REF) has established an international network of data partners and
+researchers that aim to bring out the value of health data through large-scale
+analytics by mapping all available databases to the Observational Medical
+Outcomes Partnership (OMOP) Common Data Model (CDM) [@Overhage2012]. The
+common data structure enables analyses at a very large scale. For example, in a
+recent study [@Suchard2019], a large set of first-line treatments for
+hypertension was compared with respect to 55 outcomes in a network of databases,
+including 4.9 million patients from around the world.
 
@@
 framework using existing OHDSI methods for use in the OMOP-CDM, including the
-patient-level prediction framework and the population-level effect estimation
-framework based on new-user cohort design. As a proof-of-concept we analyzed
-heterogeneity of the effects of first-line hypertension treatment: we compared
-the effect of angiotensin converting enzyme (ACE) inhibitors to beta blockers on
-9 outcomes across three different US claims databases.
+patient-level prediction framework [@Reps2018] and the population-level effect
+estimation framework [@Ryan2013] based on new-user cohort design. As a
+proof-of-concept we analyzed heterogeneity of the effects of first-line
+hypertension treatment: we compared the effect of angiotensin converting enzyme
+(ACE) inhibitors to beta blockers on 9 outcomes across three different US claims
+databases.
 
@@
 
-## Step 1: General definition of the problem The typical research aim is: "to
-compare the effect of treatment $T$ to a comparator treatment $C$ in patients
-with disease $D$ with respect to outcomes $O_1,\dots,O_n$". At least three
-cohorts are defined:
+## Step 1: General definition of the problem 
+The typical research aim is: "to compare the effect of treatment $T$ to a
+comparator treatment $C$ in patients with disease $D$ with respect to outcomes
+$O_1,\dots,O_n$". At least three cohorts are defined:
 
@@
 
-## Step 2: Identification of the database The aim of this step is the inclusion
-of databases that represent the patient population of interest. The inclusion of
-multiple databases potentially increases the generalizability of results.
-Furthermore, the cohorts should preferably have adequate sample size to ensure
-precise effect estimation, even within smaller risk strata.
+## Step 2: Identification of the database 
+The aim of this step is the inclusion of databases that represent the patient
+population of interest. The inclusion of multiple databases potentially
+increases the generalizability of results. Furthermore, the cohorts should
+preferably have adequate sample size to ensure precise effect estimation, even
+within smaller risk strata.
 
-## Step 3: Prediction We adopt the standardized framework for the generation of
-patient-level prediction models using observational data that ensures adherence
-to existing guidelines. This prediction framework requires the definition of two
-essential cohorts: a target cohort and an outcome cohort.
+## Step 3: Prediction 
+We adopt the standardized framework for the generation of patient-level
+prediction models using observational data that ensures adherence to existing
+guidelines [@Collins2015; @Moons2015]. This prediction framework requires the
+definition of two essential cohorts: a target cohort and an outcome cohort.
 
@@
 necessary to avoid deferentially fitting the prediction model to patients across
-the treatment arm in order to avoid inducing spurious interactions. To do this,
-we developed the patient-level prediction model in the propensity score-matched
-subset of the population (1:1), where treatment assignment is well-balanced. The
-propensity scores are based on LASSO logistic regression for modeling the
-association between treatment assignment and all available demographics, drug
-exposures, diagnoses, measurements and medical procedures. Finally, we need to
-define the time horizon for which we aim to make predictions and we need to
-select the machine-learning algorithm we want to use to generate patient-level
-predictions. Currently, the available options are regularized logistic
-regression, random forest, gradient boosting machines, decision tree, naive
-Bayes, K-nearest neighbors, neural network and deep learning (convolutional
-neural networks, recurrent neural network and deep nets).
+the treatment arm in order to avoid inducing spurious interactions [@Burke2014;
+@vanKlaveren2019]. To do this, we developed the patient-level prediction model
+in the propensity score-matched subset of the population (1:1), where treatment
+assignment is well-balanced. The propensity scores are based on LASSO logistic
+regression for modeling the association between treatment assignment and all
+available demographics, drug exposures, diagnoses, measurements and medical
+procedures. Finally, we need to define the time horizon for which we aim to make
+predictions and we need to select the machine-learning algorithm we want to use
+to generate patient-level predictions. Currently, the available options are
+regularized logistic regression, random forest, gradient boosting machines,
+decision tree, naive Bayes, K-nearest neighbors, neural network and deep
+learning (convolutional neural networks, recurrent neural network and deep
+nets).
 
-## Step 4: Estimation We use the patient-level prediction model to divide the
-target population into a set of equally-sized risk strata, typically 4 risk
-quarters. Then, we estimate propensity scores within risk strata. These
-propensity scores are used when estimating treatment effects, either by matching
-of patients from different treatment cohorts, by stratification of patients into
-groups with similar propensity scores, or by weighing patients’ contribution to
-the estimation process. Within risk strata we estimate treatment effect both on
-the relative and the absolute scale. It is important to evaluate treatment
-effects in both scales, as effect cannot remain constant on both the relative
-and the absolute scale at the same time, assuming a non-zero treatment
+## Step 4: Estimation 
+We use the patient-level prediction model to divide the target population into
+a set of equally-sized risk strata, typically 4 risk quarters. Then, we
+estimate propensity scores within risk strata. These propensity scores are used
+when estimating treatment effects, either by matching of patients from
+different treatment cohorts, by stratification of patients into groups with
+similar propensity scores, or by weighing patients’ contribution to the
+estimation process. Within risk strata we estimate treatment effect both on the
+relative and the absolute scale. It is important to evaluate treatment effects
+in both scales, as effect cannot remain constant on both the relative and the
+absolute scale at the same time, assuming a non-zero treatment
 effect—treatment. Any appropriate method for the evaluation of relative and
@@
 
-## Step 5: Result presentation and evaluation Our framework provides
-standardized output for each step of the analysis. The number of patients and
-person years by treatment arm along with the number of outcomes. A performance
-overview of the derived prediction models, including discrimination and
-calibration both in the propensity score matched subset, the entire population
-and separately for treated and comparator patients. This is rather relevant as
-the performance of the prediction models is directly related to our ability to
-single out patient subgroups where treatment may be highly beneficial or unsafe.
-Kent et al  demonstrated that the event rate and the discriminative ability of
-the prediction model can predict very well the distribution of predicted risk.
-The lower the event rate and the higher the c-statistic (given good calibration)
-result in high risk heterogeneity, thus making estimated average treatment
-effects uninformative. In this case, risk stratified analysis of HTE can be more
+## Step 5: Result presentation and evaluation 
+Our framework provides standardized output for each step of the analysis. The
+number of patients and person years by treatment arm along with the number of
+outcomes. A performance overview of the derived prediction models, including
+discrimination and calibration both in the propensity score matched subset, the
+entire population and separately for treated and comparator patients. This is
+rather relevant as the performance of the prediction models is directly related
+to our ability to single out patient subgroups where treatment may be highly
+beneficial or unsafe.  Kent et al [@Kent2016] demonstrated that the event rate and the
+discriminative ability of the prediction model can predict very well the
+distribution of predicted risk. The lower the event rate and the higher the
+c-statistic (given good calibration) result in high risk heterogeneity, thus
+making estimated average treatment effects uninformative. In this case, risk
+stratified analysis of HTE can be more
 effective in singling out patient subgroups that stand to benefit (or be harmed)
@@
 treatment of hypertension, more recent trials and meta-analyses have cast doubt
-on their relative effectiveness. As a result, newer US guidelines do not
-consider them for initial treatment for hypertension while in the EU guidelines
-combination with other antihypertensive treatments is recommended. However,
-another meta-analysis suggested that the efficacy profile of beta blockers is
-similar to other major treatment classes in younger hypertensive patients and,
-thus, countries like Canada still include them as a first-line candidate for the
-treatment.
+on their relative effectiveness (!!REF). As a result, newer US guidelines do
+not consider them for initial treatment for hypertension while in the EU
+guidelines combination with other antihypertensive treatments is recommended
+[@Whelton2018; @Williams2018]. However, another meta-analysis suggested that
+the efficacy profile of beta blockers is similar to other major treatment
+classes in younger hypertensive patients and, thus, countries like Canada still
+include them as a first-line candidate for the treatment [@Khan2006;
+@Rabi2020].
 
-## Step 1: General definition of the problem We demonstrate the framework with
-the following research aim: "to compare the effect of ACE-inhibitors ($T$) to
-the effect of beta blockers ($C$) in patients with established hypertension
-($D$) with respect to 9 outcomes ($O_1,\dots,O_9$)". The cohorts are:
+## Step 1: General definition of the problem 
+We demonstrate the framework with the following research aim: "to compare the
+effect of ACE-inhibitors ($T$) to the effect of beta blockers ($C$) in patients
+with established hypertension ($D$) with respect to 9 outcomes
+($O_1,\dots,O_9$)". The cohorts are:
 
@@
 
-## Step 2: Identification of the databases We used the following databases:
+## Step 2: Identification of the databases 
+We used the following databases:
 
@@
 
-## Step 3: Prediction To obtain a target cohort for developing patient-level
+## Step 3: Prediction 
+To obtain a target cohort for developing patient-level
 predictions we first merged the ACE-inhibitors cohort with the beta blockers
@@
 
-## Step 4: Estimation We used patient-level predictions to stratify the patient
-population into 4 risk quarters. We estimated relative and absolute treatment
-effects for all 9 outcomes of interest. In order to estimate risk
-quarter-specific treatment effects we first estimated propensity scores within
-risk quarters. We then used the propensity scores to stratify patients into 5
-strata.
+## Step 4: Estimation 
+We used patient-level predictions to stratify the patient population into 4
+risk quarters. We estimated relative and absolute treatment effects for all 9
+outcomes of interest. In order to estimate risk quarter-specific treatment
+effects we first estimated propensity scores within risk quarters. We then used
+the propensity scores to stratify patients into 5 strata.
 
-## Step 5: Result presentation and evaluation In the main manuscript we present
-the results of the analysis in the CCAE database with stratification based on
-risk predictions of acute MI. Results of analyses in the other databases and
-with other risk stratifications are included in the supplementary material.
+## Step 5: Result presentation and evaluation 
+In the main manuscript we present the results of the analysis in the CCAE
+database with stratification based on risk predictions of acute MI. Results of
+analyses in the other databases and with other risk stratifications are
+included in the supplementary material.
 
@@
 is in accordance with earlier findings that beta blockers should be considered
-as first-line treatment for younger hypertensive patients. More thorough
+as first-line treatment for younger hypertensive patients [@Cruickshank2007; @Khan2006]. More thorough
 evaluation of these results is required in future research.
@@
 is implemented in an open source R-package in the OHDSI methods library
-(https://github.com/OHDSI/RiskStratifiedEstimation). As a proof-of-concept, we
+(<https://github.com/OHDSI/RiskStratifiedEstimation>). As a proof-of-concept, we
 used our framework to evaluate heterogeneity of the effect of treatment with
@@
 In recent years several methods for the evaluation of treatment effect
-heterogeneity have been developed in the setting of RCTs. However, low power and
-restricted prior knowledge on the mechanisms of variation in treatment effect
-are often inherent in RCTs, which are often adequately powered only for the
-analysis of the primary outcome. Observational databases contain a large amount
-of information on treatment assignment and outcomes of interest, while also
-capturing key patient characteristics. Our framework provides a standardized
-approach that can be used to leverage available information from these data
-sources, allowing for large-scale assessment of treatment effect heterogeneity.
-Multiple outcomes can be evaluated in patient subgroups of similar baseline
-outcome risk, where different outcome risk stratification schemes can be
-considered. The standardized nature of the framework enables transportability to
-multiple databases, provided that they are mapped to the OMOP-CDM.
+heterogeneity have been developed in the setting of RCTs [@Rekkas2020].
+However, low power and restricted prior knowledge on the mechanisms of variation
+in treatment effect are often inherent in RCTs, which are often adequately
+powered only for the analysis of the primary outcome. Observational databases
+contain a large amount of information on treatment assignment and outcomes of
+interest, while also capturing key patient characteristics. Our framework
+provides a standardized approach that can be used to leverage available
+information from these data sources, allowing for large-scale assessment of
+treatment effect heterogeneity. Multiple outcomes can be evaluated in patient
+subgroups of similar baseline outcome risk, where different outcome risk
+stratification schemes can be considered. The standardized nature of the
+framework enables transportability to multiple databases, provided that they are
+mapped to the OMOP-CDM.
 
@@
 assessment of heterogeneity of treatment effect in RCT settings have been
-proposed. Our framework aims to translate these guidelines to the observational
-setting while also providing a toolset for its implementation. Several
-considerations need to be made. First, estimates may be biased due to the
-observational nature of the data. We attempt to account for potential
-confounding by estimating propensity scores within strata of predicted risk.
-These scores are estimated using regularized logistic regression on a large set
-of pre-defined covariates. This specific approach gave accurate results in
-extensive simulation studies. However, such approaches do not account for
-unobserved confounding. Several sensitivity analyses have been proposed in the
-literature for measuring the robustness of results in the presence of unobserved
-confounding. Another approach is to calibrate estimates and confidence intervals
-based on a large set of negative controls. Negative controls are
+proposed [@Kent2019; @PathEnE]. Our framework aims to translate these
+guidelines to the observational setting while also providing a toolset for its
+implementation. Several considerations need to be made. First, estimates may be
+biased due to the observational nature of the data. We attempt to account for
+potential confounding by estimating propensity scores within strata of predicted
+risk. These scores are estimated using regularized logistic regression on a
+large set of pre-defined covariates. This specific approach gave accurate
+results in extensive simulation studies [@Tian2018]. However, such approaches
+do not account for unobserved confounding [@Liu2013]. Several sensitivity
+analyses have been proposed in the literature for measuring the robustness of
+results in the presence of unobserved confounding. Another approach is to
+calibrate estimates and confidence intervals based on a large set of negative
+controls [@Schuemie2014; @Schuemie2571]. Negative controls are
 treatment-outcome pairs for which a null effect has been established. Estimating
@@
 clinical interpretation, it is not appropriate to guide clinical practice, where
-decisions need to be made at the individual level. Presentation of treatment
-effects as a continuous function of risk would be more helpful, but is
+decisions need to be made at the individual level [@Kent2019]. Presentation of
+treatment effects as a continuous function of risk would be more helpful, but is
 methodologically challenging. Future research is necessary for the development
@@
 Externally derived prediction models are preferred for analyzing treatment
-effect heterogeneity. Such external models should be well transportable. In the
-absence of such prediction models, simulations of RCTs have shown that
-internally derived models can be used to provide unbiased estimates of treatment
-effect across the spectrum of baseline risk. However, in observational databases
-treatment arms may significantly differ in sample size. Because the prediction
-model will possibly better fit to the larger treatment arm, this may introduce
-spurious treatment-covariate interactions in the prediction model, leading to
-sub-optimal risk stratification. As a remedy, we first match the patients in the
-treatment and the comparator cohorts on the basis of propensity scores.
-Additionally, we propose to assess model performance in the separate treatment
-arms to evaluate its aptness for risk stratification.
+effect heterogeneity [@Kent2010]. Such external models should be well
+transportable. In the absence of such prediction models, simulations of RCTs
+have shown that internally derived models can be used to provide unbiased
+estimates of treatment effect across the spectrum of baseline risk
+[@Burke2014]. However, in observational databases treatment arms may
+significantly differ in sample size. Because the prediction model will possibly
+better fit to the larger treatment arm, this may introduce spurious
+treatment-covariate interactions in the prediction model, leading to sub-optimal
+risk stratification. As a remedy, we first match the patients in the treatment
+and the comparator cohorts on the basis of propensity scores. Additionally, we
+propose to assess model performance in the separate treatment arms to evaluate
+its aptness for risk stratification.
 
@@
 Recently, disease risk scores have been explored as an alternative to propensity
-scores for balancing covariates. In our method, the objective of risk
-stratification is not balancing, but assessing the variation of treatment
-effects on multiple outcomes across patients with different levels of baseline
-risk. Although using the same risk model for balancing and risk-based HTE
-analysis may sound attractive, we note that our method only uses one risk model
-for stratification and one propensity score model for balancing, while separate
-disease risk score models would be required to analyze treatment effects for
-each of the multiple outcomes.
+scores for balancing covariates [@Glynn2012; @Hansen2008]. In our method, the
+objective of risk stratification is not balancing, but assessing the variation
+of treatment effects on multiple outcomes across patients with different levels
+of baseline risk. Although using the same risk model for balancing and
+risk-based HTE analysis may sound attractive, we note that our method only uses
+one risk model for stratification and one propensity score model for balancing,
+while separate disease risk score models would be required to analyze treatment
+effects for each of the multiple outcomes.
 
@@
 the overall results more actionable for medical decision making.
+
+\newpage
+# References
+\setlength{\parindent}{-0.25in}
+\setlength{\leftskip}{0.25in}
+\noindent
+<div id="refs"></div>
+\setlength{\parindent}{0in}
+\setlength{\leftskip}{0in}
+\noindent

### 2021-04-20 submission/manuscript.rmd
new version of manuscript

@@
 title: "A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases"
+author: 
+  - Alexandros Rekkas, MSc${^1}$ 
+  - David van Klaveren, PhD$^{2,3}$, 
+  - Patrick B. Ryan, PhD$^4$
+  - Ewout W. Steyerberg, PhD$^{3,5}$
+  - David M. Kent, PhD$^2$
+  - Peter R. Rijnbeek, PhD${^1}$
 output:
@@
   - \usepackage[left, pagewise]{lineno}
+  - \usepackage{caption}
 editor_options: 
@@
 
+```{r, echo=FALSE, message=FALSE}
+library(tidyverse)
+library(knitr)
+library(here)
+library(kableExtra)
+d <- function(x, decimals = 2) {
+  sprintf(paste0("%1.", decimals, "f"), x) 
+}
+knit_hooks$set(
+  inline = function(x) {
+    prettyNum(
+      x,
+      big.mark = ",",
+      decimal.mark = ".",
+      preserve.width = "individual"
+    )
+  }
+)
+```
+\vspace{20mm}
+
+$^1$ Department of Medical Informatics, Erasmus University Medical Center,
+Rotterdam, Netherlands
+
+$^2$ Predictive Analytics and Comparative Effectiveness (PACE) Center, Institute
+for Clinical Research and Health Policy Studies (ICRHPS), Tufts Medical Center,
+Boston, MA, USA
+
+$^3$ Department of Public Health, Erasmus University Medical Center, Rotterdam,
+Netherlands
+
+$^4$ Janssen Research and Development, 125 Trenton Harbourton Rd,
+Titusville,NJ 08560, USA
+
+$^5$ Department of Biomedical Data Sciences, Leiden University Medical Center,
+Leiden, The Netherlands
+
+\vspace{30mm}
+**Funding**: This work has been performed in the European Health Data and
+Evidence Network (EHDEN) project. This project has received funding from the
+Innovative Medicines Initiative 2 Joint Undertaking (JU) under grant agreement
+No 806968. The JU receives support from the European Union’s Horizon 2020
+research and innovation programme and EFPIA.
+
+\newpage
 \newpage
@@
 The Observational Health Data Sciences and Informatics (OHDSI) collaborative
-(!!REF) has established an international network of data partners and
-researchers that aim to bring out the value of health data through large-scale
-analytics by mapping all available databases to the Observational Medical
-Outcomes Partnership (OMOP) Common Data Model (CDM) [@Overhage2012]. The
-common data structure enables analyses at a very large scale. For example, in a
-recent study [@Suchard2019], a large set of first-line treatments for
+[@hripcsak2015observational] has established an international network of data
+partners and researchers that aim to bring out the value of health data through
+large-scale analytics by mapping all available databases to the Observational
+Medical Outcomes Partnership (OMOP) Common Data Model (CDM) [@Overhage2012].
+The common data structure enables analyses at a very large scale. For example,
+in a recent study [@Suchard2019], a large set of first-line treatments for
 hypertension was compared with respect to 55 outcomes in a network of databases,
@@
 presentation and evaluation of the results. A simple overview of the procedure
-can be seen in Figure XXXX.
+can be seen in Figure \@ref(fig:figure1).
+
+```{r figure1, echo=FALSE, fig.cap="Caption for figure 1", out.width = '100%'}
+# knitr::include_graphics(here("figures/Figure_1.tif"))
+grid::grid.raster(tiff::readTIFF(here("figures/Figure_1.tif")))
+```
 
@@
 treatment of hypertension, more recent trials and meta-analyses have cast doubt
-on their relative effectiveness (!!REF). As a result, newer US guidelines do
+on their relative effectiveness [@wiysonge2017beta]. As a result, newer US guidelines do
 not consider them for initial treatment for hypertension while in the EU
@@
 For each outcome and in each risk stratum there were adequate numbers of
-patients (Table XXXX). The discriminative ability of the prediction models was
-moderate in the matched development subset (c-index 0.76 for acute MI ; 0.79 for
-hospitalization with heart failure;  0.74 for stroke; ), in the general
-population (c-index 0.74 for acute MI; 0.77 for hospitalization with heart
-failure; 0.73 for stroke), in the treatment cohort (c-index for acute MI it was
-0.71, for hospitalization with heart failure was 0.76 and for stroke it was
-0.72) and in the comparator cohort (c-index for acute MI it was 0.79 for
+patients (Table \@ref(tab:table)). The discriminative ability of the prediction
+models was moderate in the matched development subset (c-index 0.76 for acute MI
+; 0.79 for hospitalization with heart failure; 0.74 for stroke; ), in the
+general population (c-index 0.74 for acute MI; 0.77 for hospitalization with
+heart failure; 0.73 for stroke), in the treatment cohort (c-index for acute MI
+it was 0.71, for hospitalization with heart failure was 0.76 and for stroke it
+was 0.72) and in the comparator cohort (c-index for acute MI it was 0.79 for
 hospitalization with heart failure was 0.79 and for stroke it was 0.75).
+\captionsetup{width=\textwidth}
+```{r table, echo=FALSE, results='asis', booktabs=TRUE}
+table <- tibble(
+  outcome = c(
+    rep("Acute myocardial infarction", 4),
+    rep("Heart failure (hosp)", 4),
+    rep("Stroke (ischemic or hemorrhagic)", 4)
+  ),
+  riskQuarter = rep(1:4, 3),
+  patientsAce = c(
+    161099, 204882, 214413, 204167,
+    146259, 188006, 218052, 230226,
+    146069, 187524, 217070, 226128
+  ),
+  personYearsAce = c(
+    276171, 372197, 393583, 351727,
+    249809, 341014, 399394, 400330,
+    294484, 340234, 397830, 393861
+  ),
+  eventsAce = c(
+    203, 534, 117, 2095,
+    228, 457, 826, 2012,
+    299, 554, 947, 1718
+  ),
+  patientsBeta = c(
+    133977, 90193, 80662, 90908,
+    126387, 84280, 83421, 98380,
+    126264, 84000, 83038, 97628
+  ),
+  personYearsBeta = c(
+    220633, 169231, 150035, 154419,
+    206706, 158425, 155222, 169139,
+    206453, 157913, 154587, 167810
+  ),
+  eventsBeta = c(
+    135, 321, 535, 1520,
+    378, 340, 570, 1773,
+    320, 351, 521, 1077
+  )
+)
+
+knitr::kable(
+  table,
+  format = "latex",
+  longtable = TRUE,
+  booktabs = TRUE,
+  col.names = c(
+    "Outcome", "Risk quarter",
+    "Persons", "Person years", "Events",
+    "Persons", "Person years", "Events"
+  ),
+   format.args = list(
+    big.mar = ","
+  ),
+  caption = "Number of patients, person years and events within quarters of predicted risk for hospitalization with heart failure for the 3 main outcomes of the study (acute myocardial infarction, hospitalization with heart failure and ischemic or hemorrhagic stroke)."
+) %>%
+  kableExtra::add_header_above(
+    c(
+      "", "", 
+      "ACE inhibitors" = 3,
+      "Beta blockers"  = 3
+    )
+  )
+```
 
@@
 increases of absolute treatment effects (ARD) with increasing acute MI risk
-(Figure XXXX). Patients in the low risk quarter did not receive absolute
-treatment benefit (ARD -0.03%) while absolute risk was 0.54% lower (95%
-confidence interval 0.36%—0.71%) for patients in the high-risk quarter. In
-contrast, the absolute and relative effects of ACE-inhibitors on safety outcomes
-(e.g. cough and angioedema) are approximately constant or even slightly
-decreasing with increasing acute MI risk (Figure XXXX and XXXX). Similar results
-were observed in the other two databases (see supplementary material).
+(Figure \@ref(fig:figure2)).
+
+```{r figure2, echo=FALSE, fig.cap="Overview of heterogeneity of ACE-inhibitors treatment within strata of predicted risk of acute MI.
+The top panel contains the observed acute rates of ACE-inhibitors and beta blockers within each quarter of predicted acute MI risk.
+These are derived using the KaplanMeier estimates at 730 days after inclusion.
+The middle panel, contains the hazard ratios of comparing ACE-inhibitors to beta blockers with regard to acute MI.
+These are estimated using Cox proportional hazards regression within quarters of predicted acute MI risk.
+The bottom panel contains absolute risk reduction for ACE-inhibitors compared to beta blockers.
+These are derived as the difference in Kaplan-Meier estimates at 730 after inclusion.
+Hazard ratios in the middle panel show a decreasing trend with increasing acute MI risk.
+Given the rather good discrimination of the prediction model (AUC=0.74), this results in an increasing trend for absolute benefit in favor of ACE-inhibitors with increasing risk.", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/Figure_2.tif")))
+```
+Patients in the low risk quarter did not receive absolute treatment benefit (ARD
+-0.03%) while absolute risk was 0.54% lower (95% confidence
+interval 0.36%—0.71%) for patients in the high-risk quarter. In contrast, the
+absolute and relative effects of ACE-inhibitors on safety outcomes (e.g. cough
+and angioedema) are approximately constant or even slightly decreasing with
+increasing acute MI risk (Figure \@ref(fig:figure3) and \@ref(fig:figure4)).
+Similar results were observed in the other two databases (see supplementary
+material).
+
+```{r figure3, echo=FALSE, fig.cap="Hazard ratios (relative treatment effects) for the main and safety outcomes, estimated by fitting stratified Cox regression models within quarters of predicted risk of acute myocardial infarction (MI).
+The four risk quarters (Q1-Q4) are defined using the internally developed model for acute MI", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/Figure_3.tif")))
+```
+```{r figure4, echo=FALSE, fig.cap="Absolute risk reduction for the main and safety outcomes, estimated by fitting stratified Cox regression models within quarters of predicted risk of acute myocardial infarction (MI).
+The four risk quarters (Q1-Q4) are defined using the internally developed model for acute MI", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/Figure_4.tif")))
+```
 
@@
 Our contribution is a translation of the PATH statement principles to the OHDSI
-methods library. Our methods encourage open science as it requires clear
+methods library. Our method encourages open science as it requires clear
 definition of the research questions translated into clear and reproducible
@@
 provides a standardized stepwise procedure for the assessment of HTE. This
-enables source code to be easily shared and evaluated. The results of these
-analyses can be reproduced in a straightforward manner. Researchers with access
-to different databases mapped to OMOP-CDM can also very easily extend their
-overall analyses with risk-based assessment of treatment effect heterogeneity.
-This enables collaboration among multiple sites with access to different patient
-populations. We propose that the framework is implemented any time treatment
-effect estimation in high-dimensional observational data is undertaken.
+enables source code to be easily shared and evaluated, allowing for reproducible
+research . Researchers with access to different databases mapped to OMOP-CDM can
+also very easily extend their overall analyses with risk-based assessment of
+treatment effect heterogeneity. This enables collaboration among multiple sites
+with access to different patient populations. We propose that the framework is
+implemented any time treatment effect estimation in high-dimensional
+observational data is undertaken.
 
@@
 # References
+\nolinenumbers
 \setlength{\parindent}{-0.25in}

### 2021-04-28 submission/Highlights.rmd
proof reading submission

---
title: "Highlights"
output:
    bookdown::pdf_document2: default
geometry: margin=1.0in
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left, pagewise]{lineno}
  - \usepackage{caption}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Highlights
  - Standardized framework for risk-based assessment of treatment effect
    heterogeneity
  - Always run the framework after population-level effect estimation problems
  - Run federated analyses across different databases mapped to OMOP common data
    model

### 2021-04-28 submission/manuscript.rmd
proof reading submission

@@
 ```
-\vspace{20mm}
+\thispagestyle{empty}
+\vspace{8mm}
 
@@
 
-\vspace{30mm}
-**Funding**: This work has been performed in the European Health Data and
+\vspace{10mm}
+**Corresponing author**
+\singlespacing 
+Alexandros Rekkas, MSc
+
+Department of Medical Informatics
+
+Erasmus University Medical Center
+
+3000 CA Rotterdam, P.O. Box 2040
+
+email: a.rekkas@erasmusmc.nl
+\onehalfspacing
+
+\vspace{10mm}
+**Funding**
+
+This work has been performed in the European Health Data and
 Evidence Network (EHDEN) project. This project has received funding from the
@@
 \newpage
-# Abstract (247 words)
+# Abstract {-}
 \singlespacing 
-**Aim:** One of the aims of the Observation Health Data Sciences
-and Informatics (OHDSI) initiative is population-level treatment effect
-estimation in large observational databases. Since treatment effects are
-well-known to vary across groups of patients with different baseline risk, we
-aimed to extend the OHDSI methods library with a framework for risk-based
-assessment of treatment effect heterogeneity.
+One of the aims of the Observation Health Data Sciences and Informatics (OHDSI)
+initiative is population-level treatment effect estimation in large
+observational databases. Since treatment effects are well-known to vary across
+groups of patients with different baseline risk, we aimed to extend the OHDSI
+methods library with a framework for risk-based assessment of treatment effect
+heterogeneity. The proposed framework consists of five steps: 1) definition of
+the problem, i.e. the population, the treatment, the comparator and the
+outcome(s) of interest; 2) identification of relevant databases; 3) development
+of a prediction model for the outcome(s) of interest; 4) estimation of
+propensity scores within strata of predicted risk and estimation of relative and
+absolute treatment effect within strata of predicted risk; 5) evaluation and
+presentation of results. We demonstrate our framework by evaluating
+heterogeneity of the effect of angiotensin-converting enzyme (ACE) inhibitors
+versus beta blockers on a set of 9 outcomes of interest across three
+observational databases. With increasing risk of acute myocardial infarction we
+observed increasing absolute benefits, i.e. from -0.03% to 0.54% in the lowest
+to highest risk groups. Cough-related absolute harms decreased from 4.1%
+to 2.6%. he proposed framework may be useful for the evaluation of heterogeneity
+of treatment effect on observational data that are mapped to the OMOP Common
+Data Model. The proof of concept study demonstrates its feasibility in large
+observational data. Further insights may arise by application to safety and
+effectiveness questions across the global data network.
 
-**Materials and Methods:** The proposed framework consists of five steps: 1)
-definition of the problem, i.e. the population, the treatment, the comparator
-and the outcome(s) of interest; 2) identification of relevant databases; 3)
-development of a prediction model for the outcome(s) of interest; 4) estimation
-of propensity scores within strata of predicted risk and estimation of relative
-and absolute treatment effect within strata of predicted risk; 5) evaluation and
-presentation of results.
+\vspace{10mm}
 
-**Results:** We demonstrate our framework by evaluating heterogeneity of the
-effect of angiotensin-converting enzyme (ACE) inhibitors versus beta blockers on
-a set of 9 outcomes of interest across three observational databases. With
-increasing risk of acute myocardial infarction we observed increasing absolute
-benefits, i.e. from -0.03% to 0.54% in the lowest to highest risk groups.
-Cough-related absolute harms decreased from 4.1% to 2.6%.
-
-**Conclusions:** The proposed framework may be useful for the evaluation of
-heterogeneity of treatment effect on observational data that are mapped to the
-OMOP Common Data Model. The proof of concept study demonstrates its feasibility
-in large observational data. Further insights may arise by application to safety
-and effectiveness questions across the global data network.
-
+**Keywords**: observational data, heterogeneity of treatment effect, risk stratification, subgroup analysis
 \newpage 
@@
 # Introduction
+
 Interest in understanding how a treatment’s effect varies across patients—a
@@
 
-The Observational Health Data Sciences and Informatics (OHDSI) collaborative
-[@hripcsak2015observational] has established an international network of data
-partners and researchers that aim to bring out the value of health data through
-large-scale analytics by mapping all available databases to the Observational
-Medical Outcomes Partnership (OMOP) Common Data Model (CDM) [@Overhage2012].
-The common data structure enables analyses at a very large scale. For example,
-in a recent study [@Suchard2019], a large set of first-line treatments for
-hypertension was compared with respect to 55 outcomes in a network of databases,
-including 4.9 million patients from around the world.
+The Observational Health Data Sciences and Informatics (OHDSI) collaborative has
+established an international network of data partners and researchers that aim
+to bring out the value of health data through large-scale analytics by mapping
+all available databases to the Observational Medical Outcomes Partnership (OMOP)
+Common Data Model (CDM) [@hripcsak2015observational; @Overhage2012]. The
+common data structure enables analyses at a very large scale. For example, in a
+recent study, a large set of first-line treatments for hypertension was compared
+with respect to 55 outcomes in a network of databases, including 4.9 million
+patients from around the world [@Suchard2019].
 
@@
 framework using existing OHDSI methods for use in the OMOP-CDM, including the
-patient-level prediction framework [@Reps2018] and the population-level effect
-estimation framework [@Ryan2013] based on new-user cohort design. As a
+patient-level prediction framework and the population-level effect estimation
+framework based on new-user cohort design [@Reps2018; @Ryan2013] . As a
 proof-of-concept we analyzed heterogeneity of the effects of first-line
@@
 
-# Methods
+# Materials and Methods
+
 The proposed framework defines 5 distinct steps that enable a standardized
@@
 
-```{r figure1, echo=FALSE, fig.cap="Caption for figure 1", out.width = '100%'}
+```{r figure1, echo=FALSE, fig.cap="(A) Starting from a treatment (top), a comparator (bottom) and an outcome (middle) cohort we estimate the propensity scores on the entire target population.
+(B) We match patients on the propensity scores and estimate the prediction model.
+Since we match patients we develop the prediction model on smaller subset of the initial population and, therefore, the number of patients is smaller in B compared to A.
+(C) We apply the prediction model on the entire population (green: lower 25\\% of the risk distribution; yellow: patients with risk between 25\\% and 50\\% of the risk distribution; orange: patients with risk between 50\\% and 75\\% of the risk distribution; red: patients at risk higher than 75\\% of the risk distribution).
+(D) We separate in risk subgroups, here quarters.
+Within risk quarters propensity scores are estimated again and relative and absolute treatment effects are estimated.", out.width = '100%'}
 # knitr::include_graphics(here("figures/Figure_1.tif"))
@@
 ## Step 1: General definition of the problem
+
 The typical research aim is: "to compare the effect of treatment $T$ to a
@@
 
-* A single treatment cohort ($T$) which includes patients with disease $D$
-receiving the target treatment of interest. For example, a set of hypertension
-patients within a database that receive angiotensin-converting enzyme
-inhibitors, followed from the time of initiation until the time of censoring.
-* A single comparator cohort ($C$) which includes patients with disease $D$
+  - A single treatment cohort ($T$) which includes patients with disease $D$
+    receiving the target treatment of interest. For example, a set of
+    hypertension patients within a database that receive angiotensin-converting
+    enzyme inhibitors, followed from the time of initiation until the time of
+    censoring.
+  - A single comparator cohort ($C$) which includes patients with disease $D$
     receiving the comparator (control) treatment. For example, a set of patients
-in a database that receive beta blockers, followed from the time of initiation
-until the time of censoring.
-* One or more outcome cohorts ($O_1,\dots,O_n$) that contain patients
-developing the outcomes of interest. For example, the
-set of patients in a database that have at least one occurrence of acute
-myocardial infarction in their record.
+    in a database that receive beta blockers, followed from the time of
+    initiation until the time of censoring.
+  - One or more outcome cohorts ($O_1,\dots,O_n$) that contain patients
+    developing the outcomes of interest. For example, the set of patients in a
+    database that have at least one occurrence of acute myocardial infarction in
+    their record.
 
 ## Step 2: Identification of the database
+
 The aim of this step is the inclusion of databases that represent the patient
@@
 ## Step 3: Prediction
+
 We adopt the standardized framework for the generation of patient-level
@@
 To generate the target cohort we pool the already defined treatment cohort and
-comparator cohort . However, for risk-based analysis of treatment effects it is
+comparator cohort. However, for risk-based analysis of treatment effects it is
 necessary to avoid deferentially fitting the prediction model to patients across
 the treatment arm in order to avoid inducing spurious interactions [@Burke2014;
-@vanKlaveren2019]. To do this, we developed the patient-level prediction model
-in the propensity score-matched subset of the population (1:1), where treatment
+@vanKlaveren2019]. To do this, we develop the patient-level prediction model in
+the propensity score-matched subset of the population (1:1), where treatment
 assignment is well-balanced. The propensity scores are based on LASSO logistic
@@
 ## Step 4: Estimation
-We use the patient-level prediction model to divide the target population into
-a set of equally-sized risk strata, typically 4 risk quarters. Then, we
-estimate propensity scores within risk strata. These propensity scores are used
-when estimating treatment effects, either by matching of patients from
-different treatment cohorts, by stratification of patients into groups with
-similar propensity scores, or by weighing patients’ contribution to the
-estimation process. Within risk strata we estimate treatment effect both on the
-relative and the absolute scale. It is important to evaluate treatment effects
-in both scales, as effect cannot remain constant on both the relative and the
-absolute scale at the same time, assuming a non-zero treatment
-effect—treatment. Any appropriate method for the evaluation of relative and
-absolute treatment effects can be considered, as long as the this is done
-consistently in all risk strata.
+
+We use the patient-level prediction model to divide the target population into a
+set of equally-sized risk strata, typically 4 risk quarters. Then, we estimate
+propensity scores within risk strata. These propensity scores are used when
+estimating treatment effects, either by matching of patients from different
+treatment cohorts, by stratification of patients into groups with similar
+propensity scores, or by weighing patients’ contribution to the estimation
+process. Within risk strata we estimate treatment effect both on the relative
+and the absolute scale. It is important to evaluate treatment effects in both
+scales, as effect cannot remain constant on both the relative and the absolute
+scale at the same time, assuming a non-zero treatment effect. Any appropriate
+method for the evaluation of relative and absolute treatment effects can be
+considered, as long as the this is done consistently in all risk strata.
 
 ## Step 5: Result presentation and evaluation
-Our framework provides standardized output for each step of the analysis. The
-number of patients and person years by treatment arm along with the number of
-outcomes. A performance overview of the derived prediction models, including
-discrimination and calibration both in the propensity score matched subset, the
-entire population and separately for treated and comparator patients. This is
-rather relevant as the performance of the prediction models is directly related
-to our ability to single out patient subgroups where treatment may be highly
-beneficial or unsafe.  Kent et al [@Kent2016] demonstrated that the event rate and the
-discriminative ability of the prediction model can predict very well the
-distribution of predicted risk. The lower the event rate and the higher the
-c-statistic (given good calibration) result in high risk heterogeneity, thus
-making estimated average treatment effects uninformative. In this case, risk
-stratified analysis of HTE can be more
+
+The number of patients and person years by treatment arm along with the number
+of outcomes should be reported. A performance overview of the derived prediction
+models, including discrimination and calibration both in the propensity score
+matched subset, the entire population and separately for treated and comparator
+patients should also be reported. This is rather relevant as the performance of
+the prediction models is directly related to our ability to single out patient
+subgroups where treatment may be highly beneficial or unsafe. Kent et al
+[@Kent2016] demonstrated that the event rate and the discriminative ability of
+the prediction model can predict very well the distribution of predicted risk.
+The lower the event rate and the higher the c-statistic (given good calibration)
+result in high risk heterogeneity, thus making estimated average treatment
+effects uninformative. In this case, risk stratified analysis of HTE can be more
 effective in singling out patient subgroups that stand to benefit (or be harmed)
@@
 
-Propensity score distributions by treatment group and covariate balance plots
-for each risk stratum. Event rates, hazard ratios and absolute risk differences
-in risk strata for a selected outcome, both in tables and in graphs. Hazard
-ratios and absolute risk differences for all analyzed outcomes by risk stratum.
-Finally, shiny application can be generated to enable easy sharing of the
-results.
+In every risk stratum propensity score distributions by treatment group and
+covariate balance plots should be made available along with event rates, hazard
+ratios and absolute risk differences. Finally, the results should be generated
+with a standardized format to allow for combination of results from multiple
+data sources. The standardized result format also enables the development of
+scalable demonstration tools such as R-shiny applications, interactive tables
+and figures.
 
 # Results
+
 As a proof of concept, we focus on the comparison of angiotensin converting
 enzyme (ACE) inhibitors to beta blockers. ACE inhibitors are among the most
-common treatment classes for hypertension, with a well-established
-effectiveness. Beta blockers, even though initially widely used for the
-treatment of hypertension, more recent trials and meta-analyses have cast doubt
-on their relative effectiveness [@wiysonge2017beta]. As a result, newer US guidelines do
-not consider them for initial treatment for hypertension while in the EU
+common treatment classes for hypertension, with well-established effectiveness.
+Beta blockers, even though initially widely used for the treatment of
+hypertension, more recent trials and meta-analyses have cast doubt on their
+relative effectiveness [@wiysonge2017beta]. As a result, newer US guidelines
+do not consider them for initial treatment for hypertension while in the EU
 guidelines combination with other antihypertensive treatments is recommended
@@
 classes in younger hypertensive patients and, thus, countries like Canada still
-include them as a first-line candidate for the treatment [@Khan2006;
-@Rabi2020].
+include them as a first-line treatment candidate [@Khan2006; @Rabi2020].
 
 ## Step 1: General definition of the problem
+
 We demonstrate the framework with the following research aim: "to compare the
@@
 
-* Treatment cohort: Patients receiving any drug within the ACE-inhibitor class
-with at least one year of follow-up before treatment start and a recorded
-hypertension diagnosis within that year.
-* Comparator cohort: Patients
-receiving any drug within the beta blocker class with at least one year of
-follow-up before treatment start and a recorded hypertension diagnosis within
-that year.
-* Outcome cohorts: We consider 3 main and 6 safety outcome cohorts.
-These are patients in the database with a diagnosis of: acute myocardial
-infarction (MI); hospitalization with heart failure; ischemic or hemorrhagic
-stroke (efficacy outcomes); hypokalemia; hyperkalemia; hypotension;
-angioedema; cough; abnormal weight gain (safety outcomes).
+  - Treatment cohort: Patients receiving any drug within the ACE-inhibitor class
+    with at least one year of follow-up before treatment initiation and a
+    recorded hypertension diagnosis within that year.
+  - Comparator cohort: Patients receiving any drug within the beta blocker class
+    with at least one year of follow-up before treatment initiation and a
+    recorded hypertension diagnosis within that year.
+  - Outcome cohorts: We consider 3 main and 6 safety outcome cohorts. These are
+    patients in the database with a diagnosis of: acute myocardial infarction
+    (MI); hospitalization with heart failure; ischemic or hemorrhagic stroke
+    (efficacy outcomes); hypokalemia; hyperkalemia; hypotension; angioedema;
+    cough; abnormal weight gain (safety outcomes).
 
@@
 ## Step 2: Identification of the databases
+
 We used the following databases:
 
-* IBM MarketScan Medicare Supplemental Beneficiaries (MDCR): Represents health
+  - IBM MarketScan Medicare Supplemental Beneficiaries (MDCR): Represents health
     services of retirees (aged 65 or older) in the United States with primary or
@@
     point-of-service or capitated health plans. These data include adjudicated
-health insurance claims (e.g. inpatient, outpatient and outpatient pharmacy).
-Additionally, it captures laboratory tests for a subset of the covered lives.
-* IBM MarketScan Medicaid (MDCD): Adjudicated US health insurance claims for
+    health insurance claims (e.g. inpatient, outpatient and outpatient
+    pharmacy). Additionally, it captures laboratory tests for a subset of the
+    covered lives.
+  - IBM MarketScan Medicaid (MDCD): Adjudicated US health insurance claims for
     Medicaid enrollees from multiple states. It includes hospital discharge
-diagnoses, outpatient diagnoses and procedures and outpatient pharmacy claims
-as well as ethnicity and Medicare eligibility.  IBM MarketScan Commercial
-* Claims and Encounters (CCAE): Data from individuals enrolled in US
-employer-sponsored   insurance   health   plans. The data includes adjudicated
-health insurance   claims (e.g. inpatient, outpatient, and outpatient
-pharmacy) as well as enrollment data from large employers and health plans who
-provide private healthcare coverage to employees, their spouses and
-dependents. Additionally, it captures laboratory tests for a subset of the
-covered lives.
+    diagnoses, outpatient diagnoses and procedures and outpatient pharmacy
+    claims as well as ethnicity and Medicare eligibility.
+  - IBM MarketScan Commercial Claims and Encounters (CCAE): Data from
+    individuals enrolled in US employer-sponsored insurance health plans. The
+    data includes adjudicated health insurance claims (e.g. inpatient,
+    outpatient, and outpatient pharmacy) as well as enrollment data from large
+    employers and health plans who provide private healthcare coverage to
+    employees, their spouses and dependents. Additionally, it captures
+    laboratory tests for a subset of the covered lives.
 
 ## Step 3: Prediction
-To obtain a target cohort for developing patient-level
-predictions we first merged the ACE-inhibitors cohort with the beta blockers
-cohort. We then matched patients in the ACE-inhibitor cohort to patients in the
-beta blockers cohort on the propensity score. We focused on the efficacy
-outcomes (acute MI, hospitalization with heart failure and hemorrhagic or
-ischemic stroke) for risk stratification of the patient population. In each
-database, for each main outcome, we developed a prediction model. We chose a
-time horizon of 2 years after inclusion into the target cohort. We developed the
-prediction models using LASSO logistic regression with 3-fold cross validation
-for hyper-parameter selection.
+
+To obtain a target cohort for developing patient-level predictions we first
+merged the ACE-inhibitors cohort with the beta blockers cohort. We then matched
+patients in the ACE-inhibitor cohort to patients in the beta blockers cohort on
+the propensity score. We focused on the efficacy outcomes (acute MI,
+hospitalization with heart failure and hemorrhagic or ischemic stroke) for risk
+stratification of the patient population. In each database, for each main
+outcome, we developed a different prediction model. We chose a time horizon of 2
+years after inclusion into the target cohort. We developed the prediction models
+using LASSO logistic regression with 3-fold cross validation for hyper-parameter
+selection.
 
 ## Step 4: Estimation
-We used patient-level predictions to stratify the patient population into 4
-risk quarters. We estimated relative and absolute treatment effects for all 9
+
+We used patient-level predictions to stratify the patient population into 4 risk
+quarters. We estimated relative and absolute treatment effects for all 9
 outcomes of interest. In order to estimate risk quarter-specific treatment
@@
 ## Step 5: Result presentation and evaluation
-In the main manuscript we present the results of the analysis in the CCAE
-database with stratification based on risk predictions of acute MI. Results of
-analyses in the other databases and with other risk stratifications are
-included in the supplementary material.
+
+Here we present the results of the analysis in the CCAE database with
+stratification based on risk predictions of acute MI. Results of analyses in the
+other databases and with other risk stratifications are included in the
+supplementary material.
 
@@
   outcome = c(
-    rep("Acute myocardial infarction", 4),
-    rep("Heart failure (hosp)", 4),
-    rep("Stroke (ischemic or hemorrhagic)", 4)
+    "Acute myocardial infarction",
+    rep("", 3),
+    "Heart failure (hosp)",
+    rep("", 3),
+    "Stroke (ischemic or hemorrhagic)",
+    rep("", 3)
   ),
@@
 ratios decreased) with increasing acute MI risk, resulting in more pronounced
-increases of absolute treatment effects (ARD) with increasing acute MI risk
-(Figure \@ref(fig:figure2)).
+absolute risk differences (ARD) with increasing acute MI risk (Figure
+\@ref(fig:figure2)).
 
@@
 ```
-Patients in the low risk quarter did not receive absolute treatment benefit (ARD
--0.03%) while absolute risk was 0.54% lower (95% confidence
-interval 0.36%—0.71%) for patients in the high-risk quarter. In contrast, the
-absolute and relative effects of ACE-inhibitors on safety outcomes (e.g. cough
-and angioedema) are approximately constant or even slightly decreasing with
-increasing acute MI risk (Figure \@ref(fig:figure3) and \@ref(fig:figure4)).
-Similar results were observed in the other two databases (see supplementary
-material).
+
+Patients in the low risk quarter did not receive absolute benefit (-0.03%)
+while absolute risk was 0.54% lower (95% confidence interval 0.36%—0.71%) for
+patients in the high-risk quarter. In contrast, the absolute and relative
+effects of ACE-inhibitors on safety outcomes (e.g. cough and angioedema) are
+approximately constant or even slightly decreasing with increasing acute MI risk
+(Figure \@ref(fig:figure3) and \@ref(fig:figure4)). Similar results were observed
+in the other two databases (see supplementary material).
 
@@
 ```
+
 ```{r figure4, echo=FALSE, fig.cap="Absolute risk reduction for the main and safety outcomes, estimated by fitting stratified Cox regression models within quarters of predicted risk of acute myocardial infarction (MI).
@@
 patients, in whom the benefit-harm tradeoff is in favor of beta blockers. This
-is in accordance with earlier findings that beta blockers should be considered
-as first-line treatment for younger hypertensive patients [@Cruickshank2007; @Khan2006]. More thorough
-evaluation of these results is required in future research.
+is in accordance with earlier findings that beta blockers should also be
+considered as first-line treatment for younger hypertensive patients
+[@Cruickshank2007; @Khan2006]. More thorough evaluation of these results is
+required in future research, however.
 
@@
 # Discussion
+
 We developed a framework for the assessment of heterogeneity of treatment effect
@@
 treatment effect heterogeneity. Multiple outcomes can be evaluated in patient
-subgroups of similar baseline outcome risk, where different outcome risk
-stratification schemes can be considered. The standardized nature of the
+subgroups of similar baseline outcome risk. Different outcome risk
+stratification schemes can also be considered. The standardized nature of the
 framework enables transportability to multiple databases, provided that they are
@@
 Our contribution is a translation of the PATH statement principles to the OHDSI
-methods library. Our method encourages open science as it requires clear
-definition of the research questions translated into clear and reproducible
-cohort definitions that can easily be shared among researchers. Our R-package
-provides a standardized stepwise procedure for the assessment of HTE. This
-enables source code to be easily shared and evaluated, allowing for reproducible
-research . Researchers with access to different databases mapped to OMOP-CDM can
-also very easily extend their overall analyses with risk-based assessment of
-treatment effect heterogeneity. This enables collaboration among multiple sites
-with access to different patient populations. We propose that the framework is
-implemented any time treatment effect estimation in high-dimensional
-observational data is undertaken.
+methods library [@Kent2019; @PathEnE]. Our method encourages open science as
+it requires clear definition of the research questions translated into clear and
+reproducible cohort definitions that can easily be shared among researchers. Our
+R-package provides a standardized stepwise procedure for the assessment of HTE.
+This enables source code to be easily shared and evaluated, allowing for
+reproducible research. Researchers with access to different databases mapped to
+OMOP-CDM can also very easily extend their overall analyses with risk-based
+assessment of treatment effect heterogeneity. This enables collaboration among
+multiple sites with access to different patient populations. We propose that the
+framework is implemented any time treatment effect estimation in
+high-dimensional observational data is undertaken.
 

### 2021-04-28 submission/manuscript.rmd
add missing reference

@@
 versus an alternative, taking into account multiple relevant patient
-characteristics. One promising approach is “risk modeling”, in which treatment
-effects are estimated in strata of predicted risk [@Kent2010; @Kent2018]. Such
-a risk-based approach first stratifies patients according to baseline risk
-predictions, using either an existing or an internally developed risk prediction
-model [@Kent2010]. Then, relative and absolute treatment effects are estimated
-within risk strata.
+characteristics [@Varadhan2013]. One promising approach is “risk modeling”, in
+which treatment effects are estimated in strata of predicted risk [@Kent2010;
+@Kent2018]. Such a risk-based approach first stratifies patients according to
+baseline risk predictions, using either an existing or an internally developed
+risk prediction model [@Kent2010]. Then, relative and absolute treatment
+effects are estimated within risk strata.
 

### 2021-04-29 submission/Highlights.rmd
add missing references

@@
   - Always run the framework after population-level effect estimation problems
-  - Run federated analyses across different databases mapped to OMOP common data
-    model
+  - Enables federated analyses across databases mapped to OMOP common data model

### 2021-04-29 submission/manuscript.rmd
add missing references

@@
 
-email: a.rekkas@erasmusmc.nl
+Email: a.rekkas@erasmusmc.nl
 \onehalfspacing
@@
 \singlespacing 
-One of the aims of the Observation Health Data Sciences and Informatics (OHDSI)
+One of the aims of the Observational Health Data Sciences and Informatics (OHDSI)
 initiative is population-level treatment effect estimation in large
@@
 to highest risk groups. Cough-related absolute harms decreased from 4.1%
-to 2.6%. he proposed framework may be useful for the evaluation of heterogeneity
+to 2.6%. The proposed framework may be useful for the evaluation of heterogeneity
 of treatment effect on observational data that are mapped to the OMOP Common
@@
 
-Interest in understanding how a treatment’s effect varies across patients—a
-concept described as heterogeneity of treatment effects (HTE)—has been growing.
-This concept is central to the agenda for both personalized (or precision)
-medicine and comparative effectiveness research. More formally, HTE has been
-defined as non-random variability in the direction or magnitude of a treatment
-effect, in which the effect is measured using clinical outcomes
-[@KRAVITZ2004]. Usually, analyses focus on the relative scale, where treatment
-effects are assessed one at a time in patient subgroups defined from single
-covariates, an approach that suffers from low power and multiplicity issues
-[@Varadhan2013]. However, even with well-established constant relative
-effects, treatment benefit (or harm) may vary substantially on the absolute
-scale.
+Understanding how a treatment’s effect varies across patients—a concept
+described as heterogeneity of treatment effects (HTE)—has been central to the
+agenda for both personalized (or precision) medicine and comparative
+effectiveness research. More formally, HTE has been defined as the non-random
+variability in the direction or magnitude of a treatment effect, in which the
+effect is measured using clinical outcomes [@KRAVITZ2004]. Usually, analyses
+focus on the relative scale, where treatment effects are assessed one at a time
+in patient subgroups defined from single covariates, an approach that suffers
+from low power and multiplicity issues [@Yusuf1991; @FernandezyGarcia2010].
+However, even with well-established constant relative effects, treatment benefit
+(or harm) may vary substantially on the absolute scale.
 
@@
 for HTE analysis, and include heterogeneous populations. However, unlike trials,
-treatment effects are subject to confounding and the unique structure of
+treatment effects are subject to confounding while the unique structure of
 different databases calls for database-specific analysis plans that are often
@@
 necessary to avoid deferentially fitting the prediction model to patients across
-the treatment arm in order to avoid inducing spurious interactions [@Burke2014;
-@vanKlaveren2019]. To do this, we develop the patient-level prediction model in
-the propensity score-matched subset of the population (1:1), where treatment
-assignment is well-balanced. The propensity scores are based on LASSO logistic
-regression for modeling the association between treatment assignment and all
-available demographics, drug exposures, diagnoses, measurements and medical
-procedures. Finally, we need to define the time horizon for which we aim to make
-predictions and we need to select the machine-learning algorithm we want to use
-to generate patient-level predictions. Currently, the available options are
-regularized logistic regression, random forest, gradient boosting machines,
-decision tree, naive Bayes, K-nearest neighbors, neural network and deep
-learning (convolutional neural networks, recurrent neural network and deep
-nets).
+the treatment arms in order to avoid inducing spurious interactions
+[@Burke2014; @vanKlaveren2019]. To do this, we develop the patient-level
+prediction model in the propensity score-matched subset of the population (1:1),
+where treatment assignment is well-balanced. The propensity scores are based on
+LASSO logistic regression for modeling the association between treatment
+assignment and all available demographics, drug exposures, diagnoses,
+measurements and medical procedures. Finally, we need to define the time horizon
+within which we aim to make predictions and we also need to select the
+machine-learning algorithm we want to use to generate patient-level predictions.
+Currently, the available options are regularized logistic regression, random
+forest, gradient boosting machines, decision tree, naive Bayes, K-nearest
+neighbors, neural network and deep learning (convolutional neural networks,
+recurrent neural network and deep nets).
 
@@
 matched subset, the entire population and separately for treated and comparator
-patients should also be reported. This is rather relevant as the performance of
+patients should also be reported. This is quite relevant as the performance of
 the prediction models is directly related to our ability to single out patient
@@
 the prediction model can predict very well the distribution of predicted risk.
-The lower the event rate and the higher the c-statistic (given good calibration)
-result in high risk heterogeneity, thus making estimated average treatment
-effects uninformative. In this case, risk stratified analysis of HTE can be more
+Lower event rate and higher c-statistic (given good calibration) result in high
+risk heterogeneity, thus making estimated average treatment effects
+uninformative. In this case, risk stratified analysis of HTE can be more
 effective in singling out patient subgroups that stand to benefit (or be harmed)
@@
 with a standardized format to allow for combination of results from multiple
-data sources. The standardized result format also enables the development of
-scalable demonstration tools such as R-shiny applications, interactive tables
-and figures.
+data sources. This also enables the development of scalable demonstration tools
+such as R-shiny applications, interactive tables and figures.
 
@@
 Our method provides a risk-stratified assessment of treatment effect
-heterogeneity. However, even though stratification can provide a rough guide for
-clinical interpretation, it is not appropriate to guide clinical practice, where
-decisions need to be made at the individual level [@Kent2019]. Presentation of
-treatment effects as a continuous function of risk would be more helpful, but is
-methodologically challenging. Future research is necessary for the development
-of methods for continuous risk-based assessment of HTE.
+heterogeneity. However, even though stratification can provide a useful overview
+for clinical interpretation, it is not appropriate to guide clinical practice,
+where decisions need to be made at the individual level [@Kent2019].
+Presentation of treatment effects as a continuous function of risk would be more
+helpful, but is methodologically challenging. Future research is necessary for
+the development of methods for continuous risk-based assessment of HTE.
 
-Externally derived prediction models are preferred for analyzing treatment
-effect heterogeneity [@Kent2010]. Such external models should be well
-transportable. In the absence of such prediction models, simulations of RCTs
-have shown that internally derived models can be used to provide unbiased
-estimates of treatment effect across the spectrum of baseline risk
-[@Burke2014]. However, in observational databases treatment arms may
-significantly differ in sample size. Because the prediction model will possibly
-better fit to the larger treatment arm, this may introduce spurious
+Externally derived and well transportable prediction models are preferred for
+analyzing treatment effect heterogeneity [@Kent2010]. In the absence of such
+prediction models, simulations of RCTs have shown that internal models can be
+used to provide unbiased estimates of treatment effect across the spectrum of
+baseline risk [@Burke2014]. However, in observational databases treatment arms
+may significantly differ in sample size. Because the prediction model will
+possibly better fit to the larger treatment arm, this may introduce spurious
 treatment-covariate interactions in the prediction model, leading to sub-optimal

### 2021-05-04 submission/manuscript.rmd
Fix typo
changes based on previous reviewer comments

@@
 
-Understanding how a treatment’s effect varies across patients—a concept
-described as heterogeneity of treatment effects (HTE)—has been central to the
+Understanding how a treatment’s effect varies across patients---a concept
+described as heterogeneity of treatment effects (HTE)---has been central to the
 agenda for both personalized (or precision) medicine and comparative
@@
 
-More recently, “predictive” HTE analyses have been described (and contrasted
-with “one-variable-at-a-time” subgroup analysis) as approaches that provide
-predictions of potential outcomes in a particular patient with one intervention
-versus an alternative, taking into account multiple relevant patient
-characteristics [@Varadhan2013]. One promising approach is “risk modeling”, in
-which treatment effects are estimated in strata of predicted risk [@Kent2010;
-@Kent2018]. Such a risk-based approach first stratifies patients according to
-baseline risk predictions, using either an existing or an internally developed
-risk prediction model [@Kent2010]. Then, relative and absolute treatment
-effects are estimated within risk strata.
+More recently, “predictive” HTE analyses have been described as approaches that
+provide predictions of potential outcomes in a particular patient with one
+intervention versus an alternative, taking into account multiple relevant
+patient characteristics [@Varadhan2013]. We focus on “risk modeling”
+approaches where patients are divided into risk strata using either an existing
+or an internally developed risk prediction model. Risk-stratum-specific
+estimates provide an overview of the evolution of treatment effects with
+increasing risk both on the relative and the absolute scale. Recently,
+systematic guidance on the application of such methods has been developed
+[@Kent2010; @Kent2018].
 
-While these approaches have generally been recommended for application to
-clinical trials, observational databases are also an appealing substrate.
+While these approaches were developed for application in randomized controlled
+trials (RCTs), observational databases are also an appealing substrate.
 Observational healthcare databases, such as administrative claims and electronic
@@
 They are also often larger than many typical trials, providing excellent power
-for HTE analysis, and include heterogeneous populations. However, unlike trials,
+for HTE analysis, including heterogeneous populations. However, unlike RCTs,
 treatment effects are subject to confounding while the unique structure of
@@
 We aimed to develop a framework for risk-based assessment of treatment effect
-heterogeneity in high-dimensional observational data. We implemented the
-framework using existing OHDSI methods for use in the OMOP-CDM, including the
-patient-level prediction framework and the population-level effect estimation
-framework based on new-user cohort design [@Reps2018; @Ryan2013] . As a
-proof-of-concept we analyzed heterogeneity of the effects of first-line
-hypertension treatment: we compared the effect of angiotensin converting enzyme
-(ACE) inhibitors to beta blockers on 9 outcomes across three different US claims
-databases.
+heterogeneity in high-dimensional observational data, which extends the existing
+guidelines of the RCT setting. We implemented the framework using existing OHDSI
+methods for use in the OMOP-CDM, including the patient-level prediction
+framework and the population-level effect estimation framework based on new-user
+cohort design [@Reps2018; @Ryan2013]. As a proof-of-concept we analyzed
+heterogeneity of the effects of first-line hypertension treatment: we compared
+the effect of angiotensin converting enzyme (ACE) inhibitors to beta blockers
+on 9 outcomes across three different US claims databases.
 
@@
 absolute and relative treatment effects are estimated within risk strata; 5)
-presentation and evaluation of the results. A simple overview of the procedure
+presentation and evaluation of the results. A overview of the procedure
 can be seen in Figure \@ref(fig:figure1).
@@
 
+Note "cohort" refers to a set of persons who satisfy one or more criteria for a
+duration of time. The term can be used interchangeably with the term phenotype.
+
 ## Step 2: Identification of the database
@@
 To generate the target cohort we pool the already defined treatment cohort and
-comparator cohort. However, for risk-based analysis of treatment effects it is
-necessary to avoid deferentially fitting the prediction model to patients across
-the treatment arms in order to avoid inducing spurious interactions
-[@Burke2014; @vanKlaveren2019]. To do this, we develop the patient-level
-prediction model in the propensity score-matched subset of the population (1:1),
-where treatment assignment is well-balanced. The propensity scores are based on
-LASSO logistic regression for modeling the association between treatment
-assignment and all available demographics, drug exposures, diagnoses,
-measurements and medical procedures. Finally, we need to define the time horizon
-within which we aim to make predictions and we also need to select the
-machine-learning algorithm we want to use to generate patient-level predictions.
-Currently, the available options are regularized logistic regression, random
-forest, gradient boosting machines, decision tree, naive Bayes, K-nearest
-neighbors, neural network and deep learning (convolutional neural networks,
-recurrent neural network and deep nets).
+comparator cohort. To avoid deferentially fitting the prediction model to
+patients across treatment arms, thus introducing spurious interactions with
+treatment [@Burke2014; @vanKlaveren2019], we develop the patient-level
+prediction model in the propensity score-matched (1:1) subset of the population.
+Finally, we need to define the time horizon within which we aim to make
+predictions and we also need to select the machine-learning algorithm we want to
+use to generate patient-level predictions. Currently, the available options are
+regularized logistic regression, random forest, gradient boosting machines,
+decision tree, naive Bayes, K-nearest neighbors, neural network and deep
+learning (convolutional neural networks, recurrent neural network and deep
+nets).
 
@@
 
-We use the patient-level prediction model to divide the target population into a
-set of equally-sized risk strata, typically 4 risk quarters. Then, we estimate
-propensity scores within risk strata. These propensity scores are used when
-estimating treatment effects, either by matching of patients from different
-treatment cohorts, by stratification of patients into groups with similar
-propensity scores, or by weighing patients’ contribution to the estimation
-process. Within risk strata we estimate treatment effect both on the relative
-and the absolute scale. It is important to evaluate treatment effects in both
-scales, as effect cannot remain constant on both the relative and the absolute
-scale at the same time, assuming a non-zero treatment effect. Any appropriate
-method for the evaluation of relative and absolute treatment effects can be
-considered, as long as the this is done consistently in all risk strata.
+The aim of this step is the estimation of treatment effects (both on the
+relative and the absolute scale) within risk strata---typically 4 risk
+quarters---defined using the prediction model of step 3. Any appropriate method
+for the evaluation of relative and absolute treatment effects can be considered,
+as long as the this is done consistently in all risk strata. Common approaches
+are odds ratios or hazard ratios for the relative scale and differences in
+proportions or differences in Kaplan-Meier estimates on the absolute scale,
+depending on the problem at hand. We estimate propensity scores within risk
+strata which we then use to match patients from different treatment cohorts
+or stratify them into groups with similar propensity scores or to weigh
+each patient's contribution to the estimation process [@Austin2011].
 
@@
 
-The number of patients and person years by treatment arm along with the number
-of outcomes should be reported. A performance overview of the derived prediction
-models, including discrimination and calibration both in the propensity score
-matched subset, the entire population and separately for treated and comparator
-patients should also be reported. This is quite relevant as the performance of
-the prediction models is directly related to our ability to single out patient
-subgroups where treatment may be highly beneficial or unsafe. Kent et al
-[@Kent2016] demonstrated that the event rate and the discriminative ability of
-the prediction model can predict very well the distribution of predicted risk.
-Lower event rate and higher c-statistic (given good calibration) result in high
-risk heterogeneity, thus making estimated average treatment effects
-uninformative. In this case, risk stratified analysis of HTE can be more
-effective in singling out patient subgroups that stand to benefit (or be harmed)
-most by treatment in question.
+In the presence of a positive treatment effect and a well-discriminating
+prediction model we expect an increasing pattern of the differences in the
+absolute scale, even if the relative effects remain constant on the relative
+scale across risk strata. Results should be presented side-by-side on a
+forest-plot-like format, so that the evolution of treatment effects across risk
+strata is visible both on the relative and the absolute scale.
 
-In every risk stratum propensity score distributions by treatment group and
-covariate balance plots should be made available along with event rates, hazard
-ratios and absolute risk differences. Finally, the results should be generated
-with a standardized format to allow for combination of results from multiple
-data sources. This also enables the development of scalable demonstration tools
-such as R-shiny applications, interactive tables and figures.
+A performance overview of the derived prediction models, including
+discrimination and calibration both in the propensity score matched subset, the
+entire population and separately for treated and comparator patients should also
+be reported. This is quite relevant as the performance of the prediction models
+is directly related to our ability to single out patient subgroups where
+treatment may be highly beneficial or unsafe. Kent et al [@Kent2016]
+demonstrated that the event rate and the discriminative ability of the
+prediction model can predict very well the distribution of predicted risk. Lower
+event rate and higher c-statistic (given good calibration) result in high risk
+heterogeneity, thus making estimated average treatment effects uninformative. In
+this case, risk stratified analysis of HTE can be more effective in singling out
+patient subgroups that stand to benefit (or be harmed) most by treatment in
+question.
+
+In each risk stratum we need to evaluate if adequate covariate balance was
+achieved accounting for measured confounding. If that is not the case,
+interpretation of the results may be problematic. Common approaches include
+evaluation of the overlap of propensity score distributions and calculation of
+standardized covariate differences before and after propensity score adjustment.
 
@@
 
-We demonstrate the framework with the following research aim: "to compare the
-effect of ACE-inhibitors ($T$) to the effect of beta blockers ($C$) in patients
-with established hypertension ($D$) with respect to 9 outcomes
-($O_1,\dots,O_9$)". The cohorts are:
+We consider the following research aim: "compare the effect of ACE-inhibitors
+($T$) to the effect of beta blockers ($C$) in patients with established
+hypertension ($D$) with respect to 9 outcomes ($O_1,\dots,O_9$)". The cohorts
+are:
 
@@
 
-To obtain a target cohort for developing patient-level predictions we first
-merged the ACE-inhibitors cohort with the beta blockers cohort. We then matched
-patients in the ACE-inhibitor cohort to patients in the beta blockers cohort on
-the propensity score. We focused on the efficacy outcomes (acute MI,
-hospitalization with heart failure and hemorrhagic or ischemic stroke) for risk
-stratification of the patient population. In each database, for each main
-outcome, we developed a different prediction model. We chose a time horizon of 2
-years after inclusion into the target cohort. We developed the prediction models
-using LASSO logistic regression with 3-fold cross validation for hyper-parameter
-selection.
+We developed separate prediction models for all efficacy outcomes (acute MI,
+hospitalization with heart failure and hemorrhagic or ischemic stroke) in each
+database. More specifically, we first estimated propensity scores using LASSO
+logistic regression and a large set of baseline covariates including
+demographics, drug exposures, diagnoses, measurements and medical devices. The
+prediction models were estimated on the propensity score matched (1:1) subset of
+the population, using caliper of 0.2 and after excluding patients having the
+outcomes any time prior to treatment initiation. We chose a time horizon of 2
+years after inclusion into the target cohort. For this demonstration, we
+developed the prediction models using LASSO logistic regression with 3-fold
+cross validation for hyper-parameter selection. We chose a time horizon of 2
+years after inclusion into the target cohort. For this demonstration, we
+developed the prediction models using LASSO logistic regression with 3-fold
+cross validation for hyper-parameter selection.
 
@@
 We used patient-level predictions to stratify the patient population into 4 risk
-quarters. We estimated relative and absolute treatment effects for all 9
-outcomes of interest. In order to estimate risk quarter-specific treatment
-effects we first estimated propensity scores within risk quarters. We then used
-the propensity scores to stratify patients into 5 strata.
+quarters. Within risk strata, relative effects were estimated using Cox
+regression and absolute effects were estimated from the Kaplan-Meier estimate
+differences at 2 years after treatment initiation. To adjust for observed
+confounding within risk strata, we estimated propensity scores using the same
+approach as in the development of prediction models. We used the estimated
+propensity scores to stratify patients into 5 strata, within each risk quarter.
 
@@
 
-Here we present the results of the analysis in the CCAE database with
+We present the results of the analysis in the CCAE database with
 stratification based on risk predictions of acute MI. Results of analyses in the
@@
 
-Relative treatment effects of ACE-inhibitors vs beta blockers increased (hazard
-ratios decreased) with increasing acute MI risk, resulting in more pronounced
-absolute risk differences (ARD) with increasing acute MI risk (Figure
-\@ref(fig:figure2)).
+Relative treatment effects of ACE-inhibitors vs beta blockers increased in favor
+of ACE-inhibitors (hazard ratios decreased) with increasing acute MI risk,
+resulting in more pronounced absolute risk differences (ARD) with increasing
+acute MI risk (Figure \@ref(fig:figure2)).
 
@@
 
-Patients in the low risk quarter did not receive absolute benefit (-0.03%)
-while absolute risk was 0.54% lower (95% confidence interval 0.36%—0.71%) for
-patients in the high-risk quarter. In contrast, the absolute and relative
-effects of ACE-inhibitors on safety outcomes (e.g. cough and angioedema) are
-approximately constant or even slightly decreasing with increasing acute MI risk
-(Figure \@ref(fig:figure3) and \@ref(fig:figure4)). Similar results were observed
-in the other two databases (see supplementary material).
+In general, with increasing acute MI risk we observed an increasing pattern in
+terms of absolute benefit for the main outcomes, while the absolute harms did
+not increase for the safety outcomes. More specifically, patients in the low
+risk quarter did not receive absolute benefit (-0.03%) while absolute risk
+was 0.54% lower (95% confidence interval 0.36%—0.71%) for patients in the
+highest risk quarter. In contrast, the absolute and relative effects of
+ACE-inhibitors on safety outcomes (e.g. cough and angioedema) are slightly
+decreasing with increasing acute MI risk (Figure \@ref(fig:figure3) and
+\@ref(fig:figure4)). Similar results were observed in the other two databases
+(see supplementary material).
 
@@
 
-This example nicely illustrates heterogeneity of absolute treatment effects,
-i.e. differences in absolute benefits and harms of ACE-inhibitors vs beta
-blockers for patients with different baseline risk. The results suggest that
-treatment with ACE-inhibitors, compared to treatment with beta blockers, may be
-focused on the higher risk patients, in whom the benefits outweigh the harms.
-However, treatment with beta blockers may be a viable option in lower risk
-patients, in whom the benefit-harm tradeoff is in favor of beta blockers. This
-is in accordance with earlier findings that beta blockers should also be
-considered as first-line treatment for younger hypertensive patients
-[@Cruickshank2007; @Khan2006]. More thorough evaluation of these results is
-required in future research, however.
+These results suggest that treatment with ACE-inhibitors, compared to treatment
+with beta blockers, may be focused on the higher risk patients, in whom the
+benefits outweigh the harms, while beta blockers may be a viable option in lower
+risk patients, in whom the benefit-harm tradeoff is more favorable. This is in
+accordance with earlier findings that beta blockers should also be considered as
+first-line treatment for younger hypertensive patients [@Cruickshank2007;
+@Khan2006]. More thorough evaluation of these results is required in future
+research, however.
 
@@
 provides a standardized approach that can be used to leverage available
-information from these data sources, allowing for large-scale assessment of
-treatment effect heterogeneity. Multiple outcomes can be evaluated in patient
-subgroups of similar baseline outcome risk. Different outcome risk
+information from these data sources, allowing for large-scale risk-based
+assessment of treatment effect heterogeneity. Multiple outcomes can be evaluated
+in patient subgroups of similar baseline outcome risk. Different outcome risk
 stratification schemes can also be considered. The standardized nature of the
@@
 risk. These scores are estimated using regularized logistic regression on a
-large set of pre-defined covariates. This specific approach gave accurate
-results in extensive simulation studies [@Tian2018]. However, such approaches
-do not account for unobserved confounding [@Liu2013]. Several sensitivity
-analyses have been proposed in the literature for measuring the robustness of
-results in the presence of unobserved confounding. Another approach is to
-calibrate estimates and confidence intervals based on a large set of negative
-controls [@Schuemie2014; @Schuemie2571]. Negative controls are
-treatment-outcome pairs for which a null effect has been established. Estimating
-these effects within available data provides an approximation of the null
-distribution that can be used to empirically recalibrate effect estimates.
-Future work may extend our framework with this type of analyses.
+large set of pre-defined covariates. However, such approaches do not account for
+unobserved confounding [@Liu2013]. Several sensitivity analyses have been
+proposed in the literature for measuring the robustness of results in the
+presence of unobserved confounding. Another approach is to calibrate estimates
+and confidence intervals based on a large set of negative controls
+[@Schuemie2014; @Schuemie2571]. Negative controls are treatment-outcome pairs
+for which a null effect has been established. Estimating these effects within
+available data provides an approximation of the null distribution that can be
+used to empirically recalibrate effect estimates. Future work may extend our
+framework with this type of analyses.
 
@@
 heterogeneity. However, even though stratification can provide a useful overview
-for clinical interpretation, it is not appropriate to guide clinical practice,
-where decisions need to be made at the individual level [@Kent2019].
+for clinical interpretation, these results cannot be applied to individuals in a straightforward manner, as we are still estimating subgroup effects [@Kent2019].
 Presentation of treatment effects as a continuous function of risk would be more
@@
 methods library [@Kent2019; @PathEnE]. Our method encourages open science as
-it requires clear definition of the research questions translated into clear and
-reproducible cohort definitions that can easily be shared among researchers. Our
-R-package provides a standardized stepwise procedure for the assessment of HTE.
-This enables source code to be easily shared and evaluated, allowing for
+it requires accurate definition of the research questions translated into clear
+and reproducible cohort definitions that can easily be shared among researchers.
+Our R-package provides a standardized stepwise procedure for the assessment of
+HTE. This enables source code to be easily shared and evaluated, allowing for
 reproducible research. Researchers with access to different databases mapped to

### 2021-05-05 submission/manuscript.rmd
typo
new version of manuscript

@@
 \vspace{10mm}
-**Corresponing author**
+**Corresponding author**
 \singlespacing 
@@
 
-More recently, “predictive” HTE analyses have been described as approaches that
+Predictive HTE analyses have been described as approaches that
 provide predictions of potential outcomes in a particular patient with one
@@
 for HTE analysis, including heterogeneous populations. However, unlike RCTs,
-treatment effects are subject to confounding while the unique structure of
-different databases calls for database-specific analysis plans that are often
+treatment effects are subject to confounding, while the unique structure of
+different databases calls for database-specific analysis plans that often are
 not easily transportable.
@@
 Common Data Model (CDM) [@hripcsak2015observational; @Overhage2012]. The
-common data structure enables analyses at a very large scale. For example, in a
-recent study, a large set of first-line treatments for hypertension was compared
-with respect to 55 outcomes in a network of databases, including 4.9 million
-patients from around the world [@Suchard2019].
-
-We aimed to develop a framework for risk-based assessment of treatment effect
-heterogeneity in high-dimensional observational data, which extends the existing
-guidelines of the RCT setting. We implemented the framework using existing OHDSI
-methods for use in the OMOP-CDM, including the patient-level prediction
-framework and the population-level effect estimation framework based on new-user
-cohort design [@Reps2018; @Ryan2013]. As a proof-of-concept we analyzed
-heterogeneity of the effects of first-line hypertension treatment: we compared
-the effect of angiotensin converting enzyme (ACE) inhibitors to beta blockers
-on 9 outcomes across three different US claims databases.
+common data structure enables analyses at a very large scale. We aimed to
+develop a framework for risk-based assessment of treatment effect heterogeneity
+in high-dimensional observational data, which extends the existing guidelines of
+the RCT setting. We implemented the framework using existing OHDSI methods for
+use in the OMOP-CDM, including the patient-level prediction framework and the
+population-level effect estimation framework based on new-user cohort design
+[@Reps2018; @Ryan2013]. As a proof-of-concept we analyzed heterogeneity of the
+effects of first-line hypertension treatment: we compared the effect of
+angiotensin converting enzyme (ACE) inhibitors to beta blockers on 9 outcomes
+across three different US claims databases.
 
@@
 absolute and relative treatment effects are estimated within risk strata; 5)
-presentation and evaluation of the results. A overview of the procedure
+presentation and evaluation of the results. An overview of the procedure
 can be seen in Figure \@ref(fig:figure1).
@@
 as long as the this is done consistently in all risk strata. Common approaches
-are odds ratios or hazard ratios for the relative scale and differences in
-proportions or differences in Kaplan-Meier estimates on the absolute scale,
-depending on the problem at hand. We estimate propensity scores within risk
-strata which we then use to match patients from different treatment cohorts
-or stratify them into groups with similar propensity scores or to weigh
+are odds ratios or hazard ratios for relative scale estimates and differences in
+observed proportions or differences in Kaplan-Meier estimates for absolute scale
+estimates, depending on the problem at hand. We estimate propensity scores
+within risk strata which we then use to match patients from different treatment
+cohorts or stratify them into groups with similar propensity scores or to weigh
 each patient's contribution to the estimation process [@Austin2011].
@@
 prediction model we expect an increasing pattern of the differences in the
-absolute scale, even if the relative effects remain constant on the relative
-scale across risk strata. Results should be presented side-by-side on a
+absolute scale, even if treatment effects remain constant on the relative scale
+across risk strata. Results should be presented side-by-side on a
 forest-plot-like format, so that the evolution of treatment effects across risk
@@
 As a proof of concept, we focus on the comparison of angiotensin converting
-enzyme (ACE) inhibitors to beta blockers. ACE inhibitors are among the most
-common treatment classes for hypertension, with well-established effectiveness.
-Beta blockers, even though initially widely used for the treatment of
-hypertension, more recent trials and meta-analyses have cast doubt on their
-relative effectiveness [@wiysonge2017beta]. As a result, newer US guidelines
-do not consider them for initial treatment for hypertension while in the EU
-guidelines combination with other antihypertensive treatments is recommended
+enzyme (ACE) inhibitors to beta blockers are among the most common treatment
+classes for hypertension, with well-established effectiveness. Beta blockers,
+even though initially widely used for the treatment of hypertension, more recent
+trials and meta-analyses have cast doubt on their relative effectiveness
+[@wiysonge2017beta]. As a result, newer US guidelines do not consider beta
+blockers for initial treatment for hypertension while in the EU guidelines
+combination with other antihypertensive treatments is recommended
 [@Whelton2018; @Williams2018]. However, another meta-analysis suggested that
@@
 
-All cohort definitions can be found in the supplementary material.
+All cohort definitions were identical to the ones used in the multinational
+study carried out within OHDSI that provided overall treatment effect estimates
+comparing all anti-hypertensive drug classes with each other [@Suchard2019].
+More information can be found in the supplementary material.
 
@@
 However, low power and restricted prior knowledge on the mechanisms of variation
-in treatment effect are often inherent in RCTs, which are often adequately
+in treatment effect are often inherent in RCTs, which are usually adequately
 powered only for the analysis of the primary outcome. Observational databases
@@
 assessment of treatment effect heterogeneity. Multiple outcomes can be evaluated
-in patient subgroups of similar baseline outcome risk. Different outcome risk
-stratification schemes can also be considered. The standardized nature of the
-framework enables transportability to multiple databases, provided that they are
-mapped to the OMOP-CDM.
+in patient subgroups of similar baseline outcome risk. Multiple outcome risk
+stratification schemes can also be considered. However, this should be done with
+caution, as it may hinder the interpretability of the results, in a similar
+manner as typical subgroup analyses. The standardized nature of the framework
+enables transportability to multiple databases, provided that they are mapped to
+the OMOP-CDM.
 
@@
 heterogeneity. However, even though stratification can provide a useful overview
-for clinical interpretation, these results cannot be applied to individuals in a straightforward manner, as we are still estimating subgroup effects [@Kent2019].
-Presentation of treatment effects as a continuous function of risk would be more
-helpful, but is methodologically challenging. Future research is necessary for
-the development of methods for continuous risk-based assessment of HTE.
+for clinical interpretation, these results cannot be applied to individuals in a
+straightforward manner, as we are still estimating subgroup effects
+[@Kent2019]. Presentation of treatment effects as a continuous function of
+risk would be more helpful, but is methodologically challenging. Future research
+is necessary for the development of methods for continuous risk-based assessment
+of HTE.
 
@@
 may significantly differ in sample size. Because the prediction model will
-possibly better fit to the larger treatment arm, this may introduce spurious
-treatment-covariate interactions in the prediction model, leading to sub-optimal
+possibly better fit to the larger treatment arm, spurious treatment-covariate
+interactions may be introduced in the prediction model, leading to sub-optimal
 risk stratification. As a remedy, we first match the patients in the treatment

### 2021-05-05 submission/manuscript.rmd
minor changes in discussion

@@
 caution, as it may hinder the interpretability of the results, in a similar
-manner as typical subgroup analyses. The standardized nature of the framework
-enables transportability to multiple databases, provided that they are mapped to
-the OMOP-CDM.
+manner as typical subgroup analyses.
 
@@
 proposed [@Kent2019; @PathEnE]. Our framework aims to translate these
-guidelines to the observational setting while also providing a toolset for its
-implementation. Several considerations need to be made. First, estimates may be
+guidelines to the observational setting while also providing a toolkit for its
+implementation within OHDSI. It encourages open science as it requires accurate
+definition of the research questions translated into clear and reproducible
+cohort definitions that can easily be shared among researchers. Researchers with
+access to different databases mapped to OMOP-CDM can also very easily extend
+their overall analyses with risk-based assessment of treatment effect
+heterogeneity. This enables collaboration among multiple sites with access to
+different patient populations. We propose that the framework is implemented any
+time treatment effect estimation in high-dimensional observational data is
+undertaken.
+
+Several considerations need to be made. First, estimates may be
 biased due to the observational nature of the data. We attempt to account for
@@
 
-Our contribution is a translation of the PATH statement principles to the OHDSI
-methods library [@Kent2019; @PathEnE]. Our method encourages open science as
-it requires accurate definition of the research questions translated into clear
-and reproducible cohort definitions that can easily be shared among researchers.
-Our R-package provides a standardized stepwise procedure for the assessment of
-HTE. This enables source code to be easily shared and evaluated, allowing for
-reproducible research. Researchers with access to different databases mapped to
-OMOP-CDM can also very easily extend their overall analyses with risk-based
-assessment of treatment effect heterogeneity. This enables collaboration among
-multiple sites with access to different patient populations. We propose that the
-framework is implemented any time treatment effect estimation in
-high-dimensional observational data is undertaken.
-
 Recently, disease risk scores have been explored as an alternative to propensity

### 2021-08-18 code/CombinedPlots.R
code to combine results from all databases

@@ -1,0 +1,152 @@
+library(tidyverse)
+
+mappedOverallRelativeResults <- readRDS(
+  "~/Documents/Projects/arekkas_HteFramework_XXXX_2021/.scratch/AceBeta9Outcomes/data/mappedOverallRelativeResults.rds"
+)
+mappedOverallAbsoluteResults <- readRDS(
+  "~/Documents/Projects/arekkas_HteFramework_XXXX_2021/.scratch/AceBeta9Outcomes/data/mappedOverallAbsoluteResults.rds"
+)
+map_outcomes <- readRDS(
+  "~/Documents/Projects/arekkas_HteFramework_XXXX_2021/.scratch/AceBeta9Outcomes/data/map_outcomes.rds"
+)
+map_exposures <- readRDS(
+  "~/Documents/Projects/arekkas_HteFramework_XXXX_2021/.scratch/AceBeta9Outcomes/data/map_exposures.rds"
+)
+
+prepareDataset <- function(data) {
+  data %>%
+    left_join(
+      map_outcomes,
+      by = c("stratOutcome" = "outcome_id")
+    ) %>%
+    select(-stratOutcome) %>%
+    rename("stratOutcome" = "outcome_name") %>%
+    left_join(
+      map_outcomes,
+      by = c("estOutcome" = "outcome_id")
+    ) %>%
+    select(-estOutcome) %>%
+    rename("estOutcome" = "outcome_name") %>%
+    left_join(
+      map_exposures,
+      by = c("treatment" = "exposure_id")
+    ) %>%
+    select(-treatment) %>%
+    rename("treatment" = "exposure_name") %>%
+    left_join(
+      map_exposures,
+      by = c("comparator" = "exposure_id")
+    ) %>%
+    select(-comparator) %>%
+    rename("comparator" = "exposure_name") %>%
+    mutate(
+      estOutcome = factor(
+        estOutcome,
+        rev(unique(estOutcome))
+      )
+    )
+}
+
+
+relative <- prepareDataset(mappedOverallRelativeResults) %>%
+  filter(
+    stratOutcome == "Acute myocardial infarction"
+  )
+
+ggplot(
+  data = relative,
+  aes(
+    y = estOutcome,
+    x = estimate,
+    color = database,
+    xmin = lower,
+    xmax = upper
+  )
+) +
+  geom_point(
+    position = position_dodge(.5)
+  ) +
+  geom_errorbar(
[…]

### 2021-08-19 code/CombinedPlots.R
add overall results to processed data

@@ -1,16 +1,16 @@
 library(tidyverse)
 
 mappedOverallRelativeResults <- readRDS(
-  "~/Documents/Projects/arekkas_HteFramework_XXXX_2021/.scratch/AceBeta9Outcomes/data/mappedOverallRelativeResults.rds"
+  "data/processed/mappedOverallRelativeResults.rds"
 )
 mappedOverallAbsoluteResults <- readRDS(
-  "~/Documents/Projects/arekkas_HteFramework_XXXX_2021/.scratch/AceBeta9Outcomes/data/mappedOverallAbsoluteResults.rds"
+  "data/processed/mappedOverallAbsoluteResults.rds"
 )
 map_outcomes <- readRDS(
-  "~/Documents/Projects/arekkas_HteFramework_XXXX_2021/.scratch/AceBeta9Outcomes/data/map_outcomes.rds"
+  "data/processed/map_outcomes.rds"
 )
 map_exposures <- readRDS(
-  "~/Documents/Projects/arekkas_HteFramework_XXXX_2021/.scratch/AceBeta9Outcomes/data/map_exposures.rds"
+  "data/processed/map_exposures.rds"
 )
 
 prepareDataset <- function(data) {

### 2021-08-19 submission/manuscript.rmd
working on reviewer 1 comments

@@
 
+< < ADD > >
+
+***Because of the latter issue, running analyses at scale demands a big investment
+of time and effort, as researchers are forced to map their analysis plans to the
+databases available to them.***
+
+< < ADD > >
+
 The Observational Health Data Sciences and Informatics (OHDSI) collaborative has
@@
 all available databases to the Observational Medical Outcomes Partnership (OMOP)
-Common Data Model (CDM) [@hripcsak2015observational; @Overhage2012]. The
-common data structure enables analyses at a very large scale. We aimed to
-develop a framework for risk-based assessment of treatment effect heterogeneity
+Common Data Model (CDM) [@hripcsak2015observational; @Overhage2012].
+
+< < EDIT > >
+
+***We aimed to develop a framework for implementing risk-based assessment of
+treatment heterogeneity in high-dimensional observational data, extending the
+existing guidelines of the RCT setting. Our contribution provides an
+out-of-the-box solution for implementing such analyses at scale within the OHDSI
+network, taking advantage of the OMOP-CDM.***
+
+< < EDIT > >
+
+< < OLD > >
+
+We aimed to develop a framework for risk-based assessment of treatment effect heterogeneity
 in high-dimensional observational data, which extends the existing guidelines of
-the RCT setting. We implemented the framework using existing OHDSI methods for
-use in the OMOP-CDM, including the patient-level prediction framework and the
-population-level effect estimation framework based on new-user cohort design
-[@Reps2018; @Ryan2013]. As a proof-of-concept we analyzed heterogeneity of the
-effects of first-line hypertension treatment: we compared the effect of
-angiotensin converting enzyme (ACE) inhibitors to beta blockers on 9 outcomes
-across three different US claims databases.
+the RCT setting.
+
+< < OLD > >
+
+We implemented the framework using existing OHDSI methods including the
+patient-level prediction framework and the population-level effect estimation
+framework based on new-user cohort design [@Reps2018; @Ryan2013]. As a
+proof-of-concept we analyzed heterogeneity of the effects of first-line
+hypertension treatment: we compared the effect of angiotensin converting enzyme
+(ACE) inhibitors to beta blockers on 9 outcomes across three different US claims
+databases.
 
@@
 prediction model in the propensity score-matched (1:1) subset of the population.
+
+< < MOVE: ps_matching > >
+
+< < ADD > >
+
+***Other methods of fitting the propensity scores, such as random forest and
+others (REF) can also be considered.***
+
+< < ADD > >
+
 Finally, we need to define the time horizon within which we aim to make
@@
 hospitalization with heart failure and hemorrhagic or ischemic stroke) in each
-database. More specifically, we first estimated propensity scores using LASSO
+database.
+
+< < MOVE: ps_matching > >
+
+***More specifically, we first estimated propensity scores using LASSO
 logistic regression and a large set of baseline covariates including
-demographics, drug exposures, diagnoses, measurements and medical devices. The
-prediction models were estimated on the propensity score matched (1:1) subset of
+demographics, drug exposures, diagnoses, measurements and medical devices.***
+
+< < MOVE > >
+
+The prediction models were estimated on the propensity score matched (1:1) subset of
 the population, using caliper of 0.2 and after excluding patients having the

### 2021-08-20 code/CombinedPlots.R
add combined plots across databases

@@ -148,5 +148,6 @@ ggplot(
     strip.background = element_rect(
       fill = "#ffffff"
     )
-  )
+  ) %>%
+  ggsave("figures/CombinedAbsolute.tif")
 

### 2021-08-20 code/CombinedPlots.R
resize plots

@@ -53,7 +53,7 @@ relative <- prepareDataset(mappedOverallRelativeResults) %>%
     stratOutcome == "Acute myocardial infarction"
   )
 
-ggplot(
+relativePlot <- ggplot(
   data = relative,
   aes(
     y = estOutcome,
@@ -93,10 +93,25 @@ ggplot(
     legend.position = "top",
     legend.title = element_blank(),
     axis.title.y = element_blank(),
+    axis.title.x = element_text(size = 20),
+    axis.text    = element_text(size = 15),
+    legend.text  = element_text(size = 15), 
     strip.background = element_rect(
       fill = "#ffffff"
-    )
+    ),
+    strip.text = element_text(size = 15)
   )
+
+ggsave(
+  "figures/CombinedRelative.tiff",
+  absolutePlot, 
+  compression = "lzw", 
+  width       = 600, 
+  height      = 300,
+  units       = "mm",
+  dpi         = 300
+)
+
 
 
 absolute <- prepareDataset(mappedOverallAbsoluteResults) %>%
@@ -105,7 +120,7 @@ absolute <- prepareDataset(mappedOverallAbsoluteResults) %>%
     estOutcome != "Cough"
   )
 
-ggplot(
+absolutePlot <- ggplot(
   data = absolute,
   aes(
     y = estOutcome,
@@ -145,9 +160,22 @@ ggplot(
     legend.position = "top",
     legend.title = element_blank(),
     axis.title.y = element_blank(),
+    axis.title.x = element_text(size = 20),
+    axis.text    = element_text(size = 15),
+    legend.text  = element_text(size = 15), 
     strip.background = element_rect(
       fill = "#ffffff"
-    )
-  ) %>%
-  ggsave("figures/CombinedAbsolute.tif")
+    ),
+    strip.text = element_text(size = 15)
+  )
+
+ggsave(
+  "figures/CombinedAbsolute.tiff",
+  absolutePlot, 
+  compression = "lzw", 
+  width       = 600, 
+  height      = 300,
+  units       = "mm",
+  dpi         = 300
+)
 

### 2021-08-20 code/CombinedPlots.R
fix bug and rebuild figure

@@ -104,7 +104,7 @@ relativePlot <- ggplot(
 
 ggsave(
   "figures/CombinedRelative.tiff",
-  absolutePlot, 
+  relativePlot, 
   compression = "lzw", 
   width       = 600, 
   height      = 300,

### 2021-08-24 submission/manuscript.rmd
take notes in preparation for rebuttal text

@@
 treatment heterogeneity in high-dimensional observational data, extending the
-existing guidelines of the RCT setting. Our contribution provides an
-out-of-the-box solution for implementing such analyses at scale within the OHDSI
-network, taking advantage of the OMOP-CDM.***
+existing guidelines of the RCT setting. Our publicly available package provides
+an out-of-the-box solution for implementing such analyses at scale within the
+OHDSI network, taking advantage of the OMOP-CDM.***
 

### 2021-08-25 submission/manuscript.rmd
responding to reviewer 1

@@
 
+\> > EDIT > >
+
+***In recent years, a large number of methods has been developed for the
+assessment of HTE, mainly in the RCT setting. Earlier work suggested separating
+HTE analyses into exploratory, confirmatory, descriptive and predictive
+[@Varadhan2013]. Exploratory analyses focus on hypothesis generation,
+confirmatory analyses test subgroup effect hypotheses, descriptive analyses aim
+at facilitating future synthesis of subgroup effects and predictive analyses
+predict probabilities of benefit or harm in individual patients. Predictive HTE
+approaches can be further subdivided into risk modeling, treatment effect
+modeling and optimal treatment regime methods, based on the reference class used
+for defining patient similarity when making individualized predictions or
+recommendations [@Rekkas2020]. We focus on “risk modeling” approaches where
+patients are divided into risk strata using either an existing or an internally
+developed risk prediction model. Risk-stratum-specific estimates provide an
+overview of the evolution of treatment effects with increasing risk both on the
+relative and the absolute scale. Recently, systematic guidance on the
+application of such methods has been developed [@Kent2010; @Kent2018].***
+
+< < EDIT < <
+
+\> > OLD > >
+
 Predictive HTE analyses have been described as approaches that
@@
 [@Kent2010; @Kent2018].
+
+< < OLD < <
 
@@
 
-< < ADD > >
+\> > ADD > >
 
@@
 
-< < ADD > >
+< < ADD < <
 
@@
 
-< < EDIT > >
+\> > EDIT > >
 
@@
 
-< < EDIT > >
+< < EDIT < <
 
-< < OLD > >
+\> > OLD > >
 
@@
 
-< < OLD > >
+< < OLD < <
 
@@
 
-## Step 5: Result presentation and evaluation
+## Step 5: Evaluation and presentation of results
 
-In the presence of a positive treatment effect and a well-discriminating
-prediction model we expect an increasing pattern of the differences in the
-absolute scale, even if treatment effects remain constant on the relative scale
-across risk strata. Results should be presented side-by-side on a
-forest-plot-like format, so that the evolution of treatment effects across risk
-strata is visible both on the relative and the absolute scale.
+In each risk stratum we need to evaluate if adequate covariate balance was
+achieved accounting for measured confounding. If that is not the case,
+interpretation of the results may be problematic. Common approaches include
+evaluation of the overlap of propensity score distributions and calculation of
+standardized covariate differences before and after propensity score adjustment.
 
@@
 
-In each risk stratum we need to evaluate if adequate covariate balance was
-achieved accounting for measured confounding. If that is not the case,
-interpretation of the results may be problematic. Common approaches include
-evaluation of the overlap of propensity score distributions and calculation of
-standardized covariate differences before and after propensity score adjustment.
+In the presence of a positive treatment effect and a well-discriminating
+prediction model we expect an increasing pattern of the differences in the
+absolute scale, even if treatment effects remain constant on the relative scale
+across risk strata. Results should be presented side-by-side on a
+forest-plot-like format, so that the evolution of treatment effects across risk
+strata is visible both on the relative and the absolute scale.
 
@@
 
-< < MOVE: ps_matching > >
+\> > MOVE: ps_matching > >
 
@@
 
-< < MOVE > >
+< < MOVE < <
 

### 2021-08-26 submission/manuscript.rmd
add responses to reviewer 1
more response work on reviewers 1 and 2

@@
 \singlespacing 
-One of the aims of the Observational Health Data Sciences and Informatics (OHDSI)
-initiative is population-level treatment effect estimation in large
+One of the aims of the Observational Health Data Sciences and Informatics
+(OHDSI) initiative is population-level treatment effect estimation in large
 observational databases. Since treatment effects are well-known to vary across
 groups of patients with different baseline risk, we aimed to extend the OHDSI
-methods library with a framework for risk-based assessment of treatment effect
-heterogeneity. The proposed framework consists of five steps: 1) definition of
-the problem, i.e. the population, the treatment, the comparator and the
-outcome(s) of interest; 2) identification of relevant databases; 3) development
-of a prediction model for the outcome(s) of interest; 4) estimation of
-propensity scores within strata of predicted risk and estimation of relative and
-absolute treatment effect within strata of predicted risk; 5) evaluation and
-presentation of results. We demonstrate our framework by evaluating
-heterogeneity of the effect of angiotensin-converting enzyme (ACE) inhibitors
-versus beta blockers on a set of 9 outcomes of interest across three
+library of open-source tools with a framework for risk-based assessment of
+treatment effect heterogeneity. The proposed framework consists of five
+steps: 1) definition of the problem, i.e. the population, the treatment, the
+comparator and the outcome(s) of interest; 2) identification of relevant
+databases; 3) development of a prediction model for the outcome(s) of
+interest; 4) estimation of propensity scores within strata of predicted risk and
+estimation of relative and absolute treatment effect within strata of predicted
+risk; 5) evaluation and presentation of results. We demonstrate our framework by
+evaluating heterogeneity of the effect of angiotensin-converting enzyme (ACE)
+inhibitors versus beta blockers on a set of 9 outcomes of interest across three
 observational databases. With increasing risk of acute myocardial infarction we
@@
 to highest risk groups. Cough-related absolute harms decreased from 4.1%
-to 2.6%. The proposed framework may be useful for the evaluation of heterogeneity
-of treatment effect on observational data that are mapped to the OMOP Common
-Data Model. The proof of concept study demonstrates its feasibility in large
-observational data. Further insights may arise by application to safety and
-effectiveness questions across the global data network.
+to 2.6%. The proposed framework may be useful for the evaluation of
+heterogeneity of treatment effect on observational data that are mapped to the
+OMOP Common Data Model. The proof of concept study demonstrates its feasibility
+in large observational data. Further insights may arise by application to safety
+and effectiveness questions across the global data network.
 
@@
 absolute and relative treatment effects are estimated within risk strata; 5)
-presentation and evaluation of the results. An overview of the procedure
-can be seen in Figure \@ref(fig:figure1).
+presentation and evaluation of the results. We developed an open-source
+R-package for the implementation of the proposed framework and made it publicly
+available. The source code can be found at
+<https://github.com/OHDSI/RiskStratifiedEstimation>. An overview of the
+procedure can be seen in Figure \@ref(fig:figure1).
 
@@
 comparator treatment $C$ in patients with disease $D$ with respect to outcomes
-$O_1,\dots,O_n$". At least three cohorts are defined:
+$O_1,\dots,O_n$".
+
+\> > ADD > >
+
+***Cohort definitions are crucial for this step of the framework. We define a
+cohort as the set of patients who satisfy one or more inclusion criteria for a
+duration of time. A cohort within the OHDSI setting is more than a set of
+specific clinical codes, providing a definition of a logic for how to use that
+code set. All cohort definitions consist of: an entry event, i.e. the time a
+patient enters a cohort; a set of inclusion criteria applied to the initial
+event cohort to further restrict the set of people, resulting in the
+construction of the construction of the qualifying cohort; cohort exit criteria
+that terminate the patient's presence in the cohort. Cohort definitions are
+transportable, meaning they can be implemented in any database, provided that it
+is mapped to the OMOP-CDM.***
+
+< < ADD < <
 
@@
 
-Note "cohort" refers to a set of persons who satisfy one or more criteria for a
-duration of time. The term can be used interchangeably with the term phenotype.
-
 ## Step 2: Identification of the database
@@
 The aim of this step is the inclusion of databases that represent the patient
-population of interest. The inclusion of multiple databases potentially
-increases the generalizability of results. Furthermore, the cohorts should
-preferably have adequate sample size to ensure precise effect estimation, even
-within smaller risk strata.
+population of interest. It is required that the databases are mapped to the
+OMOP-CDM. The inclusion of multiple databases potentially increases the
+generalizability of results. Furthermore, the cohorts should preferably have
+adequate sample size to ensure precise effect estimation, even within smaller
+risk strata.
 
@@
 guidelines [@Collins2015; @Moons2015]. This prediction framework requires the
-definition of two essential cohorts: a target cohort and an outcome cohort.
+definition of two essential cohorts: a target cohort, i.e. a set of patients
+that satisfy one or more inclusion criteria for a duration of time, and an
+outcome cohort.
 
 To generate the target cohort we pool the already defined treatment cohort and
-comparator cohort. To avoid deferentially fitting the prediction model to
-patients across treatment arms, thus introducing spurious interactions with
-treatment [@Burke2014; @vanKlaveren2019], we develop the patient-level
+comparator cohort. Further restrictions can be applied on the target cohort to
+construct the final population on which the prediction model will be developed
+(e.g. exclude patients with a prior outcome in their history, before being
+included in the target cohort). To avoid deferentially fitting the prediction
+model to patients across treatment arms, thus introducing spurious interactions
+with treatment [@Burke2014; @vanKlaveren2019], we develop the patient-level
 prediction model in the propensity score-matched (1:1) subset of the population.
@@
 
-***More specifically, we first estimated propensity scores using LASSO
+***More specifically, we first estimate propensity scores using LASSO
 logistic regression and a large set of baseline covariates including
-demographics, drug exposures, diagnoses, measurements and medical devices.***
+demographics, drug exposures, diagnoses, measurements and medical devices. We
+match patients 1-1 using a caliper, i.e. the maximum distance that is acceptable
+for any match. The default value we use is 0.2 on the standardized logit scale
+for the propensity scores.***
 
@@
 
+\> > EDIT > >
+
+***Ideally, externally derived and adequately validated prediction model would
+be preferred for analyzing treatment effect heterogeneity [@Kent2010]. In the
+absence of such prediction models an internally-developed risk prediction model
+can be considered. Earlier simulations of RCT studies have shown that internal
+models developed on the combined treatment and control arms blinded to treatment
+gave relatively unbiased estimates of treatment effect across the spectrum of
+risk [@Burke2014].***
+
+< < EDIT < <
+
+\> > OLD > >
+
 Externally derived and well transportable prediction models are preferred for
@@
 used to provide unbiased estimates of treatment effect across the spectrum of
-baseline risk [@Burke2014]. However, in observational databases treatment arms
+baseline risk [@Burke2014].
+
+< < OLD < <
+
+However, in observational databases treatment arms
 may significantly differ in sample size. Because the prediction model will

### 2021-08-31 submission/manuscript.rmd
separate evaluation from results in methods section
rewrite presentation of results

@@
 
-## Step 5: Evaluation and presentation of results
+## Step 5: Evaluation
 
@@
 
+## Step 6: Presentation of results
+
 In the presence of a positive treatment effect and a well-discriminating
@@
 absolute scale, even if treatment effects remain constant on the relative scale
-across risk strata. Results should be presented side-by-side on a
-forest-plot-like format, so that the evolution of treatment effects across risk
-strata is visible both on the relative and the absolute scale.
+across risk strata. Due to this scale-dependence of treatment effect
+heterogeneity, results should be assessed both on the relative and the absolute
+scale. We find that a side-by-side presentation on a forest-like format can give
+a very good representation of our results.
 

### 2021-09-02 submission/manuscript.rmd
remove evaluation section and distribute to prediction and estimation sections

@@
 that terminate the patient's presence in the cohort. Cohort definitions are
-transportable, meaning they can be implemented in any database, provided that it
-is mapped to the OMOP-CDM.***
+transportable, meaning that in theory they can be implemented in any database,
+provided that it is mapped to the OMOP-CDM.***
 
@@
 
+After model development, a performance overview of the derived prediction models
+including discrimination and calibration both in the propensity score matched
+subset, the entire population and separately for treated and comparator patients
+should also be reported. This is important to ensure that no overfitting of the
+prediction model in one of the cohorts has occurred. In addition, the
+performance of the prediction models is directly related to our ability to
+single out patient subgroups where treatment may be highly beneficial or unsafe.
+Kent et al [@Kent2016] demonstrated that the event rate and the discriminative
+ability of the prediction model can predict very well the distribution of
+predicted risk. Lower event rate and higher c-statistic (given good calibration)
+result in high risk heterogeneity, thus making estimated average treatment
+effects uninformative. In this case, risk stratified analysis of HTE can be more
+effective in singling out patient subgroups that stand to benefit (or be harmed)
+most by treatment in question.
+
 ## Step 4: Estimation
@@
 
-## Step 5: Evaluation
+Before focusing on the results of the estimation process we need to evaluate if
+adequate covariate balance was achieved within each risk stratum accounting for
+measured confounding. Common approaches include evaluation of the overlap of
+propensity score distributions and calculation of standardized covariate
+differences before and after propensity score adjustment.
 
-In each risk stratum we need to evaluate if adequate covariate balance was
-achieved accounting for measured confounding. If that is not the case,
-interpretation of the results may be problematic. Common approaches include
-evaluation of the overlap of propensity score distributions and calculation of
-standardized covariate differences before and after propensity score adjustment.
-
-A performance overview of the derived prediction models, including
-discrimination and calibration both in the propensity score matched subset, the
-entire population and separately for treated and comparator patients should also
-be reported. This is quite relevant as the performance of the prediction models
-is directly related to our ability to single out patient subgroups where
-treatment may be highly beneficial or unsafe. Kent et al [@Kent2016]
-demonstrated that the event rate and the discriminative ability of the
-prediction model can predict very well the distribution of predicted risk. Lower
-event rate and higher c-statistic (given good calibration) result in high risk
-heterogeneity, thus making estimated average treatment effects uninformative. In
-this case, risk stratified analysis of HTE can be more effective in singling out
-patient subgroups that stand to benefit (or be harmed) most by treatment in
-question.
-
-## Step 6: Presentation of results
+## Step 5: Presentation of results
 

### 2021-09-02 submission/manuscript.rmd
separate creation of Table 1

@@
 \captionsetup{width=\textwidth}
-```{r table, echo=FALSE, results='asis', booktabs=TRUE}
-table <- tibble(
-  outcome = c(
-    "Acute myocardial infarction",
-    rep("", 3),
-    "Heart failure (hosp)",
-    rep("", 3),
-    "Stroke (ischemic or hemorrhagic)",
-    rep("", 3)
-  ),
-  riskQuarter = rep(1:4, 3),
-  patientsAce = c(
-    161099, 204882, 214413, 204167,
-    146259, 188006, 218052, 230226,
-    146069, 187524, 217070, 226128
-  ),
-  personYearsAce = c(
-    276171, 372197, 393583, 351727,
-    249809, 341014, 399394, 400330,
-    294484, 340234, 397830, 393861
-  ),
-  eventsAce = c(
-    203, 534, 117, 2095,
-    228, 457, 826, 2012,
-    299, 554, 947, 1718
-  ),
-  patientsBeta = c(
-    133977, 90193, 80662, 90908,
-    126387, 84280, 83421, 98380,
-    126264, 84000, 83038, 97628
-  ),
-  personYearsBeta = c(
-    220633, 169231, 150035, 154419,
-    206706, 158425, 155222, 169139,
-    206453, 157913, 154587, 167810
-  ),
-  eventsBeta = c(
-    135, 321, 535, 1520,
-    378, 340, 570, 1773,
-    320, 351, 521, 1077
-  )
-)
+```{r table, echo=FALSE, warning=FALSE, message=FALSE, results='asis', booktabs=TRUE}
+table <- readr::read_csv(
+  here::here("data/processed/table1.csv"),
+  na = character()
+) %>%
+  select(-database)
 
@@
   ),
-  caption = "Number of patients, person years and events within quarters of predicted risk for hospitalization with heart failure for the 3 main outcomes of the study (acute myocardial infarction, hospitalization with heart failure and ischemic or hemorrhagic stroke)."
+  caption = "Number of patients, person years and events within quarters of predicted risk for hospitalization with heart failure for the 3 main outcomes of the study (acute myocardial infarction, hospitalization with heart failure and ischemic or hemorrhagic stroke) across the 3 databases."
 ) %>%
+  kableExtra::group_rows("CCAE", 1, 12) %>%
+  kableExtra::group_rows("MDCD", 13, 24) %>%
+  kableExtra::group_rows("MDCR", 25, 36) %>%
   kableExtra::add_header_above(
@@
     )
+  )
+```
+
+```{r tableAucs, echo=FALSE, results='asis', booktabs=TRUE}
+table <- tibble(
+  population = c(
+    "Matched set",
+    "Treatment cohort",
+    "Comparator cohort",
+    "Entire population"
+  ),
+  ccae = c(
+    75.78, 70.65, 79.1, 74.06
+  ),
+  mdcd = c(
+    78.12, 78.29, 80.26, 79.46
+  ),
+  mdcr = c(
+    71.09, 69.03, 68.53, 69.07
+  )
+)
+
+knitr::kable(
+  table,
+  format = "latex",
+  longtable = TRUE,
+  booktabs = TRUE,
+  col.names = c(
+    "Population", "CCAE", "MDCD", "MDCR"
+  ),
+   format.args = list(
+    big.mar = ","
+  ),
+  caption = "Discriminiative ability (c-statistic) of the derived prediction models for acute myocardial infarction in the matched set (development set), the treatment cohort, the comparator cohort and the etnire population in CCAE, MDCD and MDCR"
 )

### 2021-09-02 code/CombinedPlots.R

@@ -1,3 +1,5 @@
+#!/usr/bin/env Rscript
+
 library(tidyverse)
 
 mappedOverallRelativeResults <- readRDS(
@@ -116,8 +118,8 @@ ggsave(
 
 absolute <- prepareDataset(mappedOverallAbsoluteResults) %>%
   filter(
-    stratOutcome == "Acute myocardial infarction",
-    estOutcome != "Cough"
+    stratOutcome == "Acute myocardial infarction"
+    # estOutcome != "Cough"
   )
 
 absolutePlot <- ggplot(

### 2021-09-02 code/Table1.R

@@ -1,0 +1,243 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+
+tableCcae <- tibble(
+  database = "ccae",
+  outcome = c(
+    "Acute myocardial infarction",
+    rep("", 3),
+    "Heart failure (hosp)",
+    rep("", 3),
+    "Stroke (ischemic or hemorrhagic)",
+    rep("", 3)
+  ),
+  riskQuarter = rep(1:4, 3),
+  patientsAce = c(
+    161099, 204882, 214413, 204167,
+    146259, 188006, 218052, 230226,
+    146069, 187524, 217070, 226128
+  ),
+  personYearsAce = c(
+    276171, 372197, 393583, 351727,
+    249809, 341014, 399394, 400330,
+    294484, 340234, 397830, 393861
+  ),
+  eventsAce = c(
+    203, 534, 117, 2095,
+    228, 457, 826, 2012,
+    299, 554, 947, 1718
+  ),
+  patientsBeta = c(
+    133977, 90193, 80662, 90908,
+    126387, 84280, 83421, 98380,
+    126264, 84000, 83038, 97628
+  ),
+  personYearsBeta = c(
+    220633, 169231, 150035, 154419,
+    206706, 158425, 155222, 169139,
+    206453, 157913, 154587, 167810
+  ),
+  eventsBeta = c(
+    135, 321, 535, 1520,
+    378, 340, 570, 1773,
+    320, 351, 521, 1077
+  )
+)
+
+tableMdcr <- tibble(
+  database = "mdcr",
+  outcome = c(
+    "Acute myocardial infarction",
+    rep("", 3),
+    "Heart failure (hosp)",
+    rep("", 3),
+    "Stroke (ischemic or hemorrhagic)",
+    rep("", 3)
+  ),
+  riskQuarter = rep(1:4, 3),
+  patientsAce = c(
+    27853,
+    27596,
+    25893,
+    20319,
+    27530,
+    27486,
+    25482,
+    19704,
+    27291,
+    27054,
+    24763,
+    18666
+  ),
+  personYearsAce = c(
+    57541,
+    56910,
+    53202,
+    38710,
+    56886,
+    56644,
+    52475,
+    37842,
+    56413,
+    55846,
+    50976,
+    35775
+  ),
+  eventsAce = c(
+    231,
+    387,
+    560,
+    828,
+    364,
+    582,
+    965,
+    1578,
+    375,
+    490,
+    752,
+    979
[…]

### 2021-09-07 submission/manuscript.rmd
align results section to methods for prediction step

@@
     developing the outcomes of interest. For example, the set of patients in a
-    database that have at least one occurrence of acute myocardial infarction in
-    their record.
+    database that have at least one occurrence of acute myocardial infarction
+    (MI) in their record.
 
@@
   - Outcome cohorts: We consider 3 main and 6 safety outcome cohorts. These are
-    patients in the database with a diagnosis of: acute myocardial infarction
-    (MI); hospitalization with heart failure; ischemic or hemorrhagic stroke
-    (efficacy outcomes); hypokalemia; hyperkalemia; hypotension; angioedema;
-    cough; abnormal weight gain (safety outcomes).
+    patients in the database with a diagnosis of: acute MI; hospitalization with
+    heart failure; ischemic or hemorrhagic stroke (efficacy outcomes);
+    hypokalemia; hyperkalemia; hypotension; angioedema; cough; abnormal weight
+    gain (safety outcomes).
 
@@
 
-## Step 3: Prediction
+Our analyses included a total of 784,561, 66,820 and 101,661 patients initiating
+treatment with ACE inhibitors and 395,740, 45,999 and 69,798 patients initiating
+treatment with beta blockers in CCAE, MDCD and MDCD respectively (Table
+\@ref(tab:table)). Adequate numbers of patients were included in all strata of
+predicted acute MI risk.
 
-We developed separate prediction models for all efficacy outcomes (acute MI,
-hospitalization with heart failure and hemorrhagic or ischemic stroke) in each
-database.
-
-\> > MOVE: ps_matching > >
-
-***More specifically, we first estimate propensity scores using LASSO
-logistic regression and a large set of baseline covariates including
-demographics, drug exposures, diagnoses, measurements and medical devices. We
-match patients 1-1 using a caliper, i.e. the maximum distance that is acceptable
-for any match. The default value we use is 0.2 on the standardized logit scale
-for the propensity scores.***
-
-< < MOVE < <
-
-The prediction models were estimated on the propensity score matched (1:1) subset of
-the population, using caliper of 0.2 and after excluding patients having the
-outcomes any time prior to treatment initiation. We chose a time horizon of 2
-years after inclusion into the target cohort. For this demonstration, we
-developed the prediction models using LASSO logistic regression with 3-fold
-cross validation for hyper-parameter selection. We chose a time horizon of 2
-years after inclusion into the target cohort. For this demonstration, we
-developed the prediction models using LASSO logistic regression with 3-fold
-cross validation for hyper-parameter selection.
-
-## Step 4: Estimation
-
-We used patient-level predictions to stratify the patient population into 4 risk
-quarters. Within risk strata, relative effects were estimated using Cox
-regression and absolute effects were estimated from the Kaplan-Meier estimate
-differences at 2 years after treatment initiation. To adjust for observed
-confounding within risk strata, we estimated propensity scores using the same
-approach as in the development of prediction models. We used the estimated
-propensity scores to stratify patients into 5 strata, within each risk quarter.
-
-## Step 5: Result presentation and evaluation
-
-We present the results of the analysis in the CCAE database with
-stratification based on risk predictions of acute MI. Results of analyses in the
-other databases and with other risk stratifications are included in the
-supplementary material.
-
-For each outcome and in each risk stratum there were adequate numbers of
-patients (Table \@ref(tab:table)). The discriminative ability of the prediction
-models was moderate in the matched development subset (c-index 0.76 for acute MI
-; 0.79 for hospitalization with heart failure; 0.74 for stroke; ), in the
-general population (c-index 0.74 for acute MI; 0.77 for hospitalization with
-heart failure; 0.73 for stroke), in the treatment cohort (c-index for acute MI
-it was 0.71, for hospitalization with heart failure was 0.76 and for stroke it
-was 0.72) and in the comparator cohort (c-index for acute MI it was 0.79 for
-hospitalization with heart failure was 0.79 and for stroke it was 0.75).
 \captionsetup{width=\textwidth}
@@
   ),
-  caption = "Number of patients, person years and events within quarters of predicted risk for hospitalization with heart failure for the 3 main outcomes of the study (acute myocardial infarction, hospitalization with heart failure and ischemic or hemorrhagic stroke) across the 3 databases."
+  caption = "Number of patients, person years and events within quarters of predicted risk for acute MI for the 3 main outcomes of the study (acute MI, hospitalization with heart failure and ischemic or hemorrhagic stroke) across the 3 databases.
+The populations used for the evaluation of hosptialization with heart failure and stroke are subsets of the populations used for the evaluation of actute MI."
 ) %>%
@@
 
+## Step 3: Prediction
+
+We internally developed separate prediction models for acute MI in all 3
+databases.
+
+\> > MOVE: ps_matching > >
+
+***More specifically, we first estimate propensity scores using LASSO
+logistic regression and a large set of baseline covariates including
+demographics, drug exposures, diagnoses, measurements and medical devices. We
+match patients 1-1 using a caliper, i.e. the maximum distance that is acceptable
+for any match. The default value we use is 0.2 on the standardized logit scale
+for the propensity scores.***
+
+< < MOVE < <
+
+The prediction models were estimated on the propensity score matched (1:1)
+subset of the population, using caliper of 0.2 and after excluding patients
+having the outcomes any time prior to treatment initiation. We chose a time
+horizon of 2 years after inclusion into the target cohort. For this
+demonstration, we developed the prediction models using LASSO logistic
+regression with 3-fold cross validation for hyper-parameter selection. For this
+demonstration, we developed the prediction models using LASSO logistic
+regression with 3-fold cross validation for hyper-parameter selection.
+
+The models developed in the 3 databases had moderate discriminative performance
+(internally validated) with no major issues of overfitting to any cohort except
+for the case of CCAE database in which the derived prediction model performed
+better in the comparator cohort (Table \@ref(tab:tableAucs)). We also observed
+lower performance of the prediction model developed in MDCR compared to the
+other 2 databases. Results on the calibration of the prediction models can be
+found in the supplement.
+
+\captionsetup{width=\textwidth}
 ```{r tableAucs, echo=FALSE, results='asis', booktabs=TRUE}
@@
 ```
+
+## Step 4: Estimation
+
+We used patient-level predictions to stratify the patient population into 4 risk
+quarters. Within risk strata, relative effects were estimated using Cox
+regression and absolute effects were estimated from the Kaplan-Meier estimate
+differences at 2 years after treatment initiation. To adjust for observed
+confounding within risk strata, we estimated propensity scores using the same
+approach as in the development of prediction models. We used the estimated
+propensity scores to stratify patients into 5 strata, within each risk quarter.
+
+## Step 5: Result presentation and evaluation
+
+We present the results of the analysis in the CCAE database with
+stratification based on risk predictions of acute MI. Results of analyses in the
+other databases and with other risk stratifications are included in the
+supplementary material.
+
+For each outcome and in each risk stratum there were adequate numbers of
+patients (Table \@ref(tab:table)). The discriminative ability of the prediction
+models was moderate in the matched development subset (c-index 0.76 for acute MI
+; 0.79 for hospitalization with heart failure; 0.74 for stroke; ), in the
+general population (c-index 0.74 for acute MI; 0.77 for hospitalization with
+heart failure; 0.73 for stroke), in the treatment cohort (c-index for acute MI
+it was 0.71, for hospitalization with heart failure was 0.76 and for stroke it
+was 0.72) and in the comparator cohort (c-index for acute MI it was 0.79 for
+hospitalization with heart failure was 0.79 and for stroke it was 0.75).
 

### 2021-09-08 submission/manuscript.rmd
add evaluation plots to estimation step

@@
 propensity scores to stratify patients into 5 strata, within each risk quarter.
+```{r figure8, echo=FALSE, fig.cap="Absolute risk reduction for the main and safety outcomes, estimated by fitting stratified Cox regression models within quarters of predicted risk of acute myocardial infarction (MI).
+The four risk quarters (Q1-Q4) are defined using the internally developed model for acute MI", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/PsDensity.tiff")))
+```
+
+```{r figure9, echo=FALSE, fig.cap="Absolute risk reduction for the main and safety outcomes, estimated by fitting stratified Cox regression models within quarters of predicted risk of acute myocardial infarction (MI).
+The four risk quarters (Q1-Q4) are defined using the internally developed model for acute MI", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/CovariateBalance.tiff")))
+```
 

### 2021-09-08 code/CovariateBalancePlot.R

@@ -1,0 +1,54 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+
+databases <- c("ccae", "mdcd", "mdcr")
+pp <- list()
+
+for (i in seq_along(databases)) {
+  pattern <- paste0("^bal.*", databases[i], ".*104_104*.rds")
+  pp[[i]] <- list.files(
+    path = ".scratch/AceBeta9Outcomes/data",
+    pattern = pattern,
+    full.names = TRUE
+  ) %>%
+    map(readRDS) %>%
+    bind_rows() %>%
+    tibble() %>%	
+    CohortMethod::plotCovariateBalanceScatterPlot(
+      beforeLabel = "",
+      afterLabel = toupper(databases[i])
+    ) +
+    ggplot2::facet_grid(
+      ~riskStratum
+    ) +
+    theme_bw() +
+    theme(
+      plot.title   = element_blank(),
+      axis.title.x = element_blank(),
+      axis.text.x  = element_text(size = 22),
+      axis.text.y  = element_text(size = 22),
+      axis.title   = element_text(size = 30),
+      strip.text   = element_text(size = 25),
+      strip.background = element_rect(colour = "black", fill = NA)
+    )
+  
+  if (i != 1) {
+    pp[[i]] <- pp[[i]] +
+      theme(
+        strip.text       = element_blank(),
+        strip.background = element_blank()
+      )
+  }
+}
+
+plot <- gridExtra::grid.arrange(pp[[1]], pp[[2]], pp[[3]], nrow = 3)
+ggsave(
+  "figures/CovariateBalance.tiff",
+  plot, 
+  compression = "lzw", 
+  width       = 600, 
+  height      = 350,
+  units       = "mm",
+  dpi         = 300
+)

### 2021-09-08 code/PsDensityPlot.R

@@ -1,0 +1,81 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+
+databases <- c("ccae", "mdcd", "mdcr")
+pp <- list()
+
+for (i in seq_along(databases)) {
+  pattern <- paste0("^ps.*", databases[i], ".*104_104.*.rds")
+  pp[[i]] <- list.files(
+    path = ".scratch/AceBeta9Outcomes/data",
+    pattern = pattern,
+    full.names = TRUE
+  ) %>%
+    map(readRDS) %>%
+    bind_rows() %>%
+    mutate(
+      treatment = as.factor(treatment)
+    ) %>%
+    tibble() %>%	
+    ggplot2::ggplot(
+      ggplot2::aes(
+        x = x,
+        y = y
+      )
+    ) +
+    ggplot2::geom_density(
+      stat = "identity",
+      ggplot2::aes(
+        color = treatment,
+        group = treatment,
+        fill = treatment
+      )
+    ) +
+    scale_x_continuous(breaks = seq(0, 1, .5)) +
+    ggplot2::facet_grid(~riskStratum) +
+    ggplot2::ylab(
+      label = toupper(databases[i])
+    ) +
+    ggplot2::xlab(
+      label = "Preference score"
+    ) +
+    ggplot2::scale_fill_manual(
+      values = alpha(c("#fc8d59", "#91bfdb"), .6)
+    ) +
+    ggplot2::scale_color_manual(
+      values = alpha(c("#fc8d59", "#91bfdb"), .9)
+    ) +
+    theme_classic() +
+    ggplot2::theme(
+      legend.title    = ggplot2::element_blank(),
+      legend.position = "none",
+      axis.title.x    = element_blank(),
+      axis.line.y     = element_blank(),
+      axis.ticks.y    = element_blank(),
+      axis.text.y     = element_blank(),
+      axis.text.x     = element_text(size = 22),
+      axis.title      = element_text(size = 30),
+      strip.text      = element_text(size = 25)
+    )  
+  if (i != 1) {
+    pp[[i]] <- pp[[i]] +
+      theme(
+        strip.text       = element_blank(),
+        strip.background = element_blank()
+      )
+  }
+}
+
+
+plot <- gridExtra::grid.arrange(pp[[1]], pp[[2]], pp[[3]], nrow = 3)
+ggsave(
+  "figures/PsDensity.tiff",
+  plot, 
+  compression = "lzw", 
[…]

### 2021-09-09 submission/manuscript.rmd
modify results section to demonstrate power of OMOP

--- 2021-09-07 submission/manuscript.rmd
@@
 \captionsetup{width=\textwidth}
-```{r table, echo=FALSE, warning=FALSE, message=FALSE, results='asis', booktabs=TRUE}
+```{r table, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, results='asis', booktabs=TRUE}
 table <- readr::read_csv(
@@
 \captionsetup{width=\textwidth}
-```{r tableAucs, echo=FALSE, results='asis', booktabs=TRUE}
+```{r tableAucs, cache=TRUE, echo=FALSE, results='asis', booktabs=TRUE}
 table <- tibble(
@@
 
-## Step 5: Result presentation and evaluation
+In general, there was sufficient overlap of propensity score distribution in all
+risk strata (Figure \@ref(fig:psDensity)). If empirical equipoise was not
+achieved, the validity of the comparative effectiveness estimates would be
+questionable.
 
-We present the results of the analysis in the CCAE database with
-stratification based on risk predictions of acute MI. Results of analyses in the
-other databases and with other risk stratifications are included in the
-supplementary material.
-
-For each outcome and in each risk stratum there were adequate numbers of
-patients (Table \@ref(tab:table)). The discriminative ability of the prediction
-models was moderate in the matched development subset (c-index 0.76 for acute MI
-; 0.79 for hospitalization with heart failure; 0.74 for stroke; ), in the
-general population (c-index 0.74 for acute MI; 0.77 for hospitalization with
-heart failure; 0.73 for stroke), in the treatment cohort (c-index for acute MI
-it was 0.71, for hospitalization with heart failure was 0.76 and for stroke it
-was 0.72) and in the comparator cohort (c-index for acute MI it was 0.79 for
-hospitalization with heart failure was 0.79 and for stroke it was 0.75).
-
-Relative treatment effects of ACE-inhibitors vs beta blockers increased in favor
-of ACE-inhibitors (hazard ratios decreased) with increasing acute MI risk,
-resulting in more pronounced absolute risk differences (ARD) with increasing
-acute MI risk (Figure \@ref(fig:figure2)).
-
-```{r figure2, echo=FALSE, fig.cap="Overview of heterogeneity of ACE-inhibitors treatment within strata of predicted risk of acute MI.
-The top panel contains the observed acute rates of ACE-inhibitors and beta blockers within each quarter of predicted acute MI risk.
-These are derived using the KaplanMeier estimates at 730 days after inclusion.
-The middle panel, contains the hazard ratios of comparing ACE-inhibitors to beta blockers with regard to acute MI.
-These are estimated using Cox proportional hazards regression within quarters of predicted acute MI risk.
-The bottom panel contains absolute risk reduction for ACE-inhibitors compared to beta blockers.
-These are derived as the difference in Kaplan-Meier estimates at 730 after inclusion.
-Hazard ratios in the middle panel show a decreasing trend with increasing acute MI risk.
-Given the rather good discrimination of the prediction model (AUC=0.74), this results in an increasing trend for absolute benefit in favor of ACE-inhibitors with increasing risk.", out.width = '100%'}
-grid::grid.raster(tiff::readTIFF(here("figures/Figure_2.tif")))
+```{r psDensity, cache = TRUE, echo=FALSE, fig.cap="Preference score distributions for the evaluation of heterogeneity of the effect of ACE inhibitors compared to beta blockers on acute MI based on quarters of predicted acute MI risk.
+The preference score is a transformation of the propensity score that adjusts for prevalence differences between populations.", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/PsDensity.tiff")))
 ```
 
-In general, with increasing acute MI risk we observed an increasing pattern in
-terms of absolute benefit for the main outcomes, while the absolute harms did
-not increase for the safety outcomes. More specifically, patients in the low
-risk quarter did not receive absolute benefit (-0.03%) while absolute risk
-was 0.54% lower (95% confidence interval 0.36%—0.71%) for patients in the
-highest risk quarter. In contrast, the absolute and relative effects of
-ACE-inhibitors on safety outcomes (e.g. cough and angioedema) are slightly
-decreasing with increasing acute MI risk (Figure \@ref(fig:figure3) and
-\@ref(fig:figure4)). Similar results were observed in the other two databases
-(see supplementary material).
+Propensity score adjustment achieved balance for most of the considered
+covariates, measured using standardized mean differences before and after
+adjustment (Figure \@ref(fig:covariateBalance)). However, in lower risk strata
+imbalances persisted for a substantial subset of the covariates, all related to
+pregnancy findings. This was anticipated (use of ACE inhibitors is specifically
+contraindicated during pregnancy) and was already pointed out in
+[@Suchard2019].
 
-```{r figure3, echo=FALSE, fig.cap="Hazard ratios (relative treatment effects) for the main and safety outcomes, estimated by fitting stratified Cox regression models within quarters of predicted risk of acute myocardial infarction (MI).
-The four risk quarters (Q1-Q4) are defined using the internally developed model for acute MI", out.width = '100%'}
-grid::grid.raster(tiff::readTIFF(here("figures/Figure_3.tif")))
+```{r covariateBalance, cache = TRUE, echo=FALSE, fig.cap="Patient charactersistic balance for ACE inhibitors and beta blockers before and after stratification on the propensity scores.
+Each dot represents the standardized difference of means for a single covariate before (x-axis) and after (y-axis) stratification", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/CovariateBalance.tiff")))
 ```
 
-```{r figure4, echo=FALSE, fig.cap="Absolute risk reduction for the main and safety outcomes, estimated by fitting stratified Cox regression models within quarters of predicted risk of acute myocardial infarction (MI).
-The four risk quarters (Q1-Q4) are defined using the internally developed model for acute MI", out.width = '100%'}
-grid::grid.raster(tiff::readTIFF(here("figures/Figure_4.tif")))
+## Step 5: Presentation of results
+
+For hospitalization with acute MI there was an increasing trend in favor ACE
+inhibitors compared to beta blockers on the relative scale (hazard ratios
+decreased) with increasing acute MI risk. More specifically, hazard ratios
+decreased from 0.98 (0.77 to 1.26; 95% CI), 1.30 (0.51 to 3.22; 95% CI) and 1.03
+(0.82 to 1.29; 95% CI) to 0.76 (0.71 to 0.82; 95% CI), 0.94 (0.82 to 1.07; 95%
+CI) and 1.03 (0.93 to 1.15; 95% CI) in CCAE, MDCD and MDCR respectively (Figure
+\@ref(fig:combinedRelative)). In terms of hospitalization with heart failure
+relative treatment effect estimates favored ACE inhibitors across all risk
+strata in all databases. We found no differences between the two treatments in
+their effect on stroke on the relative scale. In terms of the safety outcomes we
+found an increased risk of cough and angioedema on the relative scale across all
+risk strata. In the case of cough, this effect decreased with increasing risk of
+acute MI---from 1.37 (1.33 to 1.41; 95% CI), 1.35 (1.24 to 1.48; 95% CI)
+and 1.37 (1.29 to 1.45; 95% CI) in the lowest acute MI risk quarter to 1.26
+(1.22 to 1.29; 95% CI), 1.07 (1.00 to 1.14) and 1.10 (1.04 to 1.17; 95% CI) in
+the highest acute MI risk quarter in CCAE, MDCD and MDCR respectively.
+
+```{r combinedRelative, cache=TRUE, echo=FALSE, fig.cap="Overview of heterogeneity of ACE-inhibitors treatment on the relative scale (hazard ratios) within strata of predicted risk of acute MI.
+Values below 1 favor ACE inhibitors, while values above 1 favor beta blockers.", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/CombinedRelative.tiff")))
+```
+
+We observed an increasing trend of treatment effect on the absolute scale with
+increasing acute MI risk in favor of ACE inhibitors in terms of acute MI in all
+databases except for MDCR---from -0.03% (-0.08% to 0.01%; 95% CI), -0.05%
+(-0.18% to 0.08%; 95% CI) and -0.02% (-0.24% to 0.19%; 95% CI) in the lowest
+acute MI risk quarter to 0.54% (0.36% to 0.71%; 95% CI), 0.29% (-0.39%
+to 0.97%; 95% CI) and -0.39% (-0.96% to 0.18%; 95% CI) in the highest acute MI
+risk quarter in CCAE, MDCD and MDCR, respectively (Figure \@ref(fig:figure12)).
+We found no difference on the absolute scale for stroke across risk strata.
+Absolute risk differences did not favor ACE inhibitors compared to beta blockers
+in terms of cough, even though this effect again diminished with increasing
+acute MI risk---from -4.14% (-4.62% to -3.66%; 95% CI), -6.45% (-9.12% to
+-3.78%; 95% CI) and -4.81% (-5.76% to -3.85%; 95% CI) in the lowest acute MI
+risk quarter to -2.57% (-2.99% to -2.15%; 95% CI), -1.11% (-2.93% to 0.70%; 95%
+CI) and -1.69% (-2.83% to -0.55%; 95% CI) in the highest acute MI risk quarter
+in CCAE, MDCD and MDCR, respectively. In terms of angioedema absolute risk
+differences were very small due to the rarity of the outcome.
+
+```{r figure12, cache=TRUE, echo=FALSE, fig.cap="Overview of heterogeneity of ACE-inhibitors treatment on the absolute scale within strata of predicted risk of acute MI.
+Estimates of absolute treatment effect are derived as the difference in Kaplan-Meier estimates at 730 after inclusion.
+Values above 0 favor ACE inhibitors, while values below 0 favor beta blockers.", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/CombinedAbsolute.tiff")))
 ```
@@
 first-line treatment for younger hypertensive patients [@Cruickshank2007;
-@Khan2006]. More thorough evaluation of these results is required in future
-research, however.
+@Khan2006]. This analysis, however, was carried out as demonstration of the
+framework and more rigorous analyses are required to make any suggestions for
+clinical practice.
 

### 2021-09-09 code/CombinedPlots.R

@@ -53,6 +53,9 @@ prepareDataset <- function(data) {
 relative <- prepareDataset(mappedOverallRelativeResults) %>%
   filter(
     stratOutcome == "Acute myocardial infarction"
+  ) %>%
+  mutate(
+    database = toupper(database)
   )
 
 relativePlot <- ggplot(
@@ -66,10 +69,12 @@ relativePlot <- ggplot(
   )
 ) +
   geom_point(
-    position = position_dodge(.5)
+    position = position_dodge(.5),
+    size     = 3.5
   ) +
   geom_errorbar(
     position = position_dodge(.5),
+    size = 1.1,
     width = 0
   ) +
   facet_grid(~riskStratum) +
@@ -85,9 +90,9 @@ relativePlot <- ggplot(
       "#8da0cb"
     ),
     breaks = c(
-      "mdcr",
-      "mdcd",
-      "ccae"
+      "MDCR",
+      "MDCD",
+      "CCAE"
     )
   ) +
   theme_bw() +
@@ -95,13 +100,14 @@ relativePlot <- ggplot(
     legend.position = "top",
     legend.title = element_blank(),
     axis.title.y = element_blank(),
-    axis.title.x = element_text(size = 20),
-    axis.text    = element_text(size = 15),
-    legend.text  = element_text(size = 15), 
+    axis.text.x  = element_text(size = 25),
+    axis.text.y  = element_text(size = 25),
+    axis.title   = element_text(size = 35),
+    legend.text  = element_text(size = 30), 
+    strip.text   = element_text(size = 25),
     strip.background = element_rect(
       fill = "#ffffff"
     ),
-    strip.text = element_text(size = 15)
   )
 
 ggsave(
@@ -109,7 +115,7 @@ ggsave(
   relativePlot, 
   compression = "lzw", 
   width       = 600, 
-  height      = 300,
+  height      = 350,
   units       = "mm",
   dpi         = 300
 )
@@ -120,6 +126,9 @@ absolute <- prepareDataset(mappedOverallAbsoluteResults) %>%
   filter(
     stratOutcome == "Acute myocardial infarction"
     # estOutcome != "Cough"
+  ) %>%
+  mutate(
+    database = toupper(database)
   )
 
 absolutePlot <- ggplot(
@@ -133,10 +142,12 @@ absolutePlot <- ggplot(
   )
 ) +
   geom_point(
-    position = position_dodge(.5)
+    position = position_dodge(.5),
+    size = 3.5
[…]

### 2021-09-12 submission/manuscript.rmd
include resutls from all databases; closes #1
first draft of response

@@
 
-We aimed to develop a framework for risk-based assessment of treatment effect heterogeneity
-in high-dimensional observational data, which extends the existing guidelines of
-the RCT setting.
+We aimed to develop a framework for risk-based assessment of treatment effect
+heterogeneity in high-dimensional observational data, which extends the existing
+guidelines of the RCT setting.
 
@@
 relative and the absolute scale) within risk strata---typically 4 risk
-quarters---defined using the prediction model of step 3. Any appropriate method
-for the evaluation of relative and absolute treatment effects can be considered,
-as long as the this is done consistently in all risk strata. Common approaches
-are odds ratios or hazard ratios for relative scale estimates and differences in
-observed proportions or differences in Kaplan-Meier estimates for absolute scale
-estimates, depending on the problem at hand. We estimate propensity scores
-within risk strata which we then use to match patients from different treatment
-cohorts or stratify them into groups with similar propensity scores or to weigh
-each patient's contribution to the estimation process [@Austin2011].
+quarters---defined using the prediction model of step 3.
+
+\> > ADD > >
+
+***Effect estimation may be focused on the difference in outcomes for a randomly
+selected person from the risk stratum (average treatment effect) or for a
+randomly selected person from the treatment cohort within the risk stratum
+receiving the treatment under study (average treatment effect on the
+treated).***
+
+< < ADD < <
+
+Any appropriate method for the evaluation of relative and absolute treatment
+effects can be considered, as long as the this is done consistently in all risk
+strata. Common approaches are odds ratios or hazard ratios for relative scale
+estimates and differences in observed proportions or differences in Kaplan-Meier
+estimates for absolute scale estimates, depending on the problem at hand. We
+estimate propensity scores within risk strata which we then use to match
+patients from different treatment cohorts or stratify them into groups with
+similar propensity scores or to weigh each patient's contribution to the
+estimation process [@Austin2011].
 
@@
 
+\> > ADD > >
+
+***Our aim was to estimate the average treatment effects on the relative and the
+absolute scale within strata of predicted acute MI risk.***
+
+< < ADD < <
+
 We used patient-level predictions to stratify the patient population into 4 risk
@@
 
-```{r covariateBalance, cache = TRUE, echo=FALSE, fig.cap="Patient charactersistic balance for ACE inhibitors and beta blockers before and after stratification on the propensity scores.
+```{r covariateBalance, cache = TRUE, echo=FALSE, fig.cap="Patient characteristic balance for ACE inhibitors and beta blockers before and after stratification on the propensity scores.
 Each dot represents the standardized difference of means for a single covariate before (x-axis) and after (y-axis) stratification", out.width = '100%'}

### 2021-09-13 submission/manuscript.rmd
more changes for response

--- 2021-09-09 submission/manuscript.rmd
@@
 
-```{r, echo=FALSE, message=FALSE}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 library(tidyverse)
@@
 observational databases. With increasing risk of acute myocardial infarction we
-observed increasing absolute benefits, i.e. from -0.03% to 0.54% in the lowest
+observed increasing absolute benefits, i.e. from -0.05% to -0.02% in the lowest risk quarter to 
 to highest risk groups. Cough-related absolute harms decreased from 4.1%
@@
 
-\> > EDIT > >
-
-***In recent years, a large number of methods has been developed for the
+In recent years, a large number of methods has been developed for the
 assessment of HTE, mainly in the RCT setting. Earlier work suggested separating
@@
 relative and the absolute scale. Recently, systematic guidance on the
-application of such methods has been developed [@Kent2010; @Kent2018].***
-
-< < EDIT < <
-
-\> > OLD > >
-
-Predictive HTE analyses have been described as approaches that
-provide predictions of potential outcomes in a particular patient with one
-intervention versus an alternative, taking into account multiple relevant
-patient characteristics [@Varadhan2013]. We focus on “risk modeling”
-approaches where patients are divided into risk strata using either an existing
-or an internally developed risk prediction model. Risk-stratum-specific
-estimates provide an overview of the evolution of treatment effects with
-increasing risk both on the relative and the absolute scale. Recently,
-systematic guidance on the application of such methods has been developed
-[@Kent2010; @Kent2018].
-
-< < OLD < <
+application of such methods has been developed [@Kent2010; @Kent2018].
 
@@
 different databases calls for database-specific analysis plans that often are
-not easily transportable.
-
-\> > ADD > >
-
-***Because of the latter issue, running analyses at scale demands a big investment
-of time and effort, as researchers are forced to map their analysis plans to the
-databases available to them.***
-
-< < ADD < <
+not easily transportable. Because of the latter issue, running analyses at scale
+demands a big investment of time and effort, as researchers are forced to map
+their analysis plans to the databases available to them.
 
@@
 
-\> > EDIT > >
-
-***We aimed to develop a framework for implementing risk-based assessment of
-treatment heterogeneity in high-dimensional observational data, extending the
-existing guidelines of the RCT setting. Our publicly available package provides
-an out-of-the-box solution for implementing such analyses at scale within the
-OHDSI network, taking advantage of the OMOP-CDM.***
-
-< < EDIT < <
-
-\> > OLD > >
-
-We aimed to develop a framework for risk-based assessment of treatment effect heterogeneity
-in high-dimensional observational data, which extends the existing guidelines of
-the RCT setting.
-
-< < OLD < <
-
-We implemented the framework using existing OHDSI methods including the
-patient-level prediction framework and the population-level effect estimation
-framework based on new-user cohort design [@Reps2018; @Ryan2013]. As a
-proof-of-concept we analyzed heterogeneity of the effects of first-line
-hypertension treatment: we compared the effect of angiotensin converting enzyme
-(ACE) inhibitors to beta blockers on 9 outcomes across three different US claims
-databases.
+We aimed to develop a framework for implementing a risk-based predictive
+approach for evaluating HTE in high-dimensional observational data, extending
+the existing guidelines of the RCT setting. Our publicly available package
+provides an out-of-the-box solution for implementing such analyses at scale
+within the OHDSI network, taking advantage of the OMOP-CDM. We implemented the
+framework using existing OHDSI methods including the patient-level prediction
+framework and the population-level effect estimation framework based on new-user
+cohort design [@Reps2018; @Ryan2013]. As a proof-of-concept we analyzed
+heterogeneity of the effects of first-line hypertension treatment: we compared
+the effect of angiotensin converting enzyme (ACE) inhibitors to beta blockers
+on 9 outcomes across three different US claims databases.
 
@@
 absolute and relative treatment effects are estimated within risk strata; 5)
-presentation and evaluation of the results. We developed an open-source
-R-package for the implementation of the proposed framework and made it publicly
-available. The source code can be found at
-<https://github.com/OHDSI/RiskStratifiedEstimation>. An overview of the
-procedure can be seen in Figure \@ref(fig:figure1).
-
-```{r figure1, echo=FALSE, fig.cap="(A) Starting from a treatment (top), a comparator (bottom) and an outcome (middle) cohort we estimate the propensity scores on the entire target population.
-(B) We match patients on the propensity scores and estimate the prediction model.
-Since we match patients we develop the prediction model on smaller subset of the initial population and, therefore, the number of patients is smaller in B compared to A.
-(C) We apply the prediction model on the entire population (green: lower 25\\% of the risk distribution; yellow: patients with risk between 25\\% and 50\\% of the risk distribution; orange: patients with risk between 50\\% and 75\\% of the risk distribution; red: patients at risk higher than 75\\% of the risk distribution).
-(D) We separate in risk subgroups, here quarters.
-Within risk quarters propensity scores are estimated again and relative and absolute treatment effects are estimated.", out.width = '100%'}
-# knitr::include_graphics(here("figures/Figure_1.tif"))
-grid::grid.raster(tiff::readTIFF(here("figures/Figure_1.tif")))
-```
+presentation of the results. We developed an open-source R-package for the
+implementation of the proposed framework and made it publicly available. The
+source code can be found at <https://github.com/OHDSI/RiskStratifiedEstimation>.
 
@@
 
-\> > ADD > >
-
-***Cohort definitions are crucial for this step of the framework. We define a
+Cohort definitions are crucial for this step of the framework. We define a
 cohort as the set of patients who satisfy one or more inclusion criteria for a
@@
 transportable, meaning that in theory they can be implemented in any database,
-provided that it is mapped to the OMOP-CDM.***
+provided that it is mapped to the OMOP-CDM.
 
-< < ADD < <
+Our framework uses a comparative cohort design. This means that at least 3
+cohorts of patients need to be defined at this stage of the framework.
 
@@
 generalizability of results. Furthermore, the cohorts should preferably have
-adequate sample size to ensure precise effect estimation, even within smaller
-risk strata.
+adequate sample size with adequate follow-up time to ensure precise effect
+estimation, even within smaller risk strata. Other issues that may be of
+importance for database inclusion are the depth of data capture (the precision
+at which measurements, lab tests, conditions are recorded), the reliability of
+data entry and many more, also depending on the task at hand.
 
@@
 
-< < MOVE: ps_matching > >
-
-\> > ADD > >
-
-***Other methods of fitting the propensity scores, such as random forest and
-others (REF) can also be considered.***
-
-< < ADD < <
+More specifically, we first estimate propensity scores using LASSO logistic
+regression and a large set of baseline covariates including demographics, drug
+exposures, diagnoses, measurements and medical devices. We match patients 1-1
+using a caliper, i.e. the maximum distance that is acceptable for any match. The
+default value we use is 0.2 on the standardized logit scale for the propensity
+scores. Other methods of fitting the propensity scores, such as random forest
+and others can also be considered.
 
@@
 relative and the absolute scale) within risk strata---typically 4 risk
-quarters---defined using the prediction model of step 3. Any appropriate method
-for the evaluation of relative and absolute treatment effects can be considered,
-as long as the this is done consistently in all risk strata. Common approaches
-are odds ratios or hazard ratios for relative scale estimates and differences in
-observed proportions or differences in Kaplan-Meier estimates for absolute scale
-estimates, depending on the problem at hand. We estimate propensity scores
-within risk strata which we then use to match patients from different treatment
-cohorts or stratify them into groups with similar propensity scores or to weigh
-each patient's contribution to the estimation process [@Austin2011].
+quarters---defined using the prediction model of step 3. Effect estimation may
+be focused on the difference in outcomes for a randomly selected person from the
+risk stratum (average treatment effect) or for a randomly selected person from
+the treatment cohort within the risk stratum receiving the treatment under study
+(average treatment effect on the treated).
+
+Any appropriate method for the evaluation of relative and absolute treatment
+effects can be considered, as long as the this is done consistently in all risk
+strata. Common approaches are odds ratios or hazard ratios for relative scale
+estimates and differences in observed proportions or differences in Kaplan-Meier
+estimates for absolute scale estimates, depending on the problem at hand. We
+estimate propensity scores within risk strata which we then use to match
+patients from different treatment cohorts or stratify them into groups with
+similar propensity scores or to weigh each patient's contribution to the
+estimation process [@Austin2011].
 
@@
 differences before and after propensity score adjustment.
+
+A schematic overview of the prediction and estimation steps is presented in
+Figure \@ref(fig:figure1).
+
+```{r figure1, echo=FALSE, fig.cap="(A) Starting from a treatment (top), a comparator (bottom) and an outcome (middle) cohort we estimate the propensity scores on the entire target population.
+(B) We match patients on the propensity scores and estimate the prediction model.
+Since we match patients we develop the prediction model on smaller subset of the initial population and, therefore, the number of patients is smaller in B compared to A.
+(C) We apply the prediction model on the entire population (green: lower 25\\% of the risk distribution; yellow: patients with risk between 25\\% and 50\\% of the risk distribution; orange: patients with risk between 50\\% and 75\\% of the risk distribution; red: patients at risk higher than 75\\% of the risk distribution).
+(D) We separate in risk subgroups, here quarters.
+Within risk quarters propensity scores are estimated again and relative and absolute treatment effects are estimated.", out.width = '100%'}
+# knitr::include_graphics(here("figures/Figure_1.tif"))
+grid::grid.raster(tiff::readTIFF(here("figures/Figure_1.tif")))
+```
 
@@
 
-\> > MOVE: ps_matching > >
-
-***More specifically, we first estimate propensity scores using LASSO
-logistic regression and a large set of baseline covariates including
-demographics, drug exposures, diagnoses, measurements and medical devices. We
-match patients 1-1 using a caliper, i.e. the maximum distance that is acceptable
-for any match. The default value we use is 0.2 on the standardized logit scale
-for the propensity scores.***
-
-< < MOVE < <
-
 The prediction models were estimated on the propensity score matched (1:1)
@@
 
+Our aim was to estimate the average treatment effects on the relative and the
+absolute scale within strata of predicted acute MI risk.
+
 We used patient-level predictions to stratify the patient population into 4 risk
@@
 information from these data sources, allowing for large-scale risk-based
-assessment of treatment effect heterogeneity. Multiple outcomes can be evaluated
-in patient subgroups of similar baseline outcome risk. Multiple outcome risk
-stratification schemes can also be considered. However, this should be done with
-caution, as it may hinder the interpretability of the results, in a similar
-manner as typical subgroup analyses.
+assessment of treatment effect heterogeneity. It is an addition to the rapidly
+expanding literature of approaches for evaluating treatment effect
+heterogeneity. Multiple outcomes can be evaluated in patient subgroups of
+similar baseline outcome risk. Multiple outcome risk stratification schemes can
+also be considered. However, this should be done with caution, as it may hinder
+the interpretability of the results, in a similar manner as typical subgroup
+analyses.
 
@@
 
-\> > EDIT > >
-
-***Ideally, externally derived and adequately validated prediction model would
-be preferred for analyzing treatment effect heterogeneity [@Kent2010]. In the
+Ideally, externally derived and adequately validated prediction model would be
+preferred for analyzing treatment effect heterogeneity [@Kent2010]. In the
 absence of such prediction models an internally-developed risk prediction model
@@
 gave relatively unbiased estimates of treatment effect across the spectrum of
-risk [@Burke2014].***
-
-< < EDIT < <
-
-\> > OLD > >
-
-Externally derived and well transportable prediction models are preferred for
-analyzing treatment effect heterogeneity [@Kent2010]. In the absence of such
-prediction models, simulations of RCTs have shown that internal models can be
-used to provide unbiased estimates of treatment effect across the spectrum of
-baseline risk [@Burke2014].
-
-< < OLD < <
-
-However, in observational databases treatment arms
-may significantly differ in sample size. Because the prediction model will
-possibly better fit to the larger treatment arm, spurious treatment-covariate
+risk [@Burke2014]. However, in observational databases treatment arms may
+significantly differ in sample size. Because the prediction model will possibly
+better fit to the larger treatment arm, spurious treatment-covariate
 interactions may be introduced in the prediction model, leading to sub-optimal

### 2021-09-13 code/CombinedPlots.R

@@ -128,7 +128,10 @@ absolute <- prepareDataset(mappedOverallAbsoluteResults) %>%
     # estOutcome != "Cough"
   ) %>%
   mutate(
-    database = toupper(database)
+    database = toupper(database),
+    estimate = 100 * estimate,
+    lower    = 100 * lower,
+    upper    = 100 * upper
   )
 
 absolutePlot <- ggplot(

### 2021-09-13 submission/manuscript.rmd
update abstract

@@
 databases; 3) development of a prediction model for the outcome(s) of
-interest; 4) estimation of propensity scores within strata of predicted risk and
-estimation of relative and absolute treatment effect within strata of predicted
-risk; 5) evaluation and presentation of results. We demonstrate our framework by
-evaluating heterogeneity of the effect of angiotensin-converting enzyme (ACE)
-inhibitors versus beta blockers on a set of 9 outcomes of interest across three
-observational databases. With increasing risk of acute myocardial infarction we
-observed increasing absolute benefits, i.e. from -0.05% to -0.02% in the lowest risk quarter to 
-to highest risk groups. Cough-related absolute harms decreased from 4.1%
-to 2.6%. The proposed framework may be useful for the evaluation of
-heterogeneity of treatment effect on observational data that are mapped to the
-OMOP Common Data Model. The proof of concept study demonstrates its feasibility
+interest; 4) estimation of relative and absolute treatment effect within strata
+of predicted risk, after adjusting for observed confounding; 5) presentation of
+the results. We demonstrate our framework by evaluating heterogeneity of the
+effect of angiotensin-converting enzyme (ACE) inhibitors versus beta blockers on
+3 efficacy and 6 safety outcomes across three observational databases. Patients
+at low risk of acute myocardial infarction (MI) received negligible absolute
+benefits for all 3 efficacy outcomes, though they were more pronounced in the
+highest risk quarter, especially for hospitalization with heart failure. The
+substantial risk increase of cough and angioedema with ACE inhibitors across all
+risk strata suggests that beta blockers provide a viable alternative to patients
+at low risk of acute MI. The proof of concept study demonstrates its feasibility
 in large observational data. Further insights may arise by application to safety

### 2022-05-04 submission/manuscript.rmd
updating manuscript with new results

@@
 }
-knit_hooks$set(
-  inline = function(x) {
-    prettyNum(
-      x,
-      big.mark = ",",
-      decimal.mark = ".",
-      preserve.width = "individual"
-    )
+inline_hook <- function(x) {
+  if (is.numeric(x)) {
+    if (abs(x - round(x)) < .Machine$double.eps) {
+      formatted <- format(x, digits = 1, big.mark = ",")
+    } else {
+      formatted <- format(x, digits = 1 ,nsmall = 2, big.mark = ",")
     }
-)
+  } else {
+    formatted <- x
+  }
+
+  return(formatted)
+}
+
+knit_hooks$set(inline = inline_hook)
 ```
@@
 
-Understanding how a treatment’s effect varies across patients---a concept
-described as heterogeneity of treatment effects (HTE)---has been central to the
-agenda for both personalized (or precision) medicine and comparative
-effectiveness research. More formally, HTE has been defined as the non-random
-variability in the direction or magnitude of a treatment effect, in which the
-effect is measured using clinical outcomes [@KRAVITZ2004]. Usually, analyses
-focus on the relative scale, where treatment effects are assessed one at a time
-in patient subgroups defined from single covariates, an approach that suffers
-from low power and multiplicity issues [@Yusuf1991; @FernandezyGarcia2010].
-However, even with well-established constant relative effects, treatment benefit
-(or harm) may vary substantially on the absolute scale.
+The Observational Health Data Science and Informatics (OHDSI) collaborative has
+established a global network of data partners and researchers that aim to bring
+out the value of health data through large-scale analytics by mapping local
+databases to the Observational Medical Outcomes Partnership (OMOP) Common Data
+Model (CDM) [@hripcsak2015observational;@Overhage2012]. A standardized framework
+applying current best practices for comparative effectiveness studies within the
+OHDSI setting has been proposed [@Ryan2013]. This framework was implemented on a
+large scale for estimation of average effects of all first-line hypertension
+treatment classes on a total of 52 outcomes of interest across a global network
+of 9 observational databases [@Suchard2019].
 
-In recent years, a large number of methods has been developed for the
-assessment of HTE, mainly in the RCT setting. Earlier work suggested separating
-HTE analyses into exploratory, confirmatory, descriptive and predictive
-[@Varadhan2013]. Exploratory analyses focus on hypothesis generation,
-confirmatory analyses test subgroup effect hypotheses, descriptive analyses aim
-at facilitating future synthesis of subgroup effects and predictive analyses
-predict probabilities of benefit or harm in individual patients. Predictive HTE
-approaches can be further subdivided into risk modeling, treatment effect
-modeling and optimal treatment regime methods, based on the reference class used
-for defining patient similarity when making individualized predictions or
-recommendations [@Rekkas2020]. We focus on “risk modeling” approaches where
-patients are divided into risk strata using either an existing or an internally
-developed risk prediction model. Risk-stratum-specific estimates provide an
-overview of the evolution of treatment effects with increasing risk both on the
-relative and the absolute scale. Recently, systematic guidance on the
-application of such methods has been developed [@Kent2010; @Kent2018].
+Treatment effects can often vary substantially across individual patients,
+causing overall effect estimates to be inaccurate for a significant proportion
+of the patients at hand [@Rothwell1995;@KRAVITZ2004]. Understanding
+heterogeneity of treatment effects (HTE) has been central to the agenda for both
+personalized (or precision) medicine and comparative effectiveness research,
+giving rise to a wide range of approaches for its discovery, evaluation and
+application in clinical practice. While exploratory and confirmatory HTE
+analyses focus on generating and testing hypotheses on subgroup effects,
+predictive HTE analyses use all available patient data to predict the benefits
+or harms of treatment in individual patients [@Varadhan2013;@Kent2018].
 
-While these approaches were developed for application in randomized controlled
-trials (RCTs), observational databases are also an appealing substrate.
-Observational healthcare databases, such as administrative claims and electronic
-health records, are already highly available for the analysis of
-pharmacoepidemiologic research questions [@AdlerMilstein2017; @Dahabreh2014].
-They are also often larger than many typical trials, providing excellent power
-for HTE analysis, including heterogeneous populations. However, unlike RCTs,
-treatment effects are subject to confounding, while the unique structure of
-different databases calls for database-specific analysis plans that often are
-not easily transportable. Because of the latter issue, running analyses at scale
-demands a big investment of time and effort, as researchers are forced to map
-their analysis plans to the databases available to them.
+Baseline risk has been recognized as a robust predictor of treatment effect,
+because of the inherent relationship between baseline risk and the absolute
+effects of treatment
+[@Rothwell2005;@Hayward2006;@Kent2007;@Kent2008;@Kent2010]. For example, an
+invasive coronary procedure – in comparison with medical treatment – improves
+survival in patients with myocardial infarction at high (predicted) baseline
+risk but not in those at low baseline risk [@Thune2005]. It has also been shown
+that high-risk patients with pre-diabetes benefit substantially more from a
+lifestyle modification program than low-risk patients [@Sussman2015].
 
-The Observational Health Data Sciences and Informatics (OHDSI) collaborative has
-established an international network of data partners and researchers that aim
-to bring out the value of health data through large-scale analytics by mapping
-all available databases to the Observational Medical Outcomes Partnership (OMOP)
-Common Data Model (CDM) [@hripcsak2015observational; @Overhage2012].
+Recently, systematic guidance on the application of risk-based methods has been
+developed for RCT data [@Kent2019;@PathEnE]. Patients are divided into risk strata using
+either an existing or an internally developed risk prediction
+model. Risk-stratum-specific estimates provide an overview of the evolution of
+treatment effects with increasing risk both on the relative and the absolute
+scale. Several methods for predictive HTE analysis have been adapted for use in
+observational data, but risk-based methods for predictive HTE in observational
+data are still lacking and have been highlighted as an important future research
+need [@PathEnE].
 
-We aimed to develop a framework for implementing a risk-based predictive
-approach for evaluating HTE in high-dimensional observational data, extending
-the existing guidelines of the RCT setting. Our publicly available package
-provides an out-of-the-box solution for implementing such analyses at scale
-within the OHDSI network, taking advantage of the OMOP-CDM. We implemented the
-framework using existing OHDSI methods including the patient-level prediction
-framework and the population-level effect estimation framework based on new-user
-cohort design [@Reps2018; @Ryan2013]. As a proof-of-concept we analyzed
-heterogeneity of the effects of first-line hypertension treatment: we compared
-the effect of angiotensin converting enzyme (ACE) inhibitors to beta blockers
-on 9 outcomes across three different US claims databases.
+We aimed to develop a framework for risk-based assessment of treatment effect
+heterogeneity in observational healthcare databases, extending the existing
+methodology developed for the RCT setting. We implemented the framework in a
+publicly available package providing an out-of-the-box solution for implementing
+such analyses at scale within any observational database mapped to OMOP-CDM. In
+a case study we analyzed heterogeneity of the effects of first-line hypertension
+treatment: we compared the effect of angiotensin converting enzyme (ACE)
+inhibitors to beta blockers on 9 outcomes across three different US claims
+databases.
 
-# Materials and Methods
+# Methods
 
@@
 databases mapped to the OMOP-CDM. These are: 1) general definition of the
-research aim; 2) identification of the database within which the analyses will
+research aim; 2) identification of the databases within which the analyses will
 be performed; 3) a prediction step where internal or external prediction models
@@
 implementation of the proposed framework and made it publicly available. The
-source code can be found at <https://github.com/OHDSI/RiskStratifiedEstimation>.
+source code can be found at https://github.com/OHDSI/RiskStratifiedEstimation.
 
-## Step 1: General definition of the problem
+## Step 1: General definition of the research aim
 
-The typical research aim is: "to compare the effect of treatment $T$ to a
-comparator treatment $C$ in patients with disease $D$ with respect to outcomes
-$O_1,\dots,O_n$".
-
-Cohort definitions are crucial for this step of the framework. We define a
-cohort as the set of patients who satisfy one or more inclusion criteria for a
-duration of time. A cohort within the OHDSI setting is more than a set of
-specific clinical codes, providing a definition of a logic for how to use that
-code set. All cohort definitions consist of: an entry event, i.e. the time a
-patient enters a cohort; a set of inclusion criteria applied to the initial
-event cohort to further restrict the set of people, resulting in the
-construction of the construction of the qualifying cohort; cohort exit criteria
-that terminate the patient's presence in the cohort. Cohort definitions are
-transportable, meaning that in theory they can be implemented in any database,
-provided that it is mapped to the OMOP-CDM.
+The typical research aim is: “to compare the effect of treatment to a comparator
+treatment in patients with disease with respect to outcomes $O_1,\dots,O_n$”.
 
 Our framework uses a comparative cohort design. This means that at least 3
-cohorts of patients need to be defined at this stage of the framework.
+cohorts of patients need to be defined at this stage of the framework:
 
-  - A single treatment cohort ($T$) which includes patients with disease $D$
-    receiving the target treatment of interest. For example, a set of
-    hypertension patients within a database that receive angiotensin-converting
-    enzyme inhibitors, followed from the time of initiation until the time of
-    censoring.
-  - A single comparator cohort ($C$) which includes patients with disease $D$
-    receiving the comparator (control) treatment. For example, a set of patients
-    in a database that receive beta blockers, followed from the time of
-    initiation until the time of censoring.
+- A single treatment cohort ($T$) which includes patients with disease receiving
+  the target treatment of interest.
+- A single comparator cohort ($C$) which includes patients with disease receiving
+  the comparator treatment.
 - One or more outcome cohorts ($O_1,\dots,O_n$) that contain patients
-    developing the outcomes of interest. For example, the set of patients in a
-    database that have at least one occurrence of acute myocardial infarction
-    (MI) in their record.
+  developing the outcomes of interest
 
-## Step 2: Identification of the database
+## Step 2: Identification of the databases {#methods2}
 
-The aim of this step is the inclusion of databases that represent the patient
-population of interest. It is required that the databases are mapped to the
-OMOP-CDM. The inclusion of multiple databases potentially increases the
-generalizability of results. Furthermore, the cohorts should preferably have
-adequate sample size with adequate follow-up time to ensure precise effect
-estimation, even within smaller risk strata. Other issues that may be of
-importance for database inclusion are the depth of data capture (the precision
-at which measurements, lab tests, conditions are recorded), the reliability of
-data entry and many more, also depending on the task at hand.
+Including in our analyses multiple databases representing the population of
+interest potentially increases the generalizability of results. Furthermore, the
+cohorts should preferably have adequate sample size with adequate follow-up time
+to ensure precise effect estimation, even within smaller risk strata. Other
+issues that may be of importance for database inclusion are the depth of data
+capture (the precision at which measurements, lab tests, conditions are
+recorded) and the reliability of data entry.
 
-## Step 3: Prediction
+## Step 3: Prediction {#methods3}
 
@@
 prediction models using observational data that ensures adherence to existing
-guidelines [@Collins2015; @Moons2015]. This prediction framework requires the
-definition of two essential cohorts: a target cohort, i.e. a set of patients
-that satisfy one or more inclusion criteria for a duration of time, and an
-outcome cohort.
-
-To generate the target cohort we pool the already defined treatment cohort and
-comparator cohort. Further restrictions can be applied on the target cohort to
-construct the final population on which the prediction model will be developed
-(e.g. exclude patients with a prior outcome in their history, before being
-included in the target cohort). To avoid deferentially fitting the prediction
-model to patients across treatment arms, thus introducing spurious interactions
-with treatment [@Burke2014; @vanKlaveren2019], we develop the patient-level
-prediction model in the propensity score-matched (1:1) subset of the population.
-
-More specifically, we first estimate propensity scores using LASSO logistic
-regression and a large set of baseline covariates including demographics, drug
-exposures, diagnoses, measurements and medical devices. We match patients 1-1
-using a caliper, i.e. the maximum distance that is acceptable for any match. The
-default value we use is 0.2 on the standardized logit scale for the propensity
-scores. Other methods of fitting the propensity scores, such as random forest
-and others can also be considered.
-
-Finally, we need to define the time horizon within which we aim to make
-predictions and we also need to select the machine-learning algorithm we want to
-use to generate patient-level predictions. Currently, the available options are
-regularized logistic regression, random forest, gradient boosting machines,
-decision tree, naive Bayes, K-nearest neighbors, neural network and deep
-learning (convolutional neural networks, recurrent neural network and deep
-nets).
+guidelines [@Collins2015;@Moons2015]. To generate the target cohort (i.e. the
+set of patients on which the prediction model will be developed), we pool the
+already defined treatment cohort and comparator cohort. To avoid deferentially
+fitting the prediction model to patients across treatment arms, thus introducing
+spurious interactions with treatment [@Burke2014;@vanKlaveren2019], we develop
+the patient-level prediction model in the propensity score-matched (1:1) subset
+of the sample. Finally, we need to define the time horizon within which we aim
+to make predictions and we also need to select the machine-learning algorithm we
+want to use to generate patient-level predictions.
 
@@
 including discrimination and calibration both in the propensity score matched
-subset, the entire population and separately for treated and comparator patients
-should also be reported. This is important to ensure that no overfitting of the
-prediction model in one of the cohorts has occurred. In addition, the
-performance of the prediction models is directly related to our ability to
-single out patient subgroups where treatment may be highly beneficial or unsafe.
-Kent et al [@Kent2016] demonstrated that the event rate and the discriminative
-ability of the prediction model can predict very well the distribution of
-predicted risk. Lower event rate and higher c-statistic (given good calibration)
-result in high risk heterogeneity, thus making estimated average treatment
-effects uninformative. In this case, risk stratified analysis of HTE can be more
-effective in singling out patient subgroups that stand to benefit (or be harmed)
-most by treatment in question.
+subset, the entire sample and separately for treated and comparator patients
+should also be reported.
 
-## Step 4: Estimation
+## Step 4: Estimation {#methods4}
 
-The aim of this step is the estimation of treatment effects (both on the
-relative and the absolute scale) within risk strata---typically 4 risk
-quarters---defined using the prediction model of step 3. Effect estimation may
-be focused on the difference in outcomes for a randomly selected person from the
-risk stratum (average treatment effect) or for a randomly selected person from
-the treatment cohort within the risk stratum receiving the treatment under study
-(average treatment effect on the treated).
+We estimate treatment effects (both on the relative and the absolute scale)
+within risk strata—typically 4 risk quarters—defined using the prediction model
+of step 3. Effect estimation may be focused on the difference in outcomes for a
+randomly selected person from the risk stratum (average treatment effect) or for
+a randomly selected person from the treatment cohort within the risk stratum
+receiving the treatment under study (average treatment effect on the treated).
 
-Any appropriate method for the evaluation of relative and absolute treatment
+Any appropriate method for the analysis of relative and absolute treatment
 effects can be considered, as long as the this is done consistently in all risk
-strata. Common approaches are odds ratios or hazard ratios for relative scale
-estimates and differences in observed proportions or differences in Kaplan-Meier
-estimates for absolute scale estimates, depending on the problem at hand. We
-estimate propensity scores within risk strata which we then use to match
-patients from different treatment cohorts or stratify them into groups with
-similar propensity scores or to weigh each patient's contribution to the
+strata. Common statistical metrics are odds ratios or hazard ratios for relative
+scale estimates and differences in observed proportions or differences in
+Kaplan-Meier estimates for absolute scale estimates, depending on the problem at
+hand. We estimate propensity scores within risk strata which we then use to
+match patients from different treatment cohorts or to stratify them into groups
+with similar propensity scores or to weigh each patient’s contribution to the
 estimation process [@Austin2011].
 
-Before focusing on the results of the estimation process we need to evaluate if
-adequate covariate balance was achieved within each risk stratum accounting for
-measured confounding. Common approaches include evaluation of the overlap of
-propensity score distributions and calculation of standardized covariate
-differences before and after propensity score adjustment.
+Prior to analyzing the results, we need to assess if adequate covariate balance
+was achieved within each risk stratum accounting for measured
+confounding. Common approaches include analysis of the overlap of propensity
+score distributions and calculation of standardized mean differences of the
+covariates before and after propensity score adjustment.
 
 A schematic overview of the prediction and estimation steps is presented in
-Figure \@ref(fig:figure1).
+Figure XX.
 
-```{r figure1, echo=FALSE, fig.cap="(A) Starting from a treatment (top), a comparator (bottom) and an outcome (middle) cohort we estimate the propensity scores on the entire target population.
-(B) We match patients on the propensity scores and estimate the prediction model.
-Since we match patients we develop the prediction model on smaller subset of the initial population and, therefore, the number of patients is smaller in B compared to A.
-(C) We apply the prediction model on the entire population (green: lower 25\\% of the risk distribution; yellow: patients with risk between 25\\% and 50\\% of the risk distribution; orange: patients with risk between 50\\% and 75\\% of the risk distribution; red: patients at risk higher than 75\\% of the risk distribution).
-(D) We separate in risk subgroups, here quarters.
-Within risk quarters propensity scores are estimated again and relative and absolute treatment effects are estimated.", out.width = '100%'}
-# knitr::include_graphics(here("figures/Figure_1.tif"))
-grid::grid.raster(tiff::readTIFF(here("figures/Figure_1.tif")))
-```
-
-## Step 5: Presentation of results
+## Step 5: Presentation of results {#methods5}
 
@@
 
+## Case study
+
+As a demonstration of our framework, we evaluated if our proposed method was
+able to identify treatment effect heterogeneity of ACE inhibitors compared to
+beta blockers using acute myocardial infarction (MI) risk quarter specific
+effect estimates, both on the relative and on the absolute scale. We focused on
+3 efficacy outcomes (acute MI, hospitalization with heart failure and ischemic
+or hemorrhagic stroke) and 6 safety outcomes (hypokalemia, hyperkalemia,
+hypotension, angioedema, cough and abnormal weight gain). We used data from 3
+US-based claims databases. The analysis plan was the framework outlined in
+steps 1 through 5.
+
+A recent meta-analysis has shown that beta-blockers are on average less
+efficacious in preventing cardiovascular events than ACE inhibitors
+[@Wiysonge2017]. However, meta-analyses focusing on age of the patients have
+shown that beta blockers have similar efficacy in younger hypertensive patients,
+but should not be used as first-line treatment of hypertension in older patients
+mainly due to their increased risk of cardiovascular events compared to other
+first-line antihypertensive treatments [@Khan2006;@Vgele2017].Our aim was to
+replicate these conclusions in our analyses.
+
 # Results
 
-As a proof of concept, we focus on the comparison of angiotensin converting
-enzyme (ACE) inhibitors to beta blockers are among the most common treatment
-classes for hypertension, with well-established effectiveness. Beta blockers,
-even though initially widely used for the treatment of hypertension, more recent
-trials and meta-analyses have cast doubt on their relative effectiveness
-[@wiysonge2017beta]. As a result, newer US guidelines do not consider beta
-blockers for initial treatment for hypertension while in the EU guidelines
-combination with other antihypertensive treatments is recommended
-[@Whelton2018; @Williams2018]. However, another meta-analysis suggested that
-the efficacy profile of beta blockers is similar to other major treatment
-classes in younger hypertensive patients and, thus, countries like Canada still
-include them as a first-line treatment candidate [@Khan2006; @Rabi2020].
+## Step 1: General definition of the research aim
 
-## Step 1: General definition of the problem
-
-We consider the following research aim: "compare the effect of ACE-inhibitors
-($T$) to the effect of beta blockers ($C$) in patients with established
-hypertension ($D$) with respect to 9 outcomes ($O_1,\dots,O_9$)". The cohorts
-are:
+We considered the following research aim: “compare the effect of ACE-inhibitors
+($T$) to the effect of beta blockers ($C$) in patients with established hypertension
+with respect to 9 outcomes ($O_1,\dots,O_9$)”. The cohorts are:
 
 - Treatment cohort: Patients receiving any drug within the ACE-inhibitor class
-    with at least one year of follow-up before treatment initiation and a
-    recorded hypertension diagnosis within that year.
+  with at least one year of follow-up before treatment initiation and a recorded
+  hypertension diagnosis within that year.
 - Comparator cohort: Patients receiving any drug within the beta blocker class
-    with at least one year of follow-up before treatment initiation and a
-    recorded hypertension diagnosis within that year.
-  - Outcome cohorts: We consider 3 main and 6 safety outcome cohorts. These are
-    patients in the database with a diagnosis of: acute MI; hospitalization with
-    heart failure; ischemic or hemorrhagic stroke (efficacy outcomes);
+  with at least one year of follow-up before treatment initiation and a recorded
+  hypertension diagnosis within that year.
+- Outcome cohorts: We considered 3 efficacy and 6 safety outcome cohorts. These
+  were patients in the database with a diagnosis of: acute MI; hospitalization
+  with heart failure; ischemic or hemorrhagic stroke (efficacy outcomes);
   hypokalemia; hyperkalemia; hypotension; angioedema; cough; abnormal weight
@@
 All cohort definitions were identical to the ones used in the multinational
-study carried out within OHDSI that provided overall treatment effect estimates
-comparing all anti-hypertensive drug classes with each other [@Suchard2019].
-More information can be found in the supplementary material.
+study that provided overall treatment effect estimates comparing all
+anti-hypertensive drug classes with each other [@Suchard2019]. More information
+can be found in the supplementary material.
 
@@
 
-We used the following databases:
-
-  - IBM MarketScan Medicare Supplemental Beneficiaries (MDCR): Represents health
-    services of retirees (aged 65 or older) in the United States with primary or
-    Medicare supplemental coverage through privately insured fee-for-service,
-    point-of-service or capitated health plans. These data include adjudicated
-    health insurance claims (e.g. inpatient, outpatient and outpatient
-    pharmacy). Additionally, it captures laboratory tests for a subset of the
-    covered lives.
-  - IBM MarketScan Medicaid (MDCD): Adjudicated US health insurance claims for
-    Medicaid enrollees from multiple states. It includes hospital discharge
-    diagnoses, outpatient diagnoses and procedures and outpatient pharmacy
-    claims as well as ethnicity and Medicare eligibility.
-  - IBM MarketScan Commercial Claims and Encounters (CCAE): Data from
-    individuals enrolled in US employer-sponsored insurance health plans. The
-    data includes adjudicated health insurance claims (e.g. inpatient,
-    outpatient, and outpatient pharmacy) as well as enrollment data from large
-    employers and health plans who provide private healthcare coverage to
-    employees, their spouses and dependents. Additionally, it captures
-    laboratory tests for a subset of the covered lives.
-
-Our analyses included a total of 784,561, 66,820 and 101,661 patients initiating
-treatment with ACE inhibitors and 395,740, 45,999 and 69,798 patients initiating
-treatment with beta blockers in CCAE, MDCD and MDCD respectively (Table
-\@ref(tab:table)). Adequate numbers of patients were included in all strata of
-predicted acute MI risk.
-
-\captionsetup{width=\textwidth}
-```{r table, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, results='asis', booktabs=TRUE}
-table <- readr::read_csv(
-  here::here("data/processed/table1.csv"),
-  na = character()
-) %>%
-  select(-database)
-
-knitr::kable(
-  table,
-  format = "latex",
-  longtable = TRUE,
-  booktabs = TRUE,
-  col.names = c(
-    "Outcome", "Risk quarter",
-    "Persons", "Person years", "Events",
-    "Persons", "Person years", "Events"
-  ),
-   format.args = list(
-    big.mar = ","
-  ),
-  caption = "Number of patients, person years and events within quarters of predicted risk for acute MI for the 3 main outcomes of the study (acute MI, hospitalization with heart failure and ischemic or hemorrhagic stroke) across the 3 databases.
-The populations used for the evaluation of hosptialization with heart failure and stroke are subsets of the populations used for the evaluation of actute MI."
-) %>%
-  kableExtra::group_rows("CCAE", 1, 12) %>%
-  kableExtra::group_rows("MDCD", 13, 24) %>%
-  kableExtra::group_rows("MDCR", 25, 36) %>%
-  kableExtra::add_header_above(
-    c(
-      "", "", 
-      "ACE inhibitors" = 3,
-      "Beta blockers"  = 3
-    )
-  )
-```
+For our analyses we used data from 3 US claims databases, namely IBM MarketScan
+Commercial Claims and Encounters (CCAE), IBM MarketScan Medicaid (MDCD), and IBM
+MarketScan Medicare Supplemental Beneficiaries (MDCR). Our analyses included a
+total of 784,561, 66,820 and 101,661 patients initiating treatment with ACE
+inhibitors and 395,740, 45,999 and 69,798 patients initiating treatment with
+beta blockers in CCAE, MDCD and MDCR respectively (Table 1). Adequate numbers of
+patients were included in all strata of predicted acute MI risk.
 
@@
 We internally developed separate prediction models for acute MI in all 3
-databases.
-
-The prediction models were estimated on the propensity score matched (1:1)
-subset of the population, using caliper of 0.2 and after excluding patients
-having the outcomes any time prior to treatment initiation. We chose a time
-horizon of 2 years after inclusion into the target cohort. For this
-demonstration, we developed the prediction models using LASSO logistic
-regression with 3-fold cross validation for hyper-parameter selection. For this
-demonstration, we developed the prediction models using LASSO logistic
-regression with 3-fold cross validation for hyper-parameter selection.
+databases. The prediction models were estimated on the propensity score matched
+(1:1) subset of the sample, using caliper of 0.2 and after excluding patients
+having the outcome any time prior to treatment initiation. We chose a time
+horizon of 2 years after inclusion into the target cohort. We developed the
+prediction models using LASSO logistic regression with 3-fold cross validation
+for hyper-parameter selection.
 
@@
 for the case of CCAE database in which the derived prediction model performed
-better in the comparator cohort (Table \@ref(tab:tableAucs)). We also observed
-lower performance of the prediction model developed in MDCR compared to the
-other 2 databases. Results on the calibration of the prediction models can be
-found in the supplement.
-
-\captionsetup{width=\textwidth}
-```{r tableAucs, cache=TRUE, echo=FALSE, results='asis', booktabs=TRUE}
-table <- tibble(
-  population = c(
-    "Matched set",
-    "Treatment cohort",
-    "Comparator cohort",
-    "Entire population"
-  ),
-  ccae = c(
-    75.78, 70.65, 79.1, 74.06
-  ),
-  mdcd = c(
-    78.12, 78.29, 80.26, 79.46
-  ),
-  mdcr = c(
-    71.09, 69.03, 68.53, 69.07
-  )
-)
-
-knitr::kable(
-  table,
-  format = "latex",
-  longtable = TRUE,
-  booktabs = TRUE,
-  col.names = c(
-    "Population", "CCAE", "MDCD", "MDCR"
-  ),
-   format.args = list(
-    big.mar = ","
-  ),
-  caption = "Discriminiative ability (c-statistic) of the derived prediction models for acute myocardial infarction in the matched set (development set), the treatment cohort, the comparator cohort and the etnire population in CCAE, MDCD and MDCR"
-)
-```
+better in the comparator cohort (Table 2). We also observed lower performance of
+the prediction model developed in MDCR compared to the other 2 databases (see
+supplementary material).
 
@@
 
-We used patient-level predictions to stratify the patient population into 4 risk
-quarters. Within risk strata, relative effects were estimated using Cox
+We used patient-level predictions to stratify the sample into 4 MI risk
+quarters. Within risk quarters, relative effects were estimated using Cox
 regression and absolute effects were estimated from the Kaplan-Meier estimate
@@
 approach as in the development of prediction models. We used the estimated
-propensity scores to stratify patients into 5 strata, within each risk quarter.
+propensity scores to stratify patients into 5 strata, within each risk
+quarter. The risk quarter-specific effect estimates were derived by averaging
+over the estimates within the propensity score fifths.
 
-In general, there was sufficient overlap of propensity score distribution in all
-risk strata (Figure \@ref(fig:psDensity)). If empirical equipoise was not
-achieved, the validity of the comparative effectiveness estimates would be
-questionable.
-
-```{r psDensity, cache = TRUE, echo=FALSE, fig.cap="Preference score distributions for the evaluation of heterogeneity of the effect of ACE inhibitors compared to beta blockers on acute MI based on quarters of predicted acute MI risk.
-The preference score is a transformation of the propensity score that adjusts for prevalence differences between populations.", out.width = '100%'}
-grid::grid.raster(tiff::readTIFF(here("figures/PsDensity.tiff")))
-```
-
-Propensity score adjustment achieved balance for most of the considered
-covariates, measured using standardized mean differences before and after
-adjustment (Figure \@ref(fig:covariateBalance)). However, in lower risk strata
-imbalances persisted for a substantial subset of the covariates, all related to
-pregnancy findings. This was anticipated (use of ACE inhibitors is specifically
+There was sufficient overlap of propensity score distribution in all risk strata
+(Figure 2). Propensity score adjustment achieved balance for most of the
+considered covariates, measured using standardized mean differences before and
+after adjustment (Figure 3). However, in lower risk strata imbalances persisted
+for a substantial subset of the covariates, all related to pregnancy
+findings. This was anticipated (use of ACE inhibitors is specifically
 contraindicated during pregnancy) and was already pointed out in
-[@Suchard2019].
-
-```{r covariateBalance, cache = TRUE, echo=FALSE, fig.cap="Patient characteristic balance for ACE inhibitors and beta blockers before and after stratification on the propensity scores.
-Each dot represents the standardized difference of means for a single covariate before (x-axis) and after (y-axis) stratification", out.width = '100%'}
-grid::grid.raster(tiff::readTIFF(here("figures/CovariateBalance.tiff")))
-```
+[@Suchard2019]. Therefore, treatment effect estimates in the lowest risk
+quarters of CCAE and MDCR may still be subject to residual confounding and
+should be interpreted with care.
 
 ## Step 5: Presentation of results
+
+```{r, echo=FALSE}
+map_outcomes <- readRDS(here("data/processed/multipleRseeAnalyses/map_outcomes.rds"))
+map_exposures <- readRDS(here("data/processed/multipleRseeAnalyses/map_exposures.rds"))
+relative <- readRDS(here("data/processed/multipleRseeAnalyses/mappedOverallRelativeResults.rds")) %>%
+  left_join(map_outcomes, by = c("estOutcome" = "outcome_id")) %>%
+  select(-estOutcome) %>%
+  rename("estOutcome" = "outcome_name") %>%
+  left_join(map_outcomes, by = c("stratOutcome" = "outcome_id")) %>%
+  select(-stratOutcome) %>%
+  rename("stratOutcome" = "outcome_name") %>%
+  left_join(map_exposures, by = c("treatment" = "exposure_id")) %>%
+  select(-treatment) %>%
+  rename("treatment" = "exposure_name") %>%
+  left_join(map_exposures, by = c("comparator" = "exposure_id")) %>%
+  select(-comparator) %>%
+  rename("comparator" = "exposure_name")
+
+absolute <- readRDS(here("data/processed/multipleRseeAnalyses/mappedOverallAbsoluteResults.rds")) %>%
+  left_join(map_outcomes, by = c("estOutcome" = "outcome_id")) %>%
+  select(-estOutcome) %>%
+  rename("estOutcome" = "outcome_name") %>%
+  left_join(map_outcomes, by = c("stratOutcome" = "outcome_id")) %>%
+  select(-stratOutcome) %>%
+  rename("stratOutcome" = "outcome_name") %>%
+  left_join(map_exposures, by = c("treatment" = "exposure_id")) %>%
+  select(-treatment) %>%
+  rename("treatment" = "exposure_name") %>%
+  left_join(map_exposures, by = c("comparator" = "exposure_id")) %>%
+  select(-comparator) %>%
+  rename("comparator" = "exposure_name") %>%
+  mutate(
+    estimate = 100 * estimate,
+    lower    = 100 * lower,
+    upper    = 100 * upper
+  )
+
+acute_mi_relative <- relative %>%
+  filter(
+    stratOutcome == "acute_myocardial_infarction",
+    estOutcome == "acute_myocardial_infarction"
+  )
+
+acute_mi_absolute <- absolute %>%
+  filter(
+    stratOutcome == "acute_myocardial_infarction",
+    estOutcome == "acute_myocardial_infarction"
+  )
+acute_cough_relative <- relative %>%
+  filter(
+    stratOutcome == "acute_myocardial_infarction",
+    estOutcome == "cough"
+  )
+
+acute_cough_absolute <- absolute %>%
+  filter(
+    stratOutcome == "acute_myocardial_infarction",
+    estOutcome == "cough"
+  )
+```
 
@@
 decreased) with increasing acute MI risk. More specifically, hazard ratios
-decreased from 0.98 (0.77 to 1.26; 95% CI), 1.30 (0.51 to 3.22; 95% CI) and 1.03
-(0.82 to 1.29; 95% CI) to 0.76 (0.71 to 0.82; 95% CI), 0.94 (0.82 to 1.07; 95%
-CI) and 1.03 (0.93 to 1.15; 95% CI) in CCAE, MDCD and MDCR respectively (Figure
-\@ref(fig:combinedRelative)). In terms of hospitalization with heart failure
-relative treatment effect estimates favored ACE inhibitors across all risk
-strata in all databases. We found no differences between the two treatments in
-their effect on stroke on the relative scale. In terms of the safety outcomes we
-found an increased risk of cough and angioedema on the relative scale across all
-risk strata. In the case of cough, this effect decreased with increasing risk of
-acute MI---from 1.37 (1.33 to 1.41; 95% CI), 1.35 (1.24 to 1.48; 95% CI)
-and 1.37 (1.29 to 1.45; 95% CI) in the lowest acute MI risk quarter to 1.26
-(1.22 to 1.29; 95% CI), 1.07 (1.00 to 1.14) and 1.10 (1.04 to 1.17; 95% CI) in
-the highest acute MI risk quarter in CCAE, MDCD and MDCR respectively.
-
-```{r combinedRelative, cache=TRUE, echo=FALSE, fig.cap="Overview of heterogeneity of ACE-inhibitors treatment on the relative scale (hazard ratios) within strata of predicted risk of acute MI.
-Values below 1 favor ACE inhibitors, while values above 1 favor beta blockers.", out.width = '100%'}
-grid::grid.raster(tiff::readTIFF(here("figures/CombinedRelative.tiff")))
-```
+decreased from 
+`r acute_mi_relative %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(estimate)`
+(`r acute_mi_relative %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(lower)` to
+`r acute_mi_relative %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(upper)`;
+95\% CI) and
+`r acute_mi_relative %>% filter(database == "mdcd", riskStratum == "Q1") %>% pull(estimate)`
+(`r acute_mi_relative %>% filter(database == "mdcd", riskStratum == "Q1") %>% pull(lower)` to
+`r acute_mi_relative %>% filter(database == "mdcd", riskStratum == "Q1") %>% pull(upper)`;
+95\% CI) to
+`r acute_mi_relative %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(estimate)`
+(`r acute_mi_relative %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(lower)` to
+`r acute_mi_relative %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(upper)`;
+95\% CI),
+`r acute_mi_relative %>% filter(database == "mdcd", riskStratum == "Q4") %>% pull(estimate)`
+(`r acute_mi_relative %>% filter(database == "mdcd", riskStratum == "Q4") %>% pull(lower)` to
+`r acute_mi_relative %>% filter(database == "mdcd", riskStratum == "Q4") %>% pull(upper)`;
+95\% CI) in CCAE and MDCD respectively (Figure 4). In MDCR hazard ratios remained stable
+from
+`r acute_mi_relative %>% filter(database == "mdcr", riskStratum == "Q1") %>% pull(estimate)`
+(`r acute_mi_relative %>% filter(database == "mdcr", riskStratum == "Q1") %>% pull(lower)` to
+`r acute_mi_relative %>% filter(database == "mdcr", riskStratum == "Q1") %>% pull(upper)`;
+95\% CI) in the lowest MI risk quarter to 
+`r acute_mi_relative %>% filter(database == "mdcr", riskStratum == "Q4") %>% pull(estimate)`
+(`r acute_mi_relative %>% filter(database == "mdcr", riskStratum == "Q4") %>% pull(lower)` to
+`r acute_mi_relative %>% filter(database == "mdcr", riskStratum == "Q4") %>% pull(upper)`;
+95\% CI). In terms of hospitalization with heart failure relative treatment
+effect estimates favored ACE inhibitors across all risk strata in all
+databases. We found no differences between the two treatments in their effect on
+stroke on the relative scale. In terms of the safety outcomes, we found an
+increased ACE-inhibitor risk of cough and angioedema on the relative scale
+across all risk strata. In the case of cough, this effect decreased with
+increasing risk of acute MI---from 
+`r acute_cough_relative %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(estimate)`
+(`r acute_cough_relative %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(lower)` to
+`r acute_cough_relative %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(upper)`;
+95\% CI),
+`r acute_cough_relative %>% filter(database == "mdcd", riskStratum == "Q1") %>% pull(estimate)`
+(`r acute_cough_relative %>% filter(database == "mdcd", riskStratum == "Q1") %>% pull(lower)` to
+`r acute_cough_relative %>% filter(database == "mdcd", riskStratum == "Q1") %>% pull(upper)`;
+95\% CI), and
+`r acute_cough_relative %>% filter(database == "mdcr", riskStratum == "Q1") %>% pull(estimate)`
+(`r acute_cough_relative %>% filter(database == "mdcr", riskStratum == "Q1") %>% pull(lower)` to
+`r acute_cough_relative %>% filter(database == "mdcr", riskStratum == "Q1") %>% pull(upper)`;
+95\% CI) to
+`r acute_cough_relative %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(estimate)`
+(`r acute_cough_relative %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(lower)` to
+`r acute_cough_relative %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(upper)`;
+95\% CI),
+`r acute_cough_relative %>% filter(database == "mdcd", riskStratum == "Q4") %>% pull(estimate)`
+(`r acute_cough_relative %>% filter(database == "mdcd", riskStratum == "Q4") %>% pull(lower)` to
+`r acute_cough_relative %>% filter(database == "mdcd", riskStratum == "Q4") %>% pull(upper)`;
+95\% CI), and
+`r acute_cough_relative %>% filter(database == "mdcr", riskStratum == "Q4") %>% pull(estimate)`
+(`r acute_cough_relative %>% filter(database == "mdcr", riskStratum == "Q4") %>% pull(lower)` to
+`r acute_cough_relative %>% filter(database == "mdcr", riskStratum == "Q4") %>% pull(upper)`;
+95\% CI) in CCAE, MDCD, and MDCR, respectively.
 
@@
 increasing acute MI risk in favor of ACE inhibitors in terms of acute MI in all
-databases except for MDCR---from -0.03% (-0.08% to 0.01%; 95% CI), -0.05%
-(-0.18% to 0.08%; 95% CI) and -0.02% (-0.24% to 0.19%; 95% CI) in the lowest
-acute MI risk quarter to 0.54% (0.36% to 0.71%; 95% CI), 0.29% (-0.39%
-to 0.97%; 95% CI) and -0.39% (-0.96% to 0.18%; 95% CI) in the highest acute MI
-risk quarter in CCAE, MDCD and MDCR, respectively (Figure \@ref(fig:figure12)).
-We found no difference on the absolute scale for stroke across risk strata.
-Absolute risk differences did not favor ACE inhibitors compared to beta blockers
-in terms of cough, even though this effect again diminished with increasing
-acute MI risk---from -4.14% (-4.62% to -3.66%; 95% CI), -6.45% (-9.12% to
--3.78%; 95% CI) and -4.81% (-5.76% to -3.85%; 95% CI) in the lowest acute MI
-risk quarter to -2.57% (-2.99% to -2.15%; 95% CI), -1.11% (-2.93% to 0.70%; 95%
-CI) and -1.69% (-2.83% to -0.55%; 95% CI) in the highest acute MI risk quarter
-in CCAE, MDCD and MDCR, respectively. In terms of angioedema absolute risk
-differences were very small due to the rarity of the outcome.
+databases except for MDCR---from
+`r acute_mi_absolute %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(estimate)`\%
+(`r acute_mi_absolute %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(lower)`\% to
+`r acute_mi_absolute %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(upper)`\%;
+95\% CI), 
+`r acute_mi_absolute %>% filter(database == "mdcd", riskStratum == "Q1") %>% pull(estimate)`\%
+(`r acute_mi_absolute %>% filter(database == "mdcd", riskStratum == "Q1") %>% pull(lower)`\% to
+`r acute_mi_absolute %>% filter(database == "mdcd", riskStratum == "Q1") %>% pull(upper)`\%;
+95\% CI), and
+`r acute_mi_absolute %>% filter(database == "mdcr", riskStratum == "Q1") %>% pull(estimate)`\%
+(`r acute_mi_absolute %>% filter(database == "mdcr", riskStratum == "Q1") %>% pull(lower)`\% to
+`r acute_mi_absolute %>% filter(database == "mdcr", riskStratum == "Q1") %>% pull(upper)`\%;
+95\% CI) in the lowest acute MI risk quarter to
+`r acute_mi_absolute %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(estimate)`\%
+(`r acute_mi_absolute %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(lower)`\% to
+`r acute_mi_absolute %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(upper)`\%;
+95\% CI),
+`r acute_mi_absolute %>% filter(database == "mdcd", riskStratum == "Q4") %>% pull(estimate)`\%
+(`r acute_mi_absolute %>% filter(database == "mdcd", riskStratum == "Q4") %>% pull(lower)`\% to
+`r acute_mi_absolute %>% filter(database == "mdcd", riskStratum == "Q4") %>% pull(upper)`\%;
+95\% CI), and
+`r acute_mi_absolute %>% filter(database == "mdcr", riskStratum == "Q4") %>% pull(estimate)`\%
+(`r acute_mi_absolute %>% filter(database == "mdcr", riskStratum == "Q4") %>% pull(lower)`\% to
+`r acute_mi_absolute %>% filter(database == "mdcr", riskStratum == "Q4") %>% pull(upper)`\%;
+95\% CI) in the highest acute MI risk quarter in CCAE, MDCD, and MDCR,
+respectively (Figure 5). We found no difference on the absolute scale for stroke
+across risk strata. Absolute risk differences did not favor ACE inhibitors
+compared to beta blockers in terms of cough, even though this effect again
+diminished with increasing acute MI risk---from 
+`r acute_cough_absolute %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(estimate)`\%
+(`r acute_cough_absolute %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(lower)`\% to
+`r acute_cough_absolute %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(upper)`\%;
+95\% CI),
+`r acute_cough_absolute %>% filter(database == "mdcd", riskStratum == "Q1") %>% pull(estimate)`\%
+(`r acute_cough_absolute %>% filter(database == "mdcd", riskStratum == "Q1") %>% pull(lower)`\% to
+`r acute_cough_absolute %>% filter(database == "mdcd", riskStratum == "Q1") %>% pull(upper)`\%;
+95\% CI), and
+`r acute_cough_absolute %>% filter(database == "mdcr", riskStratum == "Q1") %>% pull(estimate)`\%
+(`r acute_cough_absolute %>% filter(database == "mdcr", riskStratum == "Q1") %>% pull(lower)`\% to
+`r acute_cough_absolute %>% filter(database == "mdcr", riskStratum == "Q1") %>% pull(upper)`\%;
+95\% CI) in the lowest acute MI risk quarter to
+`r acute_cough_absolute %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(estimate)`\%
+(`r acute_cough_absolute %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(lower)`\% to
+`r acute_cough_absolute %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(upper)`\%;
+95\% CI),
+`r acute_cough_absolute %>% filter(database == "mdcd", riskStratum == "Q4") %>% pull(estimate)`\%
+(`r acute_cough_absolute %>% filter(database == "mdcd", riskStratum == "Q4") %>% pull(lower)`\% to
+`r acute_cough_absolute %>% filter(database == "mdcd", riskStratum == "Q4") %>% pull(upper)`\%;
+95\% CI), and
+`r acute_cough_absolute %>% filter(database == "mdcr", riskStratum == "Q4") %>% pull(estimate)`\%
+(`r acute_cough_absolute %>% filter(database == "mdcr", riskStratum == "Q4") %>% pull(lower)`\% to
+`r acute_cough_absolute %>% filter(database == "mdcr", riskStratum == "Q4") %>% pull(upper)`\%;
+95\% CI) in the highest acute MI risk quarter in CCAE, MDCD, and MDCR,
+respectively. In terms of angioedema absolute risk differences were very small
+due to the rarity of the outcome.
 
-```{r figure12, cache=TRUE, echo=FALSE, fig.cap="Overview of heterogeneity of ACE-inhibitors treatment on the absolute scale within strata of predicted risk of acute MI.
-Estimates of absolute treatment effect are derived as the difference in Kaplan-Meier estimates at 730 after inclusion.
-Values above 0 favor ACE inhibitors, while values below 0 favor beta blockers.", out.width = '100%'}
-grid::grid.raster(tiff::readTIFF(here("figures/CombinedAbsolute.tiff")))
-```
+The results of all the analyses performed can be accessed and assessed through a
+publicly available web application (https://data.ohdsi.org/AceBeta9Outcomes).
 
-These results suggest that treatment with ACE-inhibitors, compared to treatment
-with beta blockers, may be focused on the higher risk patients, in whom the
-benefits outweigh the harms, while beta blockers may be a viable option in lower
-risk patients, in whom the benefit-harm tradeoff is more favorable. This is in
-accordance with earlier findings that beta blockers should also be considered as
-first-line treatment for younger hypertensive patients [@Cruickshank2007;
-@Khan2006]. This analysis, however, was carried out as demonstration of the
-framework and more rigorous analyses are required to make any suggestions for
-clinical practice.
+## Evaluation
 
-The results of the analyses performed can be accessed and assessed through a
-publicly available web application (https://data.ohdsi.org/AceBeta9Outcomes).
+Meta-analyses focusing on the age of the patients have found that beta blockers
+could only be considered as a first-line treatment for hypertension in younger
+patients, in whom risk of cardiovascular events is lower. The results of our
+case study are in line with this evidence, suggesting that treatment with
+ACE-inhibitors, compared to treatment with beta blockers, may be focused on the
+higher risk patients, in whom the benefits outweigh the harms, while beta
+blockers may be a viable option in lower risk patients, in whom the benefit-harm
+tradeoff is more favorable. Our case study however, was carried out as
+demonstration of the framework and more rigorous analyses are required to make
+any suggestions for clinical practice.
 
@@
 
-We developed a framework for the assessment of heterogeneity of treatment effect
-in large observational databases using a risk modeling approach. The framework
-is implemented in an open source R-package in the OHDSI methods library
-(<https://github.com/OHDSI/RiskStratifiedEstimation>). As a proof-of-concept, we
-used our framework to evaluate heterogeneity of the effect of treatment with
-ACE-inhibitors compared to beta blockers on 3 efficacy and 6 safety outcomes.
+The major contribution of our work is the development of a risk-based framework
+for the assessment of treatment effect heterogeneity in large observational
+databases. This fills a gap identified in the literature after the development
+of guidelines for performing such analyses in the RCT setting
+[@Kent2019;@PathEnE]. As an additional contribution we developed the software
+for implementing this framework in practice and made it publicly available. We
+made our software compatible to databases mapped to OMOP-CDM which allows
+researchers to easily implement our framework in a global network of healthcare
+databases. In our case study we demonstrated the use of our framework for the
+evaluation of treatment effect heterogeneity ACE-inhibitors compared to beta
+blockers on 3 efficacy and 6 safety outcomes. Our results were in line with the
+findings of relevant literature focusing on the age of patients treated for
+hypertension. We propose that this framework is implemented any time treatment
+effect estimation in high-dimensional observational data is undertaken.
 
-In recent years several methods for the evaluation of treatment effect
-heterogeneity have been developed in the setting of RCTs [@Rekkas2020].
-However, low power and restricted prior knowledge on the mechanisms of variation
-in treatment effect are often inherent in RCTs, which are usually adequately
+In recent years several methods for the analysis of treatment effect
+heterogeneity have been developed in the setting of RCTs [@Rekkas2020]. However,
+low power and restricted prior knowledge on the mechanisms of variation in
+treatment effect are often inherent in RCTs, which are usually adequately
 powered only for the analysis of the primary outcome. Observational databases
 contain a large amount of information on treatment assignment and outcomes of
-interest, while also capturing key patient characteristics. Our framework
-provides a standardized approach that can be used to leverage available
-information from these data sources, allowing for large-scale risk-based
-assessment of treatment effect heterogeneity. It is an addition to the rapidly
-expanding literature of approaches for evaluating treatment effect
-heterogeneity. Multiple outcomes can be evaluated in patient subgroups of
-similar baseline outcome risk. Multiple outcome risk stratification schemes can
-also be considered. However, this should be done with caution, as it may hinder
-the interpretability of the results, in a similar manner as typical subgroup
-analyses.
+interest, while also capturing key patient characteristics. They contain readily
+available data on patient subpopulations of interest for which no RCT has
+focused before either due to logistical or ethical reasons. However,
+observational databases are usually not built with research in mind and
+therefore are susceptible to biases, poorly measured outcomes and
+missingness. All these can both obscure true treatment effect heterogeneity or
+falsely introduce it when there is none [@Varadhan2013]. Therefore, inferences
+on both overall treatment effect estimates and treatment effect heterogeneity
+need to rely on strong, often unverifiable, assumptions, despite the
+advancements and guidance on best practices. However, it has been shown that
+well-designed observational studies on average replicate RCT results, even
+though often differences in magnitude may occur
+[@Concato2000;@Ioannidis2001;@Dahabreh2012@Anglemyer2014;@Franklin2017]. Our
+framework is in line with the recently suggested paradigm of high-throughput
+observational studies using consistent and standardized methods for improving
+reproducibility in observational research [@Schuemie2018].
 
-Recently, guidelines on the application of risk modeling approaches for the
-assessment of heterogeneity of treatment effect in RCT settings have been
-proposed [@Kent2019; @PathEnE]. Our framework aims to translate these
-guidelines to the observational setting while also providing a toolkit for its
-implementation within OHDSI. It encourages open science as it requires accurate
-definition of the research questions translated into clear and reproducible
-cohort definitions that can easily be shared among researchers. Researchers with
-access to different databases mapped to OMOP-CDM can also very easily extend
-their overall analyses with risk-based assessment of treatment effect
-heterogeneity. This enables collaboration among multiple sites with access to
-different patient populations. We propose that the framework is implemented any
-time treatment effect estimation in high-dimensional observational data is
-undertaken.
-
-Several considerations need to be made. First, estimates may be
-biased due to the observational nature of the data. We attempt to account for
-potential confounding by estimating propensity scores within strata of predicted
-risk. These scores are estimated using regularized logistic regression on a
-large set of pre-defined covariates. However, such approaches do not account for
-unobserved confounding [@Liu2013]. Several sensitivity analyses have been
-proposed in the literature for measuring the robustness of results in the
-presence of unobserved confounding. Another approach is to calibrate estimates
-and confidence intervals based on a large set of negative controls
-[@Schuemie2014; @Schuemie2571]. Negative controls are treatment-outcome pairs
-for which a null effect has been established. Estimating these effects within
+We attempt to account for potential confounding by estimating propensity scores
+within strata of predicted risk. These scores are estimated using regularized
+logistic regression on a large set of pre-defined covariates. However, such
+approaches do not account for unobserved confounding. Several sensitivity
+analyses have been proposed in the literature for measuring the robustness of
+results in the presence of unobserved confounding
+[@Liu2013;@Cinelli2019]. Another approach is to calibrate estimates and
+confidence intervals based on a large set of negative controls
+[@Schuemie2014;@Schuemie2571]. Negative controls are treatment-outcome pairs for
+which a null effect has been established. Estimating these effects within
 available data provides an approximation of the null distribution that can be
@@
 
-Our method provides a risk-stratified assessment of treatment effect
-heterogeneity. However, even though stratification can provide a useful overview
-for clinical interpretation, these results cannot be applied to individuals in a
-straightforward manner, as we are still estimating subgroup effects
-[@Kent2019]. Presentation of treatment effects as a continuous function of
-risk would be more helpful, but is methodologically challenging. Future research
-is necessary for the development of methods for continuous risk-based assessment
-of HTE.
+Ideally, externally derived and adequately validated prediction models would be
+preferred for analyzing treatment effect heterogeneity. In the absence of such
+prediction models an internally-developed risk prediction model can be
+considered. Earlier simulations of RCT studies have shown that internal models
+developed on the combined treatment and control arms blinded to treatment gave
+relatively unbiased estimates of treatment effect across the spectrum of risk
+[@Burke2014]. However, in observational databases treatment arms may significantly
+differ in sample size. Because the prediction model will possibly fit better to
+the larger treatment arm, spurious treatment-covariate interactions may be
+introduced in the prediction model, leading to sub-optimal risk
+stratification. As a remedy, we first match the patients in the treatment and
+the comparator cohorts on the basis of propensity scores. Additionally, we
+propose to assess model performance in the separate treatment arms to evaluate
+its aptness for risk stratification. Other methods for achieving covariate
+balance before fitting the prediction model can be considered. For example,
+weighting on the propensity scores would be another valid approach.
 
-Ideally, externally derived and adequately validated prediction model would be
-preferred for analyzing treatment effect heterogeneity [@Kent2010]. In the
-absence of such prediction models an internally-developed risk prediction model
-can be considered. Earlier simulations of RCT studies have shown that internal
-models developed on the combined treatment and control arms blinded to treatment
-gave relatively unbiased estimates of treatment effect across the spectrum of
-risk [@Burke2014]. However, in observational databases treatment arms may
-significantly differ in sample size. Because the prediction model will possibly
-better fit to the larger treatment arm, spurious treatment-covariate
-interactions may be introduced in the prediction model, leading to sub-optimal
-risk stratification. As a remedy, we first match the patients in the treatment
-and the comparator cohorts on the basis of propensity scores. Additionally, we
-propose to assess model performance in the separate treatment arms to evaluate
-its aptness for risk stratification.
-
-Recently, disease risk scores have been explored as an alternative to propensity
-scores for balancing covariates [@Glynn2012; @Hansen2008]. In our method, the
-objective of risk stratification is not balancing, but assessing the variation
-of treatment effects on multiple outcomes across patients with different levels
-of baseline risk. Although using the same risk model for balancing and
-risk-based HTE analysis may sound attractive, we note that our method only uses
-one risk model for stratification and one propensity score model for balancing,
-while separate disease risk score models would be required to analyze treatment
+Disease risk scores have been explored as an alternative to propensity scores
+for balancing covariates [@Glynn2012;@Hansen2008]. In our method, the objective
+of risk stratification is not balancing, but assessing the variation of
+treatment effects on multiple outcomes across patients with different levels of
+baseline risk. Although using the same risk model for balancing and risk-based
+HTE analysis may sound attractive, we note that our method only uses one risk
+model for stratification and one propensity score model for balancing, while
+separate disease risk score models would be required to analyze treatment
 effects for each of the multiple outcomes.
-
-In conclusion, the proof-of-concept study demonstrates the feasibility of our
-framework for risk-based assessment of treatment effect heterogeneity in large
-observational data. The standardized framework is easily applicable and highly
-informative whenever treatment effect estimation in high-dimensional
-observational data is of interest. Our framework is a supplement to the
-population-level effect estimation framework developed within OHDSI and, in the
-presence of an adequately discriminating prediction model, can be used to make
-the overall results more actionable for medical decision making.
 

### 2022-05-04 code/CombinedPlots.R

@@ -3,16 +3,16 @@
 library(tidyverse)
 
 mappedOverallRelativeResults <- readRDS(
-  "data/processed/mappedOverallRelativeResults.rds"
+  "data/processed/multipleRseeAnalyses/mappedOverallRelativeResults.rds"
 )
 mappedOverallAbsoluteResults <- readRDS(
-  "data/processed/mappedOverallAbsoluteResults.rds"
+  "data/processed/multipleRseeAnalyses/mappedOverallAbsoluteResults.rds"
 )
 map_outcomes <- readRDS(
-  "data/processed/map_outcomes.rds"
+  "data/processed/multipleRseeAnalyses/map_outcomes.rds"
 )
 map_exposures <- readRDS(
-  "data/processed/map_exposures.rds"
+  "data/processed/multipleRseeAnalyses/map_exposures.rds"
 )
 
 prepareDataset <- function(data) {
@@ -52,63 +52,120 @@ prepareDataset <- function(data) {
 
 relative <- prepareDataset(mappedOverallRelativeResults) %>%
   filter(
-    stratOutcome == "Acute myocardial infarction"
+    stratOutcome == "acute_myocardial_infarction"
   ) %>%
   mutate(
-    database = toupper(database)
+    database = toupper(database),
+    estOutcome = stringr::str_replace_all(estOutcome, "_", " "),
+    stratOutcome = stringr::str_replace_all(stratOutcome, "_", " "),
+    outcomeType = ifelse(
+      estOutcome %in% c(
+        "acute myocardial infarction",
+        "hospitalization with heart failure",
+        "stroke"
+      ),
+      "Main",
+      "Safety"
+    )
+    # estimate = log(estimate),
+    # lower = log(lower),
+    # upper = log(upper)
   )
 
 relativePlot <- ggplot(
   data = relative,
   aes(
-    y = estOutcome,
-    x = estimate,
-    color = database,
-    xmin = lower,
-    xmax = upper
+    x = riskStratum,
+    y = estimate,
+    color = estOutcome,
+    group = estOutcome
   )
 ) +
   geom_point(
-    position = position_dodge(.5),
+    #position = position_dodge(.5),
     size     = 3.5
   ) +
-  geom_errorbar(
-    position = position_dodge(.5),
-    size = 1.1,
-    width = 0
-  ) +
-  facet_grid(~riskStratum) +
-  geom_vline(
[…]

### 2022-05-04 code/Test.R

@@ -1,0 +1,215 @@
+premergeShinyResults <- function(
+  pathList,
+  negativeControlPathList,
+  saveDirectory
+) {
+
+
+  saveDir <- file.path(
+    saveDirectory,
+    "multipleRseeAnalyses"
+  )
+
+  if (!dir.exists(saveDir)) {
+    dir.create(
+      saveDir,
+      recursive = TRUE
+    )
+  }
+
+  createAnlysisPath <- function(
+    path,
+    file
+  ) {
+    res <- file.path(
+      path,
+      file
+    )
+    return(res)
+  }
+
+  mergeFun <- function(
+    analysisDirs,
+    file
+  ) {
+    lapply(
+      analysisDirs,
+      readRDS
+    ) %>%
+      dplyr::bind_rows() %>%
+      saveRDS(
+        file.path(
+          saveDir,
+          paste(
+            file,
+            "rds",
+            sep = "."
+          )
+        )
+      )
+  }
+
+  analysisDirs <- sapply(
+    pathList,
+    createAnlysisPath,
+    file = "analyses.rds"
+  )
+  lapply(
+    analysisDirs,
+    readRDS
+  ) %>%
+    plyr::join_all(
+      type = "full"
+    ) %>%
+    saveRDS(
+      file.path(
+        saveDir,
+        "analyses.rds"
+      )
+    )
+  analysisDirs <- sapply(
+    pathList,
+    createAnlysisPath,
+    file = "map_exposures.rds"
+  )
+  lapply(
+    analysisDirs,
+    readRDS
+  ) %>%
+    plyr::join_all(
+      type = "full"
+    ) %>%
+    saveRDS(
+      file.path(
+        saveDir,
+        "map_exposures.rds"
+      )
+    )
+
+  analysisDirs <- sapply(
+    pathList,
+    createAnlysisPath,
+    file = "map_outcomes.rds"
+  )
+  lapply(
+    analysisDirs,
+    readRDS
+  ) %>%
+    plyr::join_all(
+      type = "full"
+    ) %>%
+    saveRDS(
+      file.path(
+        saveDir,
+        "map_outcomes.rds"
+      )
+    )
+
+  analysisDirs <- sapply(
+    pathList,
+    createAnlysisPath,
+    file = "incidence.rds"
+  )
+
+  mergeFun(
+    analysisDirs = analysisDirs,
+    file = "incidence"
+  )
+
+  analysisDirs <- sapply(
+    pathList,
+    createAnlysisPath,
[…]

### 2022-06-02 submission/manuscript.rmd
prepare manuscript submission

@@
 bibliography: references.bib
-csl: jamia.csl
+csl: the-lancet.csl
 ---
@@
     if (abs(x - round(x)) < .Machine$double.eps) {
-      formatted <- format(x, digits = 1, big.mark = ",")
+      formatted <- format(x, digits = 1, big.mark = " ")
     } else {
-      formatted <- format(x, digits = 1 ,nsmall = 2, big.mark = ",")
+      formatted <- format(x, digits = 1, nsmall = 2, big.mark = " ", decimal.mark = intToUtf8("0x00B7"))
     }
@@
 \thispagestyle{empty}
-\vspace{8mm}
 
@@
 Innovative Medicines Initiative 2 Joint Undertaking (JU) under grant agreement
-No 806968. The JU receives support from the European Union’s Horizon 2020
+No 8069six. The JU receives support from the European Union’s Horizon 2020
 research and innovation programme and EFPIA.
@@
 \singlespacing 
-One of the aims of the Observational Health Data Sciences and Informatics
-(OHDSI) initiative is population-level treatment effect estimation in large
-observational databases. Since treatment effects are well-known to vary across
-groups of patients with different baseline risk, we aimed to extend the OHDSI
-library of open-source tools with a framework for risk-based assessment of
-treatment effect heterogeneity. The proposed framework consists of five
-steps: 1) definition of the problem, i.e. the population, the treatment, the
-comparator and the outcome(s) of interest; 2) identification of relevant
-databases; 3) development of a prediction model for the outcome(s) of
-interest; 4) estimation of relative and absolute treatment effect within strata
-of predicted risk, after adjusting for observed confounding; 5) presentation of
-the results. We demonstrate our framework by evaluating heterogeneity of the
-effect of angiotensin-converting enzyme (ACE) inhibitors versus beta blockers on
-3 efficacy and 6 safety outcomes across three observational databases. Patients
-    at low risk of acute myocardial infarction (MI) received negligible absolute
-benefits for all 3 efficacy outcomes, though they were more pronounced in the
-highest risk quarter, especially for hospitalization with heart failure. The
-substantial risk increase of cough and angioedema with ACE inhibitors across all
-risk strata suggests that beta blockers provide a viable alternative to patients
-at low risk of acute MI. The proof of concept study demonstrates its feasibility
-in large observational data. Further insights may arise by application to safety
-and effectiveness questions across the global data network.
+**Background**: Treatment effects are often anticipated to vary across groups of
+patients with different baseline risk. Implementation of a risk-based approach
+in the RCT setting has shown evidence of increase in statistical power for the
+evaluation of treatment effect heterogeneity. The aim of this study was to
+extend this approach to the observational setting using a standardized scalable
+framework.
+
+**Methods**: The proposed framework consists of
+five steps: 1) definition of the problem, i.e. the population, the treatment,
+the comparator and the outcome(s) of interest; 2) identification of relevant
+databases; 3) development of a prediction model for the outcome(s) of interest;
+4) estimation of relative and absolute treatment effect within strata of
+predicted risk, after adjusting for observed confounding; 5) presentation of the
+results. We demonstrate our framework by evaluating heterogeneity of the effect
+of angiotensin-converting enzyme (ACE) inhibitors versus beta blockers on three
+efficacy and six safety outcomes across three observational databases.
+
+**Findings**: In our demostration, patients at low risk of acute myocardial
+infarction (MI) received negligible absolute benefits for all three efficacy
+outcomes, though they were more pronounced in the highest risk quarter,
+especially for hospitalization with heart failure. However, failing diagnostics
+showed evidence of residual imbalances even after propensity score adjustment.
+
+**Interpretation**: Application of our framework allows for
+observing differential risk of outcomes across strata, which offers the
+opportunity to consider the benefit-harm tradeoff between alternative treatments
+within targeted subpopulations. Further insights may arise by application to
+safety and effectiveness questions on a large scale.
+
+**Funding**: European Health Data and Evidence Network (EHDEN) project.
 
@@
 
+Treatment effects can often vary substantially across individual patients,
+causing overall effect estimates to be inaccurate for a significant proportion
+of the patients at hand [@Rothwell1995;@KRAVITZ2004]. Understanding
+heterogeneity of treatment effects (HTE) has been crucial for both personalized
+(or precision) medicine and comparative effectiveness research, giving rise to a
+wide range of approaches for its discovery, evaluation and application in
+clinical practice. A common approach to evaluating HTE in clinical trials is
+through subgroup analyses, which are rarely adequately powered and can lead to
+false conclusions of absence of HTE or exaggerate its presence
+[@Hayward2006;@Kent2018]. In addition, patients differ with regard to multiple
+characteristics simulatneously, resulting in much richer HTE compared to the one
+explored with regular on-variable-at-a-time subgroup analyses [Kent, BMJ 2018].
+
+Baseline risk is a summary score inherently related to treatment effect that
+can represent more closely the variability in patient characteristics
+[@Rothwell2005;@Hayward2006;@Kent2007;@Kent2008;@Kent2010]. For example, an
+invasive coronary procedure---in comparison with medical treatment---improves
+survival in patients with myocardial infarction at high (predicted) baseline
+risk but not in those at low baseline risk [@Thune2005]. It has also been shown
+that high-risk patients with pre-diabetes benefit substantially more from a
+lifestyle modification program than low-risk patients [@Sussman2015].
+
+Recently, systematic guidance on the application of risk-based methods for the
+assessment of HTE has been developed for RCT data [@Kent2019;@PathEnE]. After
+risk-stratifying patients using an existing or an internally derived prediction
+model, risk stratum-specific estimates of relative and absolute treatment effect
+are evaluated. Several methods for predictive HTE analysis have been adapted for
+use in observational data, but risk-based methods are still not readily
+available and have been highlighted as an important future research need
+[@PathEnE].
+
 The Observational Health Data Science and Informatics (OHDSI) collaborative has
@@
 applying current best practices for comparative effectiveness studies within the
-OHDSI setting has been proposed [@Ryan2013]. This framework was implemented on a
-large scale for estimation of average effects of all first-line hypertension
-treatment classes on a total of 52 outcomes of interest across a global network
-of 9 observational databases [@Suchard2019].
-
-Treatment effects can often vary substantially across individual patients,
-causing overall effect estimates to be inaccurate for a significant proportion
-of the patients at hand [@Rothwell1995;@KRAVITZ2004]. Understanding
-heterogeneity of treatment effects (HTE) has been central to the agenda for both
-personalized (or precision) medicine and comparative effectiveness research,
-giving rise to a wide range of approaches for its discovery, evaluation and
-application in clinical practice. While exploratory and confirmatory HTE
-analyses focus on generating and testing hypotheses on subgroup effects,
-predictive HTE analyses use all available patient data to predict the benefits
-or harms of treatment in individual patients [@Varadhan2013;@Kent2018].
-
-Baseline risk has been recognized as a robust predictor of treatment effect,
-because of the inherent relationship between baseline risk and the absolute
-effects of treatment
-[@Rothwell2005;@Hayward2006;@Kent2007;@Kent2008;@Kent2010]. For example, an
-invasive coronary procedure – in comparison with medical treatment – improves
-survival in patients with myocardial infarction at high (predicted) baseline
-risk but not in those at low baseline risk [@Thune2005]. It has also been shown
-that high-risk patients with pre-diabetes benefit substantially more from a
-lifestyle modification program than low-risk patients [@Sussman2015].
-
-Recently, systematic guidance on the application of risk-based methods has been
-developed for RCT data [@Kent2019;@PathEnE]. Patients are divided into risk strata using
-either an existing or an internally developed risk prediction
-model. Risk-stratum-specific estimates provide an overview of the evolution of
-treatment effects with increasing risk both on the relative and the absolute
-scale. Several methods for predictive HTE analysis have been adapted for use in
-observational data, but risk-based methods for predictive HTE in observational
-data are still lacking and have been highlighted as an important future research
-need [@PathEnE].
+OHDSI setting has been proposed [@Ryan2013]. This framework was successfully
+implemented on a large scale for estimation of average effects of all first-line
+hypertension treatment classes on a total of 52 outcomes of interest across a
+global network of nine observational databases [@Suchard2019].
 
@@
 a case study we analyzed heterogeneity of the effects of first-line hypertension
-treatment: we compared the effect of angiotensin converting enzyme (ACE)
-inhibitors to beta blockers on 9 outcomes across three different US claims
-databases.
+treatment.
 
@@
 
-The proposed framework defines 5 distinct steps that enable a standardized
-approach for risk-based assessment of treatment effect heterogeneity for
-databases mapped to the OMOP-CDM. These are: 1) general definition of the
-research aim; 2) identification of the databases within which the analyses will
-be performed; 3) a prediction step where internal or external prediction models
-are used to assign patient-level risk predictions; 4) an estimation step where
-absolute and relative treatment effects are estimated within risk strata; 5)
-presentation of the results. We developed an open-source R-package for the
-implementation of the proposed framework and made it publicly available. The
-source code can be found at https://github.com/OHDSI/RiskStratifiedEstimation.
+The proposed framework defines 5 distinct steps: 1) definition of the research
+aim; 2) identification of the databases within which the analyses will be
+performed; 3) prediction of outcomes of interest; 4) estimation of absolute and
+relative treatment effects within risk strata; 5) presentation of the
+results. We developed an open-source R-package for the implementation of the
+proposed framework and made it publicly available
+(https://github.com/OHDSI/RiskStratifiedEstimation). An overview of the entire
+framework can be found in Figure \ref{fig:graphicalAbstract}.
 
@@
 
-Our framework uses a comparative cohort design. This means that at least 3
-cohorts of patients need to be defined at this stage of the framework:
+We use a comparative cohort design. This means that at least three cohorts of
+patients need to be defined at this stage of the framework:
 
@@
 to ensure precise effect estimation, even within smaller risk strata. Other
-issues that may be of importance for database inclusion are the depth of data
-capture (the precision at which measurements, lab tests, conditions are
-recorded) and the reliability of data entry.
+relevant issues such as the depth of data capture (the precision at which
+measurements, lab tests, conditions are recorded) and the reliability of data
+entry should also be considered.
 
@@
 
-We adopt the standardized framework for the generation of patient-level
-prediction models using observational data that ensures adherence to existing
-guidelines [@Collins2015;@Moons2015]. To generate the target cohort (i.e. the
-set of patients on which the prediction model will be developed), we pool the
-already defined treatment cohort and comparator cohort. To avoid deferentially
-fitting the prediction model to patients across treatment arms, thus introducing
-spurious interactions with treatment [@Burke2014;@vanKlaveren2019], we develop
-the patient-level prediction model in the propensity score-matched (1:1) subset
-of the sample. Finally, we need to define the time horizon within which we aim
-to make predictions and we also need to select the machine-learning algorithm we
-want to use to generate patient-level predictions.
+Our method relies on adequately separating patients into subgroups based on
+their baseline risk for the outcomes of interest. Therefore, a model---either an
+existing external model adequately validated on an internally developed
+one---assigning patient-level risk is required. For internally developing a risk
+prediction model we adopt the standardized framework focused on observational
+data that ensures adherence to existing guidelines [@Reps2018; @Collins2015;
+@Moons2015].
 
-After model development, a performance overview of the derived prediction models
-including discrimination and calibration both in the propensity score matched
-subset, the entire sample and separately for treated and comparator patients
-should also be reported.
+We first need to define a target cohort of patients, i.e. the set of patients on
+whom the prediction model will be developed. In our case, the target cohort is
+generated by pooling the already defined treatment and comparator cohorts. We
+develop the prediction model on the propensity score-matched (1:1) subset of the
+pooled sample to avoid differentially fitting between treatment arms, thus
+introducing spurious interactions with treatment
+[@Burke2014;@vanKlaveren2019]. We also need to define a set of patients that
+experience the outcome of interest, i.e. the outcome cohort. Finally, we need to
+decide the time frame within which the predictions will be carried out, i.e. the
+patients' time at risk. Subsequently, we can develop the prediction model.
+
+It is important that the prediction models display good discriminative ability
+to ensure that risk-based subgroups are accurately defined. A performance
+overview of the derived prediction models including discrimination and
+calibration both in the propensity score matched subset, the entire sample and
+separately for treated and comparator patients should also be reported.
 
@@
 We estimate treatment effects (both on the relative and the absolute scale)
-within risk strata—typically 4 risk quarters—defined using the prediction model
-of step 3. Effect estimation may be focused on the difference in outcomes for a
+within risk strata defined using the prediction model of step 3. We often
+consider four risk strata, but fewer or more strata can be considered depending on
+the available power for accurately estimating stratum-specific treatment
+effects. Effect estimation may be focused on the difference in outcomes for a
 randomly selected person from the risk stratum (average treatment effect) or for
@@
 
-Prior to analyzing the results, we need to assess if adequate covariate balance
-was achieved within each risk stratum accounting for measured
-confounding. Common approaches include analysis of the overlap of propensity
-score distributions and calculation of standardized mean differences of the
-covariates before and after propensity score adjustment.
-
-A schematic overview of the prediction and estimation steps is presented in
-Figure XX.
+Prior to analyzing results, it is crucial to ensure that all diagnostics are passed
+in all risk strata. The standard diagnostics we carry out include analysis of
+the overlap of propensity score distributions and calculation of standardized
+mean differences of the covariates before and after propensity score
+adjustment. Finally, we use effect estimates for a large set of negative control
+outcomes (i.e. outcomes known to not be related with any of the exposures under
+study) to evaluate the presence of residual confounding not accounted for by
+propensity score adjustment [@Schuemie2014;@Schuemie2571;@Schuemie2018].
 
@@
 heterogeneity, results should be assessed both on the relative and the absolute
-scale. We find that a side-by-side presentation on a forest-like format can give
-a very good representation of our results.
+scale.
 
@@
 
-As a demonstration of our framework, we evaluated if our proposed method was
-able to identify treatment effect heterogeneity of ACE inhibitors compared to
-beta blockers using acute myocardial infarction (MI) risk quarter specific
-effect estimates, both on the relative and on the absolute scale. We focused on
-3 efficacy outcomes (acute MI, hospitalization with heart failure and ischemic
-or hemorrhagic stroke) and 6 safety outcomes (hypokalemia, hyperkalemia,
-hypotension, angioedema, cough and abnormal weight gain). We used data from 3
-US-based claims databases. The analysis plan was the framework outlined in
-steps 1 through 5.
-
-A recent meta-analysis has shown that beta-blockers are on average less
-efficacious in preventing cardiovascular events than ACE inhibitors
-[@Wiysonge2017]. However, meta-analyses focusing on age of the patients have
-shown that beta blockers have similar efficacy in younger hypertensive patients,
-but should not be used as first-line treatment of hypertension in older patients
-mainly due to their increased risk of cardiovascular events compared to other
-first-line antihypertensive treatments [@Khan2006;@Vgele2017].Our aim was to
-replicate these conclusions in our analyses.
+As a demonstration, we evaluated if our proposed method was able to identify
+treatment effect heterogeneity of ACE inhibitors compared to beta blockers using
+acute myocardial infarction (MI) risk quarter specific effect estimates, both on
+the relative and on the absolute scale. We focused on three efficacy outcomes
+(acute MI, hospitalization with heart failure and ischemic or hemorrhagic
+stroke) and six safety outcomes (hypokalemia, hyperkalemia, hypotension,
+angioedema, cough and abnormal weight gain). We used data from three US-based
+claims databases. The analysis plan was the framework outlined in steps 1
+through 5.
 
@@
 
-We considered the following research aim: “compare the effect of ACE-inhibitors
+We considered the following research aim: “compare the effect  of ACE inhibitors
 ($T$) to the effect of beta blockers ($C$) in patients with established hypertension
-with respect to 9 outcomes ($O_1,\dots,O_9$)”. The cohorts are:
+with respect to nine outcomes ($O_1,\dots,O_9$)”. The cohorts are:
 
-- Treatment cohort: Patients receiving any drug within the ACE-inhibitor class
+- Treatment cohort: Patients receiving any drug within the ACE inhibitor class
   with at least one year of follow-up before treatment initiation and a recorded
@@
   hypertension diagnosis within that year.
-- Outcome cohorts: We considered 3 efficacy and 6 safety outcome cohorts. These
+- Outcome cohorts: We considered three efficacy and six safety outcome cohorts. These
   were patients in the database with a diagnosis of: acute MI; hospitalization
@@
   hypokalemia; hyperkalemia; hypotension; angioedema; cough; abnormal weight
-  gain (safety outcomes).
+  gain (safety outcomes). Among the safety outcomes we focus on angioedema and
+  cough which are two known adverse events linked to treatment with ACE
+  inhibitors [@Israili1992]. Results on the rest of the safety outcomes
+  are included in the supplement.
 
@@
 
-For our analyses we used data from 3 US claims databases, namely IBM MarketScan
-Commercial Claims and Encounters (CCAE), IBM MarketScan Medicaid (MDCD), and IBM
-MarketScan Medicare Supplemental Beneficiaries (MDCR). Our analyses included a
-total of 784,561, 66,820 and 101,661 patients initiating treatment with ACE
-inhibitors and 395,740, 45,999 and 69,798 patients initiating treatment with
-beta blockers in CCAE, MDCD and MDCR respectively (Table 1). Adequate numbers of
-patients were included in all strata of predicted acute MI risk.
+```{r, echo = FALSE}
+incidenceOverall <- readRDS(here::here("data/processed/incidenceOverall.rds"))
+patientNumbers <- incidenceOverall %>%
+  group_by(database) %>%
+  summarise(
+    treatment = max(treatmentPersons),
+    comparator = max(comparatorPersons)
+  )
+```
+
+For our demonstration we used data from three US claims databases, namely IBM
+MarketScan Commercial Claims and Encounters (CCAE), IBM MarketScan Medicaid
+(MDCD), and IBM MarketScan Medicare Supplemental Beneficiaries (MDCR). Our
+analyses included a total of
+`r patientNumbers %>% filter(database == "ccae") %>% pull(treatment)`,
+`r patientNumbers %>% filter(database == "mdcd") %>% pull(treatment)`, and
+`r patientNumbers %>% filter(database == "mdcr") %>% pull(treatment)`
+patients initiating treatment with ACE inhibitors and 
+`r patientNumbers %>% filter(database == "ccae") %>% pull(comparator)`,
+`r patientNumbers %>% filter(database == "mdcd") %>% pull(comparator)`, and
+`r patientNumbers %>% filter(database == "mdcr") %>% pull(comparator)`
+patients initiating treatment with beta blockers in CCAE, MDCD and MDCR
+respectively (Table \@ref(tab:incidenceOverall)). Adequate numbers of patients were included in all strata
+of predicted acute MI risk (Supplement: Table XX).
 
@@
 
-We internally developed separate prediction models for acute MI in all 3
+We internally developed separate prediction models for acute MI in all three
 databases. The prediction models were estimated on the propensity score matched
-(1:1) subset of the sample, using caliper of 0.2 and after excluding patients
-having the outcome any time prior to treatment initiation. We chose a time
-horizon of 2 years after inclusion into the target cohort. We developed the
+(1:1) subset of the sample, using caliper of 
+`r format(.2, nsmall = 1, decimal.mark = intToUtf8("0x00B7"))`
+and after excluding patients having the outcome any time prior to treatment
+initiation. We chose a 2-year time at risk for patients and developed the
 prediction models using LASSO logistic regression with 3-fold cross validation
@@
 
-The models developed in the 3 databases had moderate discriminative performance
-(internally validated) with no major issues of overfitting to any cohort except
-for the case of CCAE database in which the derived prediction model performed
-better in the comparator cohort (Table 2). We also observed lower performance of
-the prediction model developed in MDCR compared to the other 2 databases (see
-supplementary material).
+The models had moderate discriminative performance (internally validated) with
+no major issues of overfitting to any cohort except for the case of CCAE, where
+the derived prediction model performed better in the comparator cohort (Table
+\@ref(tab:predictionPerformance)). We also observed lower performance of the
+prediction model developed in MDCR compared to the other 2 databases.
 
@@
 
-Our aim was to estimate the average treatment effects on the relative and the
-absolute scale within strata of predicted acute MI risk.
-
-We used patient-level predictions to stratify the sample into 4 MI risk
+We used patient-level predictions to stratify the sample into four acute MI risk
 quarters. Within risk quarters, relative effects were estimated using Cox
-regression and absolute effects were estimated from the Kaplan-Meier estimate
-differences at 2 years after treatment initiation. To adjust for observed
-confounding within risk strata, we estimated propensity scores using the same
-approach as in the development of prediction models. We used the estimated
-propensity scores to stratify patients into 5 strata, within each risk
-quarter. The risk quarter-specific effect estimates were derived by averaging
+regression and absolute effects were derived from the Kaplan-Meier estimate
+differences at two years after treatment initiation. To adjust for observed
+confounding within each risk quarter, we estimated propensity scores using the
+same approach as step 3 and stratified patients into five propensity score
+strata. The risk quarter-specific effect estimates were derived by averaging
 over the estimates within the propensity score fifths.
 
-There was sufficient overlap of propensity score distribution in all risk strata
-(Figure 2). Propensity score adjustment achieved balance for most of the
-considered covariates, measured using standardized mean differences before and
-after adjustment (Figure 3). However, in lower risk strata imbalances persisted
-for a substantial subset of the covariates, all related to pregnancy
-findings. This was anticipated (use of ACE inhibitors is specifically
-contraindicated during pregnancy) and was already pointed out in
-[@Suchard2019]. Therefore, treatment effect estimates in the lowest risk
-quarters of CCAE and MDCR may still be subject to residual confounding and
-should be interpreted with care.
+In the lowest acute MI risk quarter of CCAE and MDCD we observed strong
+separation of the propensity score distributions, therefore, effect estimates
+derived in these strata are not well-supported (Figure
+\ref{fig:psDensity}). This problematic behavior is also visible in the covariate
+balance plots comparing standardized mean differences of patient characteristics
+before and afrer PS adjustment, where in many cases the commonly accepted bound
+of 
+`r format(.1, nsmall = 1, decimal.mark = intToUtf8("0x00B7"))`
+is violated (Figure \ref{fig:covariateBalance}). This is more pronounced in the
+lowest acute MI risk quarter of CCAE, but remains an issue for a small number of
+covariates in all CCAE risk strata. This diagnostic also fails for the two lower
+acute MI risk quarters of MDCD. Often the persisting imbalances were linked to
+pregnancy outcomes, which can be explained by the contraindication of ACE
+inhibitors in this condition. Analyses in MDCR passed all diagnostics.
+
+Finally, the distribution of the estimated relative risks with regard to 30
+negative control outcomes indicated unresolved confounding within the lowest
+acute MI risk quarter of CCAE (Figrue \ref{fig:negativeControls}). Hazard ratios
+significantly different than 1 (true effect size) were concentrated in the lower
+right part of Figure \ref{fig:negativeControls}: panel Q1. This suggests
+significant negative effects of ACE inhibitors compared to beta blockers on
+causally unrelated outcomes, pointing at unresolved differences between the two
+treatment arms. This was not the case in the other risk quarters of CCAE, or in
+any risk quarter of MDCD and MDCR (Supplement, Figure XX).
 
@@
 ```{r, echo=FALSE}
-map_outcomes <- readRDS(here("data/processed/multipleRseeAnalyses/map_outcomes.rds"))
-map_exposures <- readRDS(here("data/processed/multipleRseeAnalyses/map_exposures.rds"))
-relative <- readRDS(here("data/processed/multipleRseeAnalyses/mappedOverallRelativeResults.rds")) %>%
+map_outcomes <- readRDS(here("data/processed/map_outcomes.rds"))
+
+map_exposures <- readRDS(here("data/processed/map_exposures.rds"))
+overall <- readRDS(here("data/processed/mappedOverallResults.rds")) %>%
+  left_join(map_outcomes, by = c("outcomeId" = "outcome_id")) %>%
+  select(-outcomeId) %>%
+  rename("outcome" = "outcome_name") %>%
+  left_join(map_exposures, by = c("treatmentId" = "exposure_id")) %>%
+  select(-treatmentId) %>%
+  rename("treatment" = "exposure_name") %>%
+  left_join(map_exposures, by = c("comparatorId" = "exposure_id")) %>%
+  select(-comparatorId) %>%
+  rename("comparator" = "exposure_name") %>%
+  mutate(
+    outcome = gsub("_", " ", outcome),
+    estimate = paste0(
+      format(round(estimate, 2), nsmall = 2, decimal.mark = intToUtf8("0x00B7")),
+      " (", 
+      format(round(lower, 2), nsmall = 2, decimal.mark = intToUtf8("0x00B7")),
+      ", ", 
+      format(round(upper, 2), nsmall = 2, decimal.mark = intToUtf8("0x00B7")),
+      ")"
+    )
+  ) %>%
+  select(-c("treatment", "comparator", "analysisType", "seLogRr", "lower", "upper")) %>%
+  relocate(c("database", "outcome"))
+
+relative <- readRDS(here("data/processed/mappedOverallRelativeResults.rds")) %>%
   left_join(map_outcomes, by = c("estOutcome" = "outcome_id")) %>%
@@
 
-absolute <- readRDS(here("data/processed/multipleRseeAnalyses/mappedOverallAbsoluteResults.rds")) %>%
+absolute <- readRDS(here("data/processed/mappedOverallAbsoluteResults.rds")) %>%
   left_join(map_outcomes, by = c("estOutcome" = "outcome_id")) %>%
@@
   )
+
+acute_mi_stroke_relative <- relative %>%
+  filter(
+    stratOutcome == "acute_myocardial_infarction",
+    estOutcome == "stroke"
+  )
+
+acute_mi_stroke_absolute <- absolute %>%
+  filter(
+    stratOutcome == "acute_myocardial_infarction",
+    estOutcome == "stroke"
+  )
 ```
 
-For hospitalization with acute MI there was an increasing trend in favor ACE
-inhibitors compared to beta blockers on the relative scale (hazard ratios
-decreased) with increasing acute MI risk. More specifically, hazard ratios
-decreased from 
+The overall estimated hazard ratios for the main outcomes are presented in
+Table \@ref(tab:overallTable). For hospitalization with acute MI there was an increasing trend in
+favor ACE inhibitors compared to beta blockers on the relative scale (hazard
+ratios decreased) with increasing acute MI risk. More specifically, hazard
+ratios decreased from
 `r acute_mi_relative %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(estimate)`
@@
 `r acute_mi_relative %>% filter(database == "mdcd", riskStratum == "Q4") %>% pull(upper)`;
-95\% CI) in CCAE and MDCD respectively (Figure 4). In MDCR hazard ratios remained stable
-from
+95\% CI) in CCAE and MDCD respectively (Figure 4). In MDCR hazard ratios
+increased from
 `r acute_mi_relative %>% filter(database == "mdcr", riskStratum == "Q1") %>% pull(estimate)`
@@
 `r acute_mi_relative %>% filter(database == "mdcr", riskStratum == "Q4") %>% pull(upper)`;
-95\% CI). In terms of hospitalization with heart failure relative treatment
-effect estimates favored ACE inhibitors across all risk strata in all
-databases. We found no differences between the two treatments in their effect on
-stroke on the relative scale. In terms of the safety outcomes, we found an
-increased ACE-inhibitor risk of cough and angioedema on the relative scale
-across all risk strata. In the case of cough, this effect decreased with
-increasing risk of acute MI---from 
+95\% CI). Relative treatment effect estimates for hospitalization with heart
+failure favored ACE inhibitors across all risk strata in all databases. In the
+case of stroke in CCAE we found quite constant hazard ratios which became weaker
+in the highest risk quarter patients
+(`r acute_mi_stroke_relative %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(estimate)`
+with 95\% CI from 
+`r acute_mi_stroke_relative %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(lower)`
+to
+`r acute_mi_stroke_relative %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(upper)`).
+In the other two databases no significant relative treatment effects were
+observed for stroke.  In terms of the safety outcomes, we found an increased ACE inhibitor
+risk of cough and angioedema on the relative scale across all risk strata. In
+the case of cough, this effect decreased with increasing risk of acute MI---from
 `r acute_cough_relative %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(estimate)`
@@
 
-## Evaluation
+## Interpretation
 
-Meta-analyses focusing on the age of the patients have found that beta blockers
-could only be considered as a first-line treatment for hypertension in younger
-patients, in whom risk of cardiovascular events is lower. The results of our
-case study are in line with this evidence, suggesting that treatment with
-ACE-inhibitors, compared to treatment with beta blockers, may be focused on the
-higher risk patients, in whom the benefits outweigh the harms, while beta
-blockers may be a viable option in lower risk patients, in whom the benefit-harm
-tradeoff is more favorable. Our case study however, was carried out as
-demonstration of the framework and more rigorous analyses are required to make
-any suggestions for clinical practice.
+The overall benefits of ACE inhibitors compared to beta blockers for acute MI
+and hospitalization with heart failure are driven mainly by the higher acute MI
+risk patients in CCAE and MDCD, hence the observed increasing patterns of the
+absolute treatment benefits. In MDCR we found no significant overall difference
+on the relative scale for acute MI and, cosequently, no differences in acute MI
+risk strata were observed on any scale. For heart failure, MDCR patients at the
+lower half of acute MI risk had lower absolute benefits compared to the patients
+at the upper half. Finally, the small overall relative effect for stroke
+resulted in smaller absolute benefits of ACE inhibitors across acute MI risk
+strata in all databases.
+
+For patients at lower acute MI risk, the cough and angioedema risk increase
+related to treatment with ACE inhibitors may be important factors to consider
+for medical decision making, given the small benefits observed for the main
+outcomes. However, diagnostics failed in lower risk patients within CCAE and
+MDCR which renders these conclusions less dependable.
+
+Note that any conclusions drawn are for demonstration purposes only and should
+be interpreted under this very limited setting.
 
@@
 databases. In our case study we demonstrated the use of our framework for the
-evaluation of treatment effect heterogeneity ACE-inhibitors compared to beta
-blockers on 3 efficacy and 6 safety outcomes. Our results were in line with the
-findings of relevant literature focusing on the age of patients treated for
-hypertension. We propose that this framework is implemented any time treatment
-effect estimation in high-dimensional observational data is undertaken.
+evaluation of treatment effect heterogeneity ACE inhibitors compared to beta
+blockers on three efficacy and six safety outcomes. We propose that this framework is
+implemented any time treatment effect estimation in high-dimensional
+observational data is undertaken.
 
 In recent years several methods for the analysis of treatment effect
-heterogeneity have been developed in the setting of RCTs [@Rekkas2020]. However,
+heterogeneity have been developed in the RCT setting [@Rekkas2020]. However,
 low power and restricted prior knowledge on the mechanisms of variation in
@@
 focused before either due to logistical or ethical reasons. However,
-observational databases are usually not built with research in mind and
-therefore are susceptible to biases, poorly measured outcomes and
-missingness. All these can both obscure true treatment effect heterogeneity or
-falsely introduce it when there is none [@Varadhan2013]. Therefore, inferences
-on both overall treatment effect estimates and treatment effect heterogeneity
-need to rely on strong, often unverifiable, assumptions, despite the
-advancements and guidance on best practices. However, it has been shown that
-well-designed observational studies on average replicate RCT results, even
-though often differences in magnitude may occur
+observational databases can be susceptible to biases, poorly measured outcomes
+and missingness, which may obscure true HTE or falsely introduce it when there
+is none [@Varadhan2013]. Therefore, inferences on both overall treatment effect
+estimates and HTE need to rely on strong, often unverifiable, assumptions,
+despite the advancements and guidance on best practices. However, well-designed
+observational studies on average replicate RCT results, even though often
+differences in magnitude may occur
 [@Concato2000;@Ioannidis2001;@Dahabreh2012@Anglemyer2014;@Franklin2017]. Our
@@
 
-We attempt to account for potential confounding by estimating propensity scores
-within strata of predicted risk. These scores are estimated using regularized
-logistic regression on a large set of pre-defined covariates. However, such
-approaches do not account for unobserved confounding. Several sensitivity
-analyses have been proposed in the literature for measuring the robustness of
-results in the presence of unobserved confounding
-[@Liu2013;@Cinelli2019]. Another approach is to calibrate estimates and
-confidence intervals based on a large set of negative controls
-[@Schuemie2014;@Schuemie2571]. Negative controls are treatment-outcome pairs for
-which a null effect has been established. Estimating these effects within
-available data provides an approximation of the null distribution that can be
-used to empirically recalibrate effect estimates. Future work may extend our
-framework with this type of analyses.
+Our framework highlights the scale dependency of HTE and how it relates to
+baseline risk. Treatment effect is mathematically determined by baseline
+risk, if we assume a constant non-zero effect size [@Dahabreh2016]. Patients
+with low baseline risk can only experience minimal benefits, before their risk
+is reduced to zero. In contrast, high risk patients are capable of displaying
+much higher absolute benefits. This becomes evident when evaluating the effects
+of ACE inhibitors on cough and angioedema, compared to treatment with beta
+blockers. Despite the small relative cough risk increase of ACE inhibitors, the
+large baseline cough risk resulted in larger absolute risk differences, compared
+to the other considered outcomes. Conversely, in the case of angioedema, the
+substantial relative risk increase with ACE inhibitors only translated in a
+small absolute risk increase due to the quite low baseline angioedema risk.
 
-Ideally, externally derived and adequately validated prediction models would be
-preferred for analyzing treatment effect heterogeneity. In the absence of such
-prediction models an internally-developed risk prediction model can be
-considered. Earlier simulations of RCT studies have shown that internal models
-developed on the combined treatment and control arms blinded to treatment gave
-relatively unbiased estimates of treatment effect across the spectrum of risk
-[@Burke2014]. However, in observational databases treatment arms may significantly
-differ in sample size. Because the prediction model will possibly fit better to
-the larger treatment arm, spurious treatment-covariate interactions may be
-introduced in the prediction model, leading to sub-optimal risk
-stratification. As a remedy, we first match the patients in the treatment and
-the comparator cohorts on the basis of propensity scores. Additionally, we
-propose to assess model performance in the separate treatment arms to evaluate
-its aptness for risk stratification. Other methods for achieving covariate
-balance before fitting the prediction model can be considered. For example,
-weighting on the propensity scores would be another valid approach.
+The application of our framework in the case study is for demonstration purposes
+and there are several limitations to its conclusions. First, death could be a
+competing risk. We could expand our framework in the future to potentially
+support subdistribution hazard ratios and cumulative incidence
+reductions. Second, we only used the databases readily available to us and not
+all the available databases mapped to OMOP-CDM. Therefore, the generalizability
+of our results still needs to be explored in future studies. These studies
+should also address the particular aspects of the databases at hand, such as
+their sampling frame, the completeness of the data they capture and many others
+that were not assessed in our demonstration. Third, we did not to correct for
+multiplicity when presenting the results. We are interested in presenting trends
+in the data and not detecting the specific subgroups within which a non-null
+treatment effect is detected. The implementation of our framework, however,
+generates all the relevant output required for a researcher to correct for
+multiple testing, if that is required.
 
-Disease risk scores have been explored as an alternative to propensity scores
-for balancing covariates [@Glynn2012;@Hansen2008]. In our method, the objective
-of risk stratification is not balancing, but assessing the variation of
-treatment effects on multiple outcomes across patients with different levels of
-baseline risk. Although using the same risk model for balancing and risk-based
-HTE analysis may sound attractive, we note that our method only uses one risk
-model for stratification and one propensity score model for balancing, while
-separate disease risk score models would be required to analyze treatment
-effects for each of the multiple outcomes.
+In conclusion, the case study demonstrates the feasibility of our framework for
+risk-based assessment of treatment effect heterogeneity in large observational
+data. It is easily applicable and highly informative whenever treatment effect
+estimation in high-dimensional observational data is of interest.
 
@@
 \noindent
+
+\newpage
+
+# Tables and figures
+
+```{r incidenceOverall, echo = FALSE}
+incidence <- readRDS(here::here("data/processed/incidenceOverall.rds")) %>%
+  left_join(map_outcomes, by = c("outcomeId" = "outcome_id")) %>%
+  select(-outcomeId) %>%
+  rename("outcomeId" = "outcome_name") %>%
+  mutate(
+    treatmentYears = round(treatmentDays / 365),
+    comparatorYears = round(comparatorDays / 365),
+    outcomeId =  factor(
+    outcomeId,
+    levels = c(
+      "acute_myocardial_infarction",
+      "hospitalization_with_heart_failure",
+      "stroke"
+    ),
+    labels = c(
+        "acute myocardial infarction",
+        "hospitalization with heart failure",
+        "stroke"
+      )
+    )
+  ) %>%
+  arrange(database, outcomeId)
+
+kableExtra::kbl(
+  incidence %>%
+    select(
+      outcomeId,
+      treatmentPersons,
+      treatmentYears,
+      treatmentOutcomes,
+      comparatorPersons,
+      comparatorYears,
+      comparatorPersons,
+      comparatorOutcomes
+    ),
+  align = "llrrrrr",
+  format = "latex",
+  booktabs = TRUE,
+  longtable = TRUE,
+  escape = FALSE,
+  caption = "Number of patients, person years and events within quarters of predicted risk for acute MI for the three efficacy outcomes of the study (acute MI, hospitalization with heart failure and ischemic or hemorrhagic stroke) across the three databases.",
+  linesep = c('', '', '', '\\addlinespace'),
+  col.names = c(
+    "Outcome",
+    "Patients",
+    "Person years",
+    "Outcomes",
+    "Patients",
+    "Person years",
+    "Outcomes"
+  ),
+  format.args = list(
+    big.mar      = " ",
+    decimal.mark = "."
+  )
+) %>%
+  kableExtra::add_header_above(
+    c(
+      " " = 1,
+      "ACE inhibitors" = 3,
+      "Beta blockers" = 3
+    )
+  ) %>%
+  kableExtra::group_rows("CCAE", 1, 3) %>%
+  kableExtra::group_rows("MDCD", 4, 6) %>%
+  kableExtra::group_rows("MDCR", 7, 9) %>%
+  kableExtra::kable_styling(font_size = 7)
+
+```
+
+\newpage
+
+```{r predictionPerformance, echo = FALSE}
+predictionPerformance <- readRDS(here("data/processed/predictionPerformance.rds")) %>%
+  filter(
+    stratOutcome == 2
+  ) %>%
+  select(c("database", "cohort", "AUROC", "95% lower AUROC", "95% upper AUROC")) %>%
+  rename(
+    "lower" = "95% lower AUROC",
+    "upper" = "95% upper AUROC"
+  ) %>%
+  mutate(
+    auc = paste(
+      format(round(AUROC, 2), nsmall = 2, decimal.mark = intToUtf8("0x00B7")),
+      paste0(
+        "(",
+        format(round(lower, 2), nsmall = 2, decimal.mark = intToUtf8("0x00B7")),
+        ", ",
+        format(round(upper, 2), nsmall = 2, decimal.mark = intToUtf8("0x00B7")),
+        ")"
+      )
+    )
+  ) %>%
+  select(-c("AUROC", "lower", "upper")) %>%
+  mutate(
+    cohort = factor(
+      cohort,
+      levels = c(
+        "Matched",
+        "Treatment",
+        "Comparator",
+        "EntirePopulation"
+      ),
+      labels = c(
+        "Matched",
+        "Treatment",
+        "Comparator",
+        "Entire population"
+      )
+    ),
+    database = factor(
+      database,
+      levels = c(
+        "ccae",
+        "mdcd",
+        "mdcr"
+      )
+    )
+  ) %>%
+  arrange(database, cohort)
+
+kableExtra::kbl(
+  predictionPerformance %>% pivot_wider(id_cols = "cohort", names_from = "database", values_from = "auc"),
+  align = "lrrr",
+  format = "latex",
+  booktabs = TRUE,
+  longtable = TRUE,
+  escape = FALSE,
+  caption = "Discriminative ability (c-statistic) of the derived prediction models for acute myocardial infarction in the matched set (development set), the treatment cohort, the comparator cohort and the entire population in CCAE, MDCD and MDCR.",
+  linesep = c('', '', '', '\\addlinespace'),
+  col.names = c(
+    "Population",
+    "CCAE",
+    "MDCD",
+    "MDCR"
+  ),
+  format.args = list(
+    big.mar      = ","
+  )
+) %>%
+  kableExtra::kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r overallTable, echo = FALSE}
+kableExtra::kbl(
+  overall %>% 
+    pivot_wider(id_cols = "outcome", names_from = "database", values_from = "estimate"),
+  format = "latex",
+  align = "lrrr",
+  caption = "Overall hazard ratios.",
+  booktabs = TRUE,
+  col.names = c(
+    "Outcome",
+    "CCAE",
+    "MDCD",
+    "MDCR"
+  ),
+  format.args = list(decimal.mark = intToUtf8("0x00B7"))
+) %>%
+  kableExtra::kable_styling(latex_options = "HOLD_position")
+```
+
+\newpage
+
+```{r graphicalAbstract, cache=TRUE, echo=FALSE, fig.cap="Illustration of how the framework is applied on two observational databases, preferably mapped to OMOP-CDM.", fig.show="hold", out.width = '100%'}
+include_graphics(here("figures/graphicalAbstract.png"))
+```
+
+\newpage
+
+```{r psDensity, cache=TRUE, echo=FALSE, fig.cap="Preference score distributions for the evaluation of heterogeneity of the effect of ACE inhibitors compared to beta blockers on acute MI based on quarters of predicted acute MI risk.
+The preference score is a transformation of the propensity score that adjusts for prevalence differences between populations.", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/PsDensity.tiff")))
+```
+
+\newpage
+
+```{r covariateBalance, cache=TRUE, echo=FALSE, fig.cap="Patient characteristic balance for ACE inhibitors and beta blockers before and after stratification on the propensity scores.
+Each dot represents the standardized difference of means for a single covariate before (x-axis) and after (y-axis) stratification.", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/CovariateBalance.tiff")))
+```
+
+\newpage
+
+```{r negativeControls, cache=TRUE, echo=FALSE, fig.cap=" Systematic error.
+Effect size estimates for the negative controls (true hazard ratio = 1).
+Estimates below the diagonal dashed lines are statistically significant (alpha = 0.05) different from the true effect size.
+A well-calibrated estimator should have the true effect size within the 95 percent confidence interval 95 percent of times.", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/NegativeControlsPlot_ccae.tiff")))
+```
+
+\newpage
+
+```{r relative, cache=TRUE, echo=FALSE, fig.cap="Overview of heterogeneity of ACE inhibitors treatment on the relative scale (hazard ratios) within strata of predicted risk of acute MI.
+Values below 1 favor ACE inhibitors, while values above 1 favor beta blockers.", fig.show="hold", out.width = '100%'}
+    grid::grid.raster(tiff::readTIFF(here("figures/CombinedRelative_acute_myocardial_infarction.tiff")))
+```
+
+\newpage
+
+```{r absolute, cache=TRUE, echo=FALSE, fig.cap="Overview of heterogeneity of ACE inhibitors treatment on the absolute scale within strata of predicted risk of acute MI.
+Estimates of absolute treatment effect are derived as the difference in Kaplan-Meier estimates at two years after inclusion.
+Values above 0 favor ACE inhibitors, while values below 0 favor beta blockers.
+", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/CombinedAbsolute_acute_myocardial_infarction.tiff")))
+```

### 2022-06-02 code/CombinedAbsolutePlots.R

@@ -1,0 +1,189 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(scales)
+
+args <- commandArgs(trailingOnly = TRUE)
+outcome <- as.character(args[1])
+fileType <- as.character(args[2])
+
+mappedOverallAbsoluteResults <- readRDS(
+  "data/processed/mappedOverallAbsoluteResults.rds"
+)
+map_outcomes <- readRDS(
+  "data/processed/map_outcomes.rds"
+)
+map_exposures <- readRDS(
+  "data/processed/map_exposures.rds"
+)
+
+prepareDataset <- function(data) {
+  data %>%
+    left_join(
+      map_outcomes,
+      by = c("stratOutcome" = "outcome_id")
+    ) %>%
+    select(-stratOutcome) %>%
+    rename("stratOutcome" = "outcome_name") %>%
+    left_join(
+      map_outcomes,
+      by = c("estOutcome" = "outcome_id")
+    ) %>%
+    select(-estOutcome) %>%
+    rename("estOutcome" = "outcome_name") %>%
+    left_join(
+      map_exposures,
+      by = c("treatment" = "exposure_id")
+    ) %>%
+    select(-treatment) %>%
+    rename("treatment" = "exposure_name") %>%
+    left_join(
+      map_exposures,
+      by = c("comparator" = "exposure_id")
+    ) %>%
+    select(-comparator) %>%
+    rename("comparator" = "exposure_name") %>%
+    mutate(
+      estOutcome = factor(
+        estOutcome,
+        rev(unique(estOutcome))
+      )
+    )
+}
+
+
+absolute <- prepareDataset(mappedOverallAbsoluteResults) %>%
+  filter(
+    stratOutcome == outcome
+  ) %>%
+  mutate(
+    database = toupper(database),
+    estimate = 100 * estimate,
+    lower = 100 * lower,
+    upper = 100 * upper,
+    estOutcome = stringr::str_replace_all(estOutcome, "_", " "),
+    stratOutcome = stringr::str_replace_all(stratOutcome, "_", " "),
+    estOutcome = stringr::str_replace_all(estOutcome, "hospitalization", "hospitalisation"),
+    estOutcome = stringr::str_replace_all(estOutcome, "emia", "aemia"),
+    stratOutcome = stringr::str_replace_all(stratOutcome, "_", " "),
+    stratOutcome = stringr::str_replace_all(stratOutcome, "hospitalization", "hospitalisation"),
[…]

### 2022-06-02 code/CombinedPlots.R

@@ -1,310 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(tidyverse)
-
-mappedOverallRelativeResults <- readRDS(
-  "data/processed/multipleRseeAnalyses/mappedOverallRelativeResults.rds"
-)
-mappedOverallAbsoluteResults <- readRDS(
-  "data/processed/multipleRseeAnalyses/mappedOverallAbsoluteResults.rds"
-)
-map_outcomes <- readRDS(
-  "data/processed/multipleRseeAnalyses/map_outcomes.rds"
-)
-map_exposures <- readRDS(
-  "data/processed/multipleRseeAnalyses/map_exposures.rds"
-)
-
-prepareDataset <- function(data) {
-  data %>%
-    left_join(
-      map_outcomes,
-      by = c("stratOutcome" = "outcome_id")
-    ) %>%
-    select(-stratOutcome) %>%
-    rename("stratOutcome" = "outcome_name") %>%
-    left_join(
-      map_outcomes,
-      by = c("estOutcome" = "outcome_id")
-    ) %>%
-    select(-estOutcome) %>%
-    rename("estOutcome" = "outcome_name") %>%
-    left_join(
-      map_exposures,
-      by = c("treatment" = "exposure_id")
-    ) %>%
-    select(-treatment) %>%
-    rename("treatment" = "exposure_name") %>%
-    left_join(
-      map_exposures,
-      by = c("comparator" = "exposure_id")
-    ) %>%
-    select(-comparator) %>%
-    rename("comparator" = "exposure_name") %>%
-    mutate(
-      estOutcome = factor(
-        estOutcome,
-        rev(unique(estOutcome))
-      )
-    )
-}
-
-
-relative <- prepareDataset(mappedOverallRelativeResults) %>%
-  filter(
-    stratOutcome == "acute_myocardial_infarction"
-  ) %>%
-  mutate(
-    database = toupper(database),
-    estOutcome = stringr::str_replace_all(estOutcome, "_", " "),
-    stratOutcome = stringr::str_replace_all(stratOutcome, "_", " "),
-    outcomeType = ifelse(
-      estOutcome %in% c(
-        "acute myocardial infarction",
-        "hospitalization with heart failure",
-        "stroke"
-      ),
-      "Main",
-      "Safety"
-    )
-    # estimate = log(estimate),
-    # lower = log(lower),
-    # upper = log(upper)
-  )
-
[…]

### 2022-06-02 code/CombinedRelativePlots.R

@@ -1,0 +1,209 @@
+#!/usr/bin/env Rscript
+
+
+library(tidyverse)
+library(scales)
+
+args <- commandArgs(trailingOnly = TRUE)
+outcome <- as.character(args[1])
+fileType <- as.character(args[2])
+
+mappedOverallRelativeResults <- readRDS(
+  "data/processed/mappedOverallRelativeResults.rds"
+)
+map_outcomes <- readRDS(
+  "data/processed/map_outcomes.rds"
+)
+map_exposures <- readRDS(
+  "data/processed/map_exposures.rds"
+)
+
+prepareDataset <- function(data) {
+  data %>%
+    left_join(
+      map_outcomes,
+      by = c("stratOutcome" = "outcome_id")
+    ) %>%
+    select(-stratOutcome) %>%
+    rename("stratOutcome" = "outcome_name") %>%
+    left_join(
+      map_outcomes,
+      by = c("estOutcome" = "outcome_id")
+    ) %>%
+    select(-estOutcome) %>%
+    rename("estOutcome" = "outcome_name") %>%
+    left_join(
+      map_exposures,
+      by = c("treatment" = "exposure_id")
+    ) %>%
+    select(-treatment) %>%
+    rename("treatment" = "exposure_name") %>%
+    left_join(
+      map_exposures,
+      by = c("comparator" = "exposure_id")
+    ) %>%
+    select(-comparator) %>%
+    rename("comparator" = "exposure_name") %>%
+    mutate(
+      estOutcome = factor(
+        estOutcome,
+        rev(unique(estOutcome))
+      )
+    )
+}
+
+
+relative <- prepareDataset(mappedOverallRelativeResults) %>%
+  filter(
+    stratOutcome == outcome
+  ) %>%
+  mutate(
+    database = toupper(database),
+    estOutcome = stringr::str_replace_all(estOutcome, "_", " "),
+    stratOutcome = stringr::str_replace_all(stratOutcome, "_", " "),
+    estOutcome = stringr::str_replace_all(estOutcome, "hospitalization", "hospitalisation"),
+    estOutcome = stringr::str_replace_all(estOutcome, "emia", "aemia"),
+    stratOutcome = stringr::str_replace_all(stratOutcome, "_", " "),
+    stratOutcome = stringr::str_replace_all(stratOutcome, "hospitalization", "hospitalisation"),
+    stratOutcome = stringr::str_replace_all(stratOutcome, "emia", "aemia"),
[…]

### 2022-06-02 code/CovariateBalancePlot.R

@@ -1,23 +1,31 @@
 #!/usr/bin/env Rscript
 
 library(tidyverse)
+library(scales)
+
+args <- commandArgs(trailingOnly = TRUE)
+fileType <- as.character(args[1])
 
 databases <- c("ccae", "mdcd", "mdcr")
+map_exposures <- readRDS("data/processed/map_exposures.rds")
 pp <- list()
 
 for (i in seq_along(databases)) {
-  pattern <- paste0("^bal.*", databases[i], ".*104_104*.rds")
+  pattern <- paste0("^balance.*", databases[i], ".*2_2.*.rds")
   pp[[i]] <- list.files(
-    path = ".scratch/AceBeta9Outcomes/data",
+    path = "data/processed",
     pattern = pattern,
     full.names = TRUE
   ) %>%
-    map(readRDS) %>%
+    readRDS() %>%
     bind_rows() %>%
     tibble() %>%	
     CohortMethod::plotCovariateBalanceScatterPlot(
       beforeLabel = "",
       afterLabel = toupper(databases[i])
+    ) +
+    scale_x_continuous(
+      labels = comma_format(decimal.mark = intToUtf8("0x00B7"))
     ) +
     ggplot2::facet_grid(
       ~riskStratum
@@ -43,12 +51,32 @@ for (i in seq_along(databases)) {
 }
 
 plot <- gridExtra::grid.arrange(pp[[1]], pp[[2]], pp[[3]], nrow = 3)
-ggsave(
-  "figures/CovariateBalance.tiff",
-  plot, 
-  compression = "lzw", 
-  width       = 600, 
-  height      = 350,
-  units       = "mm",
-  dpi         = 300
-)
+
+
+if (fileType == "tiff") {
+  fileName <- "CovariateBalance.tiff"
+  ggsave(
+    file.path(
+      "figures",
+      fileName
+    ),
+    plot, 
+    compression = "lzw", 
+    width       = 650, 
+    height      = 350,
+    units       = "mm",
+    dpi         = 300
+  )
+} else if (fileType == "svg") {
+  fileName <- "CovariateBalance.svg"
+  ggsave(
+    file.path(
+      "figures",
+      fileName
+    ),
+    plot, 
+    width       = 650, 
+    height      = 350,
+    units       = "mm"
+  )
+}

### 2022-06-02 code/NegativeControlsPlot.R

@@ -1,0 +1,149 @@
+#!/usr/bin/env Rscript
+source("code/helpers/GenerateNcPlot.R")
+library(tidyverse)
+library(grid)
+
+args <- commandArgs(trailingOnly = TRUE)
+db <- as.character(args[1])
+fileType <- as.character(args[2])
+
+negativeControls <- readRDS("data/processed/negativeControls.rds") %>%
+  filter(!is.na(seLogRr))
+
+plotRiskStratifiedNegativeControls <- function(negativeControls) {
+  riskStrata <- unique(negativeControls$riskStratum)
+  plots <- list()
+  for (i in seq_along(riskStrata)) {
+    negativeControlsSubset <- negativeControls %>%
+      filter(riskStratum == riskStrata[i])
+
+    # null <- EmpiricalCalibration::fitNull(
+    #   logRr   = log(negativeControlsSubset$estimate),
+    #   seLogRr = negativeControlsSubset$seLogRr
+    # )
+
+    # plots[[i]] <- EmpiricalCalibration::plotCalibrationEffect(
+    #   logRrNegatives   = log(negativeControlsSubset$estimate),
+    #   seLogRrNegatives = negativeControlsSubset$seLogRr,
+    #   null             = null
+    # ) +
+    #   ggtitle(paste("Q", i)) +
+    #   theme(
+    #     plot.title = element_text(size = 32),
+    #     axis.title = element_blank()
+    #   )
+    plots[[i]] <- generateNcPlot(
+      logRr   = log(negativeControlsSubset$estimate),
+      seLogRr = negativeControlsSubset$seLogRr,
+      trueLogRr        = rep(0, nrow(negativeControlsSubset))
+    ) +
+      ggtitle(paste("Q", i)) +
+      theme(
+        plot.title = element_text(size = 32),
+        axis.title = element_blank()
+      )
+  }
+  gridList <- list(
+    plots[[1]] +
+      theme(
+        axis.title.x = ggplot2::element_blank(),
+        axis.title.y = ggplot2::element_blank(),
+        axis.text.x = element_blank()
+      ),
+    plots[[2]] + 
+      theme(
+        axis.title = element_blank(),
+        axis.text.x = element_blank(),
+        axis.text.y = element_blank()
+      ),
+    plots[[3]],
+    plots[[4]] +
+      theme(
[…]

### 2022-06-02 code/PsDensityPlot.R

@@ -1,23 +1,35 @@
 #!/usr/bin/env Rscript
 
 library(tidyverse)
+library(scales)
+
+
+args <- commandArgs(trailingOnly = TRUE)
+fileType <- as.character(args[1])
 
 databases <- c("ccae", "mdcd", "mdcr")
+map_exposures <- readRDS("data/processed/map_exposures.rds")
 pp <- list()
 
 for (i in seq_along(databases)) {
-  pattern <- paste0("^ps.*", databases[i], ".*104_104.*.rds")
+  pattern <- paste0("^ps.*", databases[i], ".*2_2.*.rds")
   pp[[i]] <- list.files(
-    path = ".scratch/AceBeta9Outcomes/data",
+    path = "data/processed",
     pattern = pattern,
     full.names = TRUE
   ) %>%
-    map(readRDS) %>%
-    bind_rows() %>%
-    mutate(
-      treatment = as.factor(treatment)
+    readRDS() %>%
+    dplyr::mutate(
+      treatment = ifelse(
+        treatment == 1,
+        treatmentId,
+        comparatorId
+      )
     ) %>%
-    tibble() %>%	
+    dplyr::left_join(
+    map_exposures,
+    by = c("treatment" = "exposure_id")
+  ) %>% 
     ggplot2::ggplot(
       ggplot2::aes(
         x = x,
@@ -27,12 +39,15 @@ for (i in seq_along(databases)) {
     ggplot2::geom_density(
       stat = "identity",
       ggplot2::aes(
-        color = treatment,
-        group = treatment,
-        fill = treatment
+        color = exposure_name,
+        group = exposure_name,
+        fill = exposure_name
       )
     ) +
-    scale_x_continuous(breaks = seq(0, 1, .5)) +
+    scale_x_continuous(
+      breaks = seq(0, 1, .5),
+      labels = comma_format(decimal.mark = intToUtf8("0x00B7"))
+    ) +
     ggplot2::facet_grid(~riskStratum) +
     ggplot2::ylab(
       label = toupper(databases[i])
@@ -78,4 +93,33 @@ ggsave(
   units       = "mm",
   dpi         = 300
 )
+
+if (fileType == "tiff") {
+  fileName <- "PsDensity.tiff"
+  ggsave(
+    file.path(
+      "figures",
+      fileName
+    ),
+    plot, 
+    compression = "lzw", 
+    width       = 650, 
+    height      = 350,
+    units       = "mm",
[…]

### 2022-06-02 code/Table1.R

@@ -2,242 +2,73 @@
 
 library(tidyverse)
 
-tableCcae <- tibble(
-  database = "ccae",
-  outcome = c(
-    "Acute myocardial infarction",
-    rep("", 3),
-    "Heart failure (hosp)",
-    rep("", 3),
-    "Stroke (ischemic or hemorrhagic)",
-    rep("", 3)
-  ),
-  riskQuarter = rep(1:4, 3),
-  patientsAce = c(
-    161099, 204882, 214413, 204167,
-    146259, 188006, 218052, 230226,
-    146069, 187524, 217070, 226128
-  ),
-  personYearsAce = c(
-    276171, 372197, 393583, 351727,
-    249809, 341014, 399394, 400330,
-    294484, 340234, 397830, 393861
-  ),
-  eventsAce = c(
-    203, 534, 117, 2095,
-    228, 457, 826, 2012,
-    299, 554, 947, 1718
-  ),
-  patientsBeta = c(
-    133977, 90193, 80662, 90908,
-    126387, 84280, 83421, 98380,
-    126264, 84000, 83038, 97628
-  ),
-  personYearsBeta = c(
-    220633, 169231, 150035, 154419,
-    206706, 158425, 155222, 169139,
-    206453, 157913, 154587, 167810
-  ),
-  eventsBeta = c(
-    135, 321, 535, 1520,
-    378, 340, 570, 1773,
-    320, 351, 521, 1077
-  )
-)
+map_outcomes <- readRDS("data/processed/map_outcomes.rds")
+map_exposures <- readRDS("data/processed/map_exposures.rds")
 
-tableMdcr <- tibble(
-  database = "mdcr",
-  outcome = c(
-    "Acute myocardial infarction",
-    rep("", 3),
-    "Heart failure (hosp)",
-    rep("", 3),
-    "Stroke (ischemic or hemorrhagic)",
-    rep("", 3)
-  ),
-  riskQuarter = rep(1:4, 3),
-  patientsAce = c(
-    27853,
-    27596,
-    25893,
-    20319,
-    27530,
-    27486,
-    25482,
-    19704,
-    27291,
-    27054,
-    24763,
-    18666
-  ),
-  personYearsAce = c(
-    57541,
-    56910,
-    53202,
-    38710,
-    56886,
-    56644,
-    52475,
-    37842,
-    56413,
-    55846,
-    50976,
-    35775
-  ),
-  eventsAce = c(
-    231,
-    387,
[…]

### 2022-06-02 code/helpers/GenerateNcPlot.R

@@ -1,0 +1,106 @@
+generateNcPlot <- function(
+  logRr,
+  seLogRr,
+  trueLogRr,
+  legacy = FALSE,
+  model = NULL,
+  xLabel = "Relative risk",
+  title,
+  fileName = NULL
+) {
+  alpha <- 0.05
+  if (is.null(model)) {
+    model <- EmpiricalCalibration::fitSystematicErrorModel(logRr = logRr, 
+      seLogRr = seLogRr, 
+      trueLogRr = trueLogRr,
+      estimateCovarianceMatrix = FALSE,
+      legacy = legacy)
+  } else {
+    legacy <- (names(model)[3] == "logSdIntercept")  
+  }
+  d <- data.frame(logRr = logRr, 
+    seLogRr = seLogRr, 
+    trueLogRr = trueLogRr, 
+    trueRr = exp(trueLogRr),
+    logCi95lb = logRr + qnorm(0.025) * seLogRr,
+    logCi95ub = logRr + qnorm(0.975) * seLogRr)
+  d <- d[!is.na(d$logRr), ]
+  d <- d[!is.na(d$seLogRr), ]
+  if (nrow(d) == 0) {
+    return(NULL)
+  }
+  d$Group <- as.factor(d$trueRr)
+  d$Significant <- d$logCi95lb > d$trueLogRr | d$logCi95ub < d$trueLogRr
+  
+  temp1 <- aggregate(Significant ~ trueRr, data = d, length)
+  temp2 <- aggregate(Significant ~ trueRr, data = d, mean)
+  temp1$nLabel <- paste0(formatC(temp1$Significant, big.mark = ","), " estimates")
+  temp1$Significant <- NULL
+  temp2$meanLabel <- paste0(formatC(100 * (1 - temp2$Significant), digits = 1, format = "f"),
+    "% of CIs includes ",
+    temp2$trueRr)
+  temp2$Significant <- NULL
+  dd <- merge(temp1, temp2)
+  
+  breaks <- c(0.5, 0.75, 1, 1.25, 1.5, 1.75, 2)
+  theme <- ggplot2::element_text(colour = "#000000", size = 10)
+  themeRA <- ggplot2::element_text(colour = "#000000", size = 10, hjust = 1)
+  
+  d$Group <- paste("True", tolower(xLabel), "=", d$trueRr)
+  dd$Group <- paste("True", tolower(xLabel), "=", dd$trueRr)
+  
[…]

### 2022-06-28 submission/manuscript.rmd
update structure for resubmission

@@
   - Ewout W. Steyerberg, PhD$^{3,5}$
-  - David M. Kent, PhD$^2$
+  - David M. Kent, MD$^2$
   - Peter R. Rijnbeek, PhD${^1}$
 output:
-    bookdown::pdf_document2: default
+    bookdown::pdf_document2:
+      number_sections: false
     bookdown::word_document2:
@@
   - \usepackage[left, pagewise]{lineno}
-  - \usepackage{caption}
+  - \usepackage[labelfont=bf]{caption}
 editor_options: 
@@
     if (abs(x - round(x)) < .Machine$double.eps) {
-      formatted <- format(x, digits = 1, big.mark = " ")
+      formatted <- format(x, digits = 1, big.mark = ",")
     } else {
-      formatted <- format(x, digits = 1, nsmall = 2, big.mark = " ", decimal.mark = intToUtf8("0x00B7"))
+      formatted <- format(x, digits = 1, nsmall = 2, big.mark = ",", decimal.mark = ".")
     }
@@
 \singlespacing 
-**Background**: Treatment effects are often anticipated to vary across groups of
-patients with different baseline risk. Implementation of a risk-based approach
-in the RCT setting has shown evidence of increase in statistical power for the
-evaluation of treatment effect heterogeneity. The aim of this study was to
-extend this approach to the observational setting using a standardized scalable
-framework.
-
-**Methods**: The proposed framework consists of
-five steps: 1) definition of the problem, i.e. the population, the treatment,
-the comparator and the outcome(s) of interest; 2) identification of relevant
-databases; 3) development of a prediction model for the outcome(s) of interest;
-4) estimation of relative and absolute treatment effect within strata of
-predicted risk, after adjusting for observed confounding; 5) presentation of the
-results. We demonstrate our framework by evaluating heterogeneity of the effect
-of angiotensin-converting enzyme (ACE) inhibitors versus beta blockers on three
-efficacy and six safety outcomes across three observational databases.
-
-**Findings**: In our demostration, patients at low risk of acute myocardial
-infarction (MI) received negligible absolute benefits for all three efficacy
-outcomes, though they were more pronounced in the highest risk quarter,
-especially for hospitalization with heart failure. However, failing diagnostics
-showed evidence of residual imbalances even after propensity score adjustment.
-
-**Interpretation**: Application of our framework allows for
-observing differential risk of outcomes across strata, which offers the
-opportunity to consider the benefit-harm tradeoff between alternative treatments
-within targeted subpopulations. Further insights may arise by application to
-safety and effectiveness questions on a large scale.
-
-**Funding**: European Health Data and Evidence Network (EHDEN) project.
+Treatment effects are often anticipated to vary across groups of patients with
+different baseline risk. The Predictive Approaches to Treatment Effect
+Heterogeneity (PATH) statement focused on baseline risk as a robust predictor of
+treatment effect and provided guidance on risk-based assessment of treatment
+effect heterogeneity in the RCT setting. The aim of this study was to extend
+this approach to the observational setting using a standardised scalable
+framework. The proposed framework consists of five steps: 1) definition of the
+research aim, i.e., the population, the treatment, the comparator and the
+outcome(s) of interest; 2) identification of relevant databases; 3) development
+of a prediction model for the outcome(s) of interest; 4) estimation of relative
+and absolute treatment effect within strata of predicted risk, after adjusting
+for observed confounding; 5) presentation of the results. We demonstrate our
+framework by evaluating heterogeneity of the effect of angiotensin-converting
+enzyme (ACE) inhibitors versus beta blockers on three efficacy and six safety
+outcomes across three observational databases.  The proposed framework can
+supplement any comparative effectiveness study. We provide a publicly available
+R software package for applying this framework to any database mapped to the
+Observational Medical Outcomes Partnership Common Data Model. In our
+demonstration, patients at low risk of acute myocardial infarction received
+negligible absolute benefits for all three efficacy outcomes, though they were
+more pronounced in the highest risk quarter, especially for hospitalisation with
+heart failure. However, failing diagnostics showed evidence of residual
+imbalances even after adjustment for observed confounding. Our framework allows
+for the evaluation of differential treatment effects across risk strata, which
+offers the opportunity to consider the benefit-harm trade-off between
+alternative treatments. It is easily applicable and highly informative whenever
+treatment effect estimation in observational data is of interest.
 
@@
 
-# Introduction
+# INTRODUCTION
 
@@
 
-# Methods
-
+# RESULTS
 The proposed framework defines 5 distinct steps: 1) definition of the research
@@
 
-## Step 1: General definition of the research aim
-
-The typical research aim is: “to compare the effect of treatment to a comparator
-treatment in patients with disease with respect to outcomes $O_1,\dots,O_n$”.
-
-We use a comparative cohort design. This means that at least three cohorts of
-patients need to be defined at this stage of the framework:
-
-- A single treatment cohort ($T$) which includes patients with disease receiving
-  the target treatment of interest.
-- A single comparator cohort ($C$) which includes patients with disease receiving
-  the comparator treatment.
-- One or more outcome cohorts ($O_1,\dots,O_n$) that contain patients
-  developing the outcomes of interest
-
-## Step 2: Identification of the databases {#methods2}
-
-Including in our analyses multiple databases representing the population of
-interest potentially increases the generalizability of results. Furthermore, the
-cohorts should preferably have adequate sample size with adequate follow-up time
-to ensure precise effect estimation, even within smaller risk strata. Other
-relevant issues such as the depth of data capture (the precision at which
-measurements, lab tests, conditions are recorded) and the reliability of data
-entry should also be considered.
-
-## Step 3: Prediction {#methods3}
-
-Our method relies on adequately separating patients into subgroups based on
-their baseline risk for the outcomes of interest. Therefore, a model---either an
-existing external model adequately validated on an internally developed
-one---assigning patient-level risk is required. For internally developing a risk
-prediction model we adopt the standardized framework focused on observational
-data that ensures adherence to existing guidelines [@Reps2018; @Collins2015;
-@Moons2015].
-
-We first need to define a target cohort of patients, i.e. the set of patients on
-whom the prediction model will be developed. In our case, the target cohort is
-generated by pooling the already defined treatment and comparator cohorts. We
-develop the prediction model on the propensity score-matched (1:1) subset of the
-pooled sample to avoid differentially fitting between treatment arms, thus
-introducing spurious interactions with treatment
-[@Burke2014;@vanKlaveren2019]. We also need to define a set of patients that
-experience the outcome of interest, i.e. the outcome cohort. Finally, we need to
-decide the time frame within which the predictions will be carried out, i.e. the
-patients' time at risk. Subsequently, we can develop the prediction model.
-
-It is important that the prediction models display good discriminative ability
-to ensure that risk-based subgroups are accurately defined. A performance
-overview of the derived prediction models including discrimination and
-calibration both in the propensity score matched subset, the entire sample and
-separately for treated and comparator patients should also be reported.
-
-## Step 4: Estimation {#methods4}
-
-We estimate treatment effects (both on the relative and the absolute scale)
-within risk strata defined using the prediction model of step 3. We often
-consider four risk strata, but fewer or more strata can be considered depending on
-the available power for accurately estimating stratum-specific treatment
-effects. Effect estimation may be focused on the difference in outcomes for a
-randomly selected person from the risk stratum (average treatment effect) or for
-a randomly selected person from the treatment cohort within the risk stratum
-receiving the treatment under study (average treatment effect on the treated).
-
-Any appropriate method for the analysis of relative and absolute treatment
-effects can be considered, as long as the this is done consistently in all risk
-strata. Common statistical metrics are odds ratios or hazard ratios for relative
-scale estimates and differences in observed proportions or differences in
-Kaplan-Meier estimates for absolute scale estimates, depending on the problem at
-hand. We estimate propensity scores within risk strata which we then use to
-match patients from different treatment cohorts or to stratify them into groups
-with similar propensity scores or to weigh each patient’s contribution to the
-estimation process [@Austin2011].
-
-Prior to analyzing results, it is crucial to ensure that all diagnostics are passed
-in all risk strata. The standard diagnostics we carry out include analysis of
-the overlap of propensity score distributions and calculation of standardized
-mean differences of the covariates before and after propensity score
-adjustment. Finally, we use effect estimates for a large set of negative control
-outcomes (i.e. outcomes known to not be related with any of the exposures under
-study) to evaluate the presence of residual confounding not accounted for by
-propensity score adjustment [@Schuemie2014;@Schuemie2571;@Schuemie2018].
-
-## Step 5: Presentation of results {#methods5}
-
-In the presence of a positive treatment effect and a well-discriminating
-prediction model we expect an increasing pattern of the differences in the
-absolute scale, even if treatment effects remain constant on the relative scale
-across risk strata. Due to this scale-dependence of treatment effect
-heterogeneity, results should be assessed both on the relative and the absolute
-scale.
-
-## Case study
+```{r graphicalAbstract, cache=TRUE, echo=FALSE, fig.cap="Illustration of how the framework is applied on two observational databases, preferably mapped to OMOP-CDM.", fig.show="hold", out.width = '100%'}
+include_graphics(here("figures/rsee_process.png"))
+```
 
@@
 angioedema, cough and abnormal weight gain). We used data from three US-based
-claims databases. The analysis plan was the framework outlined in steps 1
-through 5.
-
-# Results
+claims databases.
 
@@
 patients initiating treatment with beta blockers in CCAE, MDCD and MDCR
-respectively (Table \@ref(tab:incidenceOverall)). Adequate numbers of patients were included in all strata
-of predicted acute MI risk (Supplement: Table XX).
+respectively (Table \@ref(tab:incidenceOverall)). Adequate numbers of patients
+were included in all strata of predicted acute MI risk (Supplement, page 26).
+
+```{r incidenceOverall, echo = FALSE}
+map_outcomes <- readRDS(here("data/processed/map_outcomes.rds"))
+map_exposures <- readRDS(here("data/processed/map_exposures.rds"))
+
+incidence <- readRDS(here::here("data/processed/incidenceOverall.rds")) %>%
+  left_join(map_outcomes, by = c("outcomeId" = "outcome_id")) %>%
+  select(-outcomeId) %>%
+  rename("outcomeId" = "outcome_name") %>%
+  mutate(
+    treatmentYears = round(treatmentDays / 365),
+    comparatorYears = round(comparatorDays / 365),
+    outcomeId =  factor(
+    outcomeId,
+    levels = c(
+      "acute_myocardial_infarction",
+      "hospitalization_with_heart_failure",
+      "stroke"
+    ),
+    labels = c(
+        "acute myocardial infarction",
+        "hospitalization with heart failure",
+        "stroke"
+      )
+    )
+  ) %>%
+  arrange(database, outcomeId)
+
+kableExtra::kbl(
+  incidence %>%
+    select(
+      outcomeId,
+      treatmentPersons,
+      treatmentYears,
+      treatmentOutcomes,
+      comparatorPersons,
+      comparatorYears,
+      comparatorPersons,
+      comparatorOutcomes
+    ),
+  align = "llrrrrr",
+  format = "latex",
+  booktabs = TRUE,
+  longtable = TRUE,
+  escape = FALSE,
+  caption = "Number of patients, person years and events within quarters of predicted risk for acute MI for the three efficacy outcomes of the study (acute MI, hospitalization with heart failure and ischemic or hemorrhagic stroke) across the three databases.",
+  linesep = c('', '', '', '\\addlinespace'),
+  col.names = c(
+    "Outcome",
+    "Patients",
+    "Person years",
+    "Outcomes",
+    "Patients",
+    "Person years",
+    "Outcomes"
+  ),
+  format.args = list(
+    big.mar      = ",",
+    decimal.mark = "."
+  )
+) %>%
+  kableExtra::add_header_above(
+    c(
+      " " = 1,
+      "ACE inhibitors" = 3,
+      "Beta blockers" = 3
+    )
+  ) %>%
+  kableExtra::group_rows("CCAE", 1, 3) %>%
+  kableExtra::group_rows("MDCD", 4, 6) %>%
+  kableExtra::group_rows("MDCR", 7, 9) %>%
+  kableExtra::kable_styling(font_size = 7)
+
+```
 
@@
 (1:1) subset of the sample, using caliper of 
-`r format(.2, nsmall = 1, decimal.mark = intToUtf8("0x00B7"))`
+`r format(.2, nsmall = 1, decimal.mark = ".")`
 and after excluding patients having the outcome any time prior to treatment
@@
 \@ref(tab:predictionPerformance)). We also observed lower performance of the
-prediction model developed in MDCR compared to the other 2 databases.
+prediction model developed in MDCR compared to the other 2 databases
+(Supplement, page 29).
+
+```{r predictionPerformance, echo = FALSE}
+predictionPerformance <- readRDS(here("data/processed/predictionPerformance.rds")) %>%
+  filter(
+    stratOutcome == 2
+  ) %>%
+  select(c("database", "cohort", "AUROC", "95% lower AUROC", "95% upper AUROC")) %>%
+  rename(
+    "lower" = "95% lower AUROC",
+    "upper" = "95% upper AUROC"
+  ) %>%
+  mutate(
+    auc = paste(
+      format(round(AUROC, 2), nsmall = 2), 
+      paste0(
+        "(",
+        format(round(lower, 2), nsmall = 2),
+        ", ",
+        format(round(upper, 2), nsmall = 2),
+        ")"
+      )
+    )
+  ) %>%
+  select(-c("AUROC", "lower", "upper")) %>%
+  mutate(
+    cohort = factor(
+      cohort,
+      levels = c(
+        "Matched",
+        "Treatment",
+        "Comparator",
+        "EntirePopulation"
+      ),
+      labels = c(
+        "Matched",
+        "Treatment",
+        "Comparator",
+        "Entire population"
+      )
+    ),
+    database = factor(
+      database,
+      levels = c(
+        "ccae",
+        "mdcd",
+        "mdcr"
+      )
+    )
+  ) %>%
+  arrange(database, cohort)
+
+kableExtra::kbl(
+  predictionPerformance %>% pivot_wider(id_cols = "cohort", names_from = "database", values_from = "auc"),
+  align = "lrrr",
+  format = "latex",
+  booktabs = TRUE,
+  longtable = TRUE,
+  escape = FALSE,
+  caption = "Discriminative ability (c-statistic) of the derived prediction models for acute myocardial infarction in the matched set (development set), the treatment cohort, the comparator cohort and the entire population in CCAE, MDCD and MDCR.",
+  linesep = c('', '', '', '\\addlinespace'),
+  col.names = c(
+    "Population",
+    "CCAE",
+    "MDCD",
+    "MDCR"
+  ),
+  format.args = list(
+    big.mar      = ","
+  )
+) %>%
+  kableExtra::kable_styling(font_size = 7)
+```
 
@@
 of 
-`r format(.1, nsmall = 1, decimal.mark = intToUtf8("0x00B7"))`
+`r format(.1, nsmall = 1, decimal.mark = ".")`
 is violated (Figure \ref{fig:covariateBalance}). This is more pronounced in the
@@
 inhibitors in this condition. Analyses in MDCR passed all diagnostics.
+
+```{r psDensity, cache=TRUE, echo=FALSE, fig.cap="\\textbf{Preference score distributions in quarters of predicted acute MI risk}.
+The preference score is a transformation of the propensity score that adjusts for prevalence differences between populations.
+Higher overlap of the preference score distributions indicates that patients in the target and the comparator cohorts are more similar in terms of the predicted probability of receiving treatment (ACE inhibitors).", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/PsDensity.tiff")))
+```
+
+```{r covariateBalance, cache=TRUE, echo=FALSE, fig.cap="\\textbf{Patient characteristic balance for ACE inhibitors and beta blockers before and after stratification on the propensity scores}.
+Each dot represents the standardized difference of means for a single covariate before (x-axis) and after (y-axis) stratification.
+A commonly used rule of thumb suggests that standardized mean differences above 0.1 indicate insufficient covariate balance post propensity score adjustment.", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/CovariateBalance.tiff")))
+```
 
@@
 treatment arms. This was not the case in the other risk quarters of CCAE, or in
-any risk quarter of MDCD and MDCR (Supplement, Figure XX).
+any risk quarter of MDCD and MDCR (Supplement, pages 34-36).
+
+```{r negativeControls, cache=TRUE, echo=FALSE, fig.cap="\\textbf{Systematic error}.
+Effect size estimates for the negative controls (true hazard ratio = 1).
+Estimates below the diagonal dashed lines are statistically significant (alpha = 0.05) different from the true effect size.
+A well-calibrated estimator should have the true effect size within the 95 percent confidence interval 95 percent of times.", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/NegativeControlsPlot_ccae.tiff")))
+```
 
@@
 ```{r, echo=FALSE}
-map_outcomes <- readRDS(here("data/processed/map_outcomes.rds"))
-
-map_exposures <- readRDS(here("data/processed/map_exposures.rds"))
 overall <- readRDS(here("data/processed/mappedOverallResults.rds")) %>%
@@
     estimate = paste0(
-      format(round(estimate, 2), nsmall = 2, decimal.mark = intToUtf8("0x00B7")),
+      format(round(estimate, 2), nsmall = 2),
       " (", 
-      format(round(lower, 2), nsmall = 2, decimal.mark = intToUtf8("0x00B7")),
+      format(round(lower, 2), nsmall = 2),
       ", ", 
-      format(round(upper, 2), nsmall = 2, decimal.mark = intToUtf8("0x00B7")),
+      format(round(upper, 2), nsmall = 2),
       ")"
@@
 `r acute_mi_relative %>% filter(database == "mdcd", riskStratum == "Q4") %>% pull(upper)`;
-95\% CI) in CCAE and MDCD respectively (Figure 4). In MDCR hazard ratios
+95\% CI) in CCAE and MDCD respectively (Figure \ref{fig:relative}). In MDCR hazard ratios
 increased from
@@
 
+```{r overallTable, echo = FALSE}
+kableExtra::kbl(
+  overall %>% 
+    pivot_wider(id_cols = "outcome", names_from = "database", values_from = "estimate"),
+  format = "latex",
+  align = "lrrr",
+  caption = "Overall hazard ratios.",
+  booktabs = TRUE,
+  col.names = c(
+    "Outcome",
+    "CCAE",
+    "MDCD",
+    "MDCR"
+  )) %>%
+  kableExtra::kable_styling(latex_options = "HOLD_position")
+```
+
+```{r relative, cache=TRUE, echo=FALSE, fig.cap="\\textbf{Overview of heterogeneity of ACE inhibitors treatment on the relative scale (hazard ratios) within strata of predicted risk of acute MI}.
+Values below 1 favor ACE inhibitors, while values above 1 favor beta blockers.", fig.show="hold", out.width = '100%'}
+    grid::grid.raster(tiff::readTIFF(here("figures/CombinedRelative_acute_myocardial_infarction.tiff")))
+```
+
 We observed an increasing trend of treatment effect on the absolute scale with
@@
 95\% CI) in the highest acute MI risk quarter in CCAE, MDCD, and MDCR,
-respectively (Figure 5). We found no difference on the absolute scale for stroke
-across risk strata. Absolute risk differences did not favor ACE inhibitors
-compared to beta blockers in terms of cough, even though this effect again
-diminished with increasing acute MI risk---from 
+respectively (Figure \ref{fig:absolute}). We found no difference on the absolute
+scale for stroke across risk strata. Absolute risk differences did not favor ACE
+inhibitors compared to beta blockers in terms of cough, even though this effect
+again diminished with increasing acute MI risk---from
 `r acute_cough_absolute %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(estimate)`\%
@@
 
+```{r absolute, cache=TRUE, echo=FALSE, fig.cap="\\textbf{Overview of heterogeneity of ACE inhibitors treatment on the absolute scale within strata of predicted risk of acute MI}.
+Estimates of absolute treatment effect are derived as the difference in Kaplan-Meier estimates at two years after inclusion.
+Values above 0 favor ACE inhibitors, while values below 0 favor beta blockers.
+", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/CombinedAbsolute_acute_myocardial_infarction.tiff")))
+```
+
 The results of all the analyses performed can be accessed and assessed through a
@@
 
-# Discussion
+# DISCUSSION
 
@@
 
+# METHODS
+
+## Step 1: General definition of the research aim
+
+The typical research aim is: “to compare the effect of treatment to a comparator
+treatment in patients with disease with respect to outcomes $O_1,\dots,O_n$”.
+
+We use a comparative cohort design. This means that at least three cohorts of
+patients need to be defined at this stage of the framework:
+
+- A single treatment cohort ($T$) which includes patients with disease receiving
+  the target treatment of interest.
+- A single comparator cohort ($C$) which includes patients with disease receiving
+  the comparator treatment.
+- One or more outcome cohorts ($O_1,\dots,O_n$) that contain patients
+  developing the outcomes of interest
+
+## Step 2: Identification of the databases {#methods2}
+
+Including in our analyses multiple databases representing the population of
+interest potentially increases the generalizability of results. Furthermore, the
+cohorts should preferably have adequate sample size with adequate follow-up time
+to ensure precise effect estimation, even within smaller risk strata. Other
+relevant issues such as the depth of data capture (the precision at which
+measurements, lab tests, conditions are recorded) and the reliability of data
+entry should also be considered.
+
+## Step 3: Prediction {#methods3}
+
+Our method relies on adequately separating patients into subgroups based on
+their baseline risk for the outcomes of interest. Therefore, a model---either an
+existing external model adequately validated on an internally developed
+one---assigning patient-level risk is required. For internally developing a risk
+prediction model we adopt the standardized framework focused on observational
+data that ensures adherence to existing guidelines [@Reps2018; @Collins2015;
+@Moons2015].
+
+We first need to define a target cohort of patients, i.e. the set of patients on
+whom the prediction model will be developed. In our case, the target cohort is
+generated by pooling the already defined treatment and comparator cohorts. We
+develop the prediction model on the propensity score-matched (1:1) subset of the
+pooled sample to avoid differentially fitting between treatment arms, thus
+introducing spurious interactions with treatment
+[@Burke2014;@vanKlaveren2019]. We also need to define a set of patients that
+experience the outcome of interest, i.e. the outcome cohort. Finally, we need to
+decide the time frame within which the predictions will be carried out, i.e. the
+patients' time at risk. Subsequently, we can develop the prediction model.
+
+It is important that the prediction models display good discriminative ability
+to ensure that risk-based subgroups are accurately defined. A performance
+overview of the derived prediction models including discrimination and
+calibration both in the propensity score matched subset, the entire sample and
+separately for treated and comparator patients should also be reported.
+
+## Step 4: Estimation {#methods4}
+
+We estimate treatment effects (both on the relative and the absolute scale)
+within risk strata defined using the prediction model of step 3. We often
+consider four risk strata, but fewer or more strata can be considered depending on
+the available power for accurately estimating stratum-specific treatment
+effects. Effect estimation may be focused on the difference in outcomes for a
+randomly selected person from the risk stratum (average treatment effect) or for
+a randomly selected person from the treatment cohort within the risk stratum
+receiving the treatment under study (average treatment effect on the treated).
+
+Any appropriate method for the analysis of relative and absolute treatment
+effects can be considered, as long as the this is done consistently in all risk
+strata. Common statistical metrics are odds ratios or hazard ratios for relative
+scale estimates and differences in observed proportions or differences in
+Kaplan-Meier estimates for absolute scale estimates, depending on the problem at
+hand. We estimate propensity scores within risk strata which we then use to
+match patients from different treatment cohorts or to stratify them into groups
+with similar propensity scores or to weigh each patient’s contribution to the
+estimation process [@Austin2011].
+
+Prior to analyzing results, it is crucial to ensure that all diagnostics are passed
+in all risk strata. The standard diagnostics we carry out include analysis of
+the overlap of propensity score distributions and calculation of standardized
+mean differences of the covariates before and after propensity score
+adjustment. Finally, we use effect estimates for a large set of negative control
+outcomes (i.e. outcomes known to not be related with any of the exposures under
+study) to evaluate the presence of residual confounding not accounted for by
+propensity score adjustment [@Schuemie2014;@Schuemie2571;@Schuemie2018].
+
+## Step 5: Presentation of results {#methods5}
+
+In the presence of a positive treatment effect and a well-discriminating
+prediction model we expect an increasing pattern of the differences in the
+absolute scale, even if treatment effects remain constant on the relative scale
+across risk strata. Due to this scale-dependence of treatment effect
+heterogeneity, results should be assessed both on the relative and the absolute
+scale.
+
+# DATA AVAILABILITY
+The claims data are proprietary and are not publicly accessible due to
+restricted user agreement. Database descriptions are available in the
+Supplementary Material.
+
+# CODE AVAILABILITY
+Our analysis code is publicly available under the MIT license at
+https://github.com/mi-erasmusmc/AceBeta9Outcomes.
+
 \newpage
-# References
+# REFERENCES
 \nolinenumbers
@@
 
-# Tables and figures
+# AUTHOR CONTRIBUTIONS
+AR, PRR, and DVK conceptualised the study. AR, DVK, EWS, and DMK developed the
+methodology. PBR, and PRR acquired the data and AR analysed the data. AR
+developed the software and wrote drafted the manuscript, which was critically
+reviewed by DVK, PBR, EWS, DMK, and PRR. AR, PBR, and PRR had full access to the
+raw data. All authors read and approved the manuscript and had final
+responsibility for the decision to submit for publication.
 
-```{r incidenceOverall, echo = FALSE}
-incidence <- readRDS(here::here("data/processed/incidenceOverall.rds")) %>%
-  left_join(map_outcomes, by = c("outcomeId" = "outcome_id")) %>%
-  select(-outcomeId) %>%
-  rename("outcomeId" = "outcome_name") %>%
-  mutate(
-    treatmentYears = round(treatmentDays / 365),
-    comparatorYears = round(comparatorDays / 365),
-    outcomeId =  factor(
-    outcomeId,
-    levels = c(
-      "acute_myocardial_infarction",
-      "hospitalization_with_heart_failure",
-      "stroke"
-    ),
-    labels = c(
-        "acute myocardial infarction",
-        "hospitalization with heart failure",
-        "stroke"
-      )
-    )
-  ) %>%
-  arrange(database, outcomeId)
-
-kableExtra::kbl(
-  incidence %>%
-    select(
-      outcomeId,
-      treatmentPersons,
-      treatmentYears,
-      treatmentOutcomes,
-      comparatorPersons,
-      comparatorYears,
-      comparatorPersons,
-      comparatorOutcomes
-    ),
-  align = "llrrrrr",
-  format = "latex",
-  booktabs = TRUE,
-  longtable = TRUE,
-  escape = FALSE,
-  caption = "Number of patients, person years and events within quarters of predicted risk for acute MI for the three efficacy outcomes of the study (acute MI, hospitalization with heart failure and ischemic or hemorrhagic stroke) across the three databases.",
-  linesep = c('', '', '', '\\addlinespace'),
-  col.names = c(
-    "Outcome",
-    "Patients",
-    "Person years",
-    "Outcomes",
-    "Patients",
-    "Person years",
-    "Outcomes"
-  ),
-  format.args = list(
-    big.mar      = " ",
-    decimal.mark = "."
-  )
-) %>%
-  kableExtra::add_header_above(
-    c(
-      " " = 1,
-      "ACE inhibitors" = 3,
-      "Beta blockers" = 3
-    )
-  ) %>%
-  kableExtra::group_rows("CCAE", 1, 3) %>%
-  kableExtra::group_rows("MDCD", 4, 6) %>%
-  kableExtra::group_rows("MDCR", 7, 9) %>%
-  kableExtra::kable_styling(font_size = 7)
-
-```
-
-\newpage
-
-```{r predictionPerformance, echo = FALSE}
-predictionPerformance <- readRDS(here("data/processed/predictionPerformance.rds")) %>%
-  filter(
-    stratOutcome == 2
-  ) %>%
-  select(c("database", "cohort", "AUROC", "95% lower AUROC", "95% upper AUROC")) %>%
-  rename(
-    "lower" = "95% lower AUROC",
-    "upper" = "95% upper AUROC"
-  ) %>%
-  mutate(
-    auc = paste(
-      format(round(AUROC, 2), nsmall = 2, decimal.mark = intToUtf8("0x00B7")),
-      paste0(
-        "(",
-        format(round(lower, 2), nsmall = 2, decimal.mark = intToUtf8("0x00B7")),
-        ", ",
-        format(round(upper, 2), nsmall = 2, decimal.mark = intToUtf8("0x00B7")),
-        ")"
-      )
-    )
-  ) %>%
-  select(-c("AUROC", "lower", "upper")) %>%
-  mutate(
-    cohort = factor(
-      cohort,
-      levels = c(
-        "Matched",
-        "Treatment",
-        "Comparator",
-        "EntirePopulation"
-      ),
-      labels = c(
-        "Matched",
-        "Treatment",
-        "Comparator",
-        "Entire population"
-      )
-    ),
-    database = factor(
-      database,
-      levels = c(
-        "ccae",
-        "mdcd",
-        "mdcr"
-      )
-    )
-  ) %>%
-  arrange(database, cohort)
-
-kableExtra::kbl(
-  predictionPerformance %>% pivot_wider(id_cols = "cohort", names_from = "database", values_from = "auc"),
-  align = "lrrr",
-  format = "latex",
-  booktabs = TRUE,
-  longtable = TRUE,
-  escape = FALSE,
-  caption = "Discriminative ability (c-statistic) of the derived prediction models for acute myocardial infarction in the matched set (development set), the treatment cohort, the comparator cohort and the entire population in CCAE, MDCD and MDCR.",
-  linesep = c('', '', '', '\\addlinespace'),
-  col.names = c(
-    "Population",
-    "CCAE",
-    "MDCD",
-    "MDCR"
-  ),
-  format.args = list(
-    big.mar      = ","
-  )
-) %>%
-  kableExtra::kable_styling(font_size = 7)
-```
-
-\newpage
-
-```{r overallTable, echo = FALSE}
-kableExtra::kbl(
-  overall %>% 
-    pivot_wider(id_cols = "outcome", names_from = "database", values_from = "estimate"),
-  format = "latex",
-  align = "lrrr",
-  caption = "Overall hazard ratios.",
-  booktabs = TRUE,
-  col.names = c(
-    "Outcome",
-    "CCAE",
-    "MDCD",
-    "MDCR"
-  ),
-  format.args = list(decimal.mark = intToUtf8("0x00B7"))
-) %>%
-  kableExtra::kable_styling(latex_options = "HOLD_position")
-```
-
-\newpage
-
-```{r graphicalAbstract, cache=TRUE, echo=FALSE, fig.cap="Illustration of how the framework is applied on two observational databases, preferably mapped to OMOP-CDM.", fig.show="hold", out.width = '100%'}
-include_graphics(here("figures/graphicalAbstract.png"))
-```
-
-\newpage
-
-```{r psDensity, cache=TRUE, echo=FALSE, fig.cap="Preference score distributions for the evaluation of heterogeneity of the effect of ACE inhibitors compared to beta blockers on acute MI based on quarters of predicted acute MI risk.
-The preference score is a transformation of the propensity score that adjusts for prevalence differences between populations.", fig.show="hold", out.width = '100%'}
-grid::grid.raster(tiff::readTIFF(here("figures/PsDensity.tiff")))
-```
-
-\newpage
-
-```{r covariateBalance, cache=TRUE, echo=FALSE, fig.cap="Patient characteristic balance for ACE inhibitors and beta blockers before and after stratification on the propensity scores.
-Each dot represents the standardized difference of means for a single covariate before (x-axis) and after (y-axis) stratification.", fig.show="hold", out.width = '100%'}
-grid::grid.raster(tiff::readTIFF(here("figures/CovariateBalance.tiff")))
-```
-
-\newpage
-
-```{r negativeControls, cache=TRUE, echo=FALSE, fig.cap=" Systematic error.
-Effect size estimates for the negative controls (true hazard ratio = 1).
-Estimates below the diagonal dashed lines are statistically significant (alpha = 0.05) different from the true effect size.
-A well-calibrated estimator should have the true effect size within the 95 percent confidence interval 95 percent of times.", fig.show="hold", out.width = '100%'}
-grid::grid.raster(tiff::readTIFF(here("figures/NegativeControlsPlot_ccae.tiff")))
-```
-
-\newpage
-
-```{r relative, cache=TRUE, echo=FALSE, fig.cap="Overview of heterogeneity of ACE inhibitors treatment on the relative scale (hazard ratios) within strata of predicted risk of acute MI.
-Values below 1 favor ACE inhibitors, while values above 1 favor beta blockers.", fig.show="hold", out.width = '100%'}
-    grid::grid.raster(tiff::readTIFF(here("figures/CombinedRelative_acute_myocardial_infarction.tiff")))
-```
-
-\newpage
-
-```{r absolute, cache=TRUE, echo=FALSE, fig.cap="Overview of heterogeneity of ACE inhibitors treatment on the absolute scale within strata of predicted risk of acute MI.
-Estimates of absolute treatment effect are derived as the difference in Kaplan-Meier estimates at two years after inclusion.
-Values above 0 favor ACE inhibitors, while values below 0 favor beta blockers.
-", fig.show="hold", out.width = '100%'}
-grid::grid.raster(tiff::readTIFF(here("figures/CombinedAbsolute_acute_myocardial_infarction.tiff")))
-```
+# COMPETING INTERESTS
+AR and PRR work for a group that received unconditional research grants from
+Boehringer-Ingelheim, GSK, Janssen Research & Development, Novartis, Pfizer,
+Yamanouchi, and Servier. None of these grants result in a conflict of interest
+for the content of this paper. PBR is an employee of Janssen R&D, subsidiary of
+Johnson & Johnson. DVK, DMK, EWS have nothing to declare.

### 2022-09-18 submission/manuscript.rmd
latest updates for submission

@@
       number_sections: false
+      keep_tex: true
     bookdown::word_document2:
@@
 Innovative Medicines Initiative 2 Joint Undertaking (JU) under grant agreement
-No 8069six. The JU receives support from the European Union’s Horizon 2020
+No 806968. The JU receives support from the European Union’s Horizon 2020
 research and innovation programme and EFPIA.
@@
 
-```{r graphicalAbstract, cache=TRUE, echo=FALSE, fig.cap="Illustration of how the framework is applied on two observational databases, preferably mapped to OMOP-CDM.", fig.show="hold", out.width = '100%'}
+```{r graphicalAbstract, cache=TRUE, echo=FALSE, fig.cap="\\textbf{Framework execution diagram}.
+Illustration of how the framework is applied on two observational databases, preferably mapped to OMOP-CDM.", fig.show="hold", out.width = '100%'}
 include_graphics(here("figures/rsee_process.png"))
@@
 respectively (Table \@ref(tab:incidenceOverall)). Adequate numbers of patients
-were included in all strata of predicted acute MI risk (Supplement, page 26).
+were included in all strata of predicted acute MI risk (Supplementary Tables
+S1-S3).
 
@@
 \@ref(tab:predictionPerformance)). We also observed lower performance of the
-prediction model developed in MDCR compared to the other 2 databases
-(Supplement, page 29).
+prediction model developed in MDCR compared to the other 2 databases.
 
@@
 treatment arms. This was not the case in the other risk quarters of CCAE, or in
-any risk quarter of MDCD and MDCR (Supplement, pages 34-36).
+any risk quarter of MDCD and MDCR (Supplementary Figures S1-S2).
 
@@
 # CODE AVAILABILITY
-Our analysis code is publicly available under the MIT license at
+The source code for the R-package that implements our framework can be found at
+https://github.com/OHDSI/RiskStratifiedEstimation. The code for implementing the
+proof of concept study presented here is publicly available at
 https://github.com/mi-erasmusmc/AceBeta9Outcomes.
@@
 
+# ACKNOWLEDGEMENTS
+AR and PRR have received funding from the Innovative Medicines Initiative 2
+Joint Undertaking (JU) under grant agreement No 80696. The JU receives support
+from the European Union’s Horizon 2020 research and innovation programme and
+EFPIA.
+
 # AUTHOR CONTRIBUTIONS

### 2022-09-18 code/CombinedAbsolutePlots.R

@@ -63,15 +63,10 @@ absolute <- prepareDataset(mappedOverallAbsoluteResults) %>%
     upper = 100 * upper,
     estOutcome = stringr::str_replace_all(estOutcome, "_", " "),
     stratOutcome = stringr::str_replace_all(stratOutcome, "_", " "),
-    estOutcome = stringr::str_replace_all(estOutcome, "hospitalization", "hospitalisation"),
-    estOutcome = stringr::str_replace_all(estOutcome, "emia", "aemia"),
-    stratOutcome = stringr::str_replace_all(stratOutcome, "_", " "),
-    stratOutcome = stringr::str_replace_all(stratOutcome, "hospitalization", "hospitalisation"),
-    stratOutcome = stringr::str_replace_all(stratOutcome, "emia", "aemia"),
     outcomeType = ifelse(
       estOutcome %in% c(
         "acute myocardial infarction",
-        "hospitalisation with heart failure",
+        "hospitalization with heart failure",
         "stroke"
       ),
       "Main outcomes",
@@ -96,8 +91,7 @@ absolutePlot <- ggplot(
   scale_y_continuous(
     # trans = "log10",
     # limits = c(.6, 4.04),
-    name = "Absolute risk reduction (%)",
-    labels = comma_format(decimal.mark = intToUtf8("0x00B7"))
+    name = "Absolute risk reduction (%)"
   ) +
   xlab("Risk quarter") +
   facet_grid(outcomeType ~ database, scales = "free") +
@@ -118,13 +112,13 @@ absolutePlot <- ggplot(
     ),
     breaks = c(
       "acute myocardial infarction",
-      "hospitalisation with heart failure",
+      "hospitalization with heart failure",
       "stroke",
       "abnormal weight gain",
       "angioedema",
       "cough",
-      "hyperkalaemia",
-      "hypokalaemia",
+      "hyperkalemia",
+      "hypokalemia",
       "hypotension"
     )
   ) +

### 2022-09-18 code/CombinedRelativePlots.R

@@ -61,15 +61,10 @@ relative <- prepareDataset(mappedOverallRelativeResults) %>%
     database = toupper(database),
     estOutcome = stringr::str_replace_all(estOutcome, "_", " "),
     stratOutcome = stringr::str_replace_all(stratOutcome, "_", " "),
-    estOutcome = stringr::str_replace_all(estOutcome, "hospitalization", "hospitalisation"),
-    estOutcome = stringr::str_replace_all(estOutcome, "emia", "aemia"),
-    stratOutcome = stringr::str_replace_all(stratOutcome, "_", " "),
-    stratOutcome = stringr::str_replace_all(stratOutcome, "hospitalization", "hospitalisation"),
-    stratOutcome = stringr::str_replace_all(stratOutcome, "emia", "aemia"),
     outcomeType = ifelse(
       estOutcome %in% c(
         "acute myocardial infarction",
-        "hospitalisation with heart failure",
+        "hospitalization with heart failure",
         "stroke"
       ),
       "Main outcomes",
@@ -85,8 +80,7 @@ if (outcome == "hospitalization_with_heart_failure") {
     trans = "log10",
     breaks = c(.5, .7, 1, 2, 5),
     limits = c(.45, 7.5),
-    name = "Hazard ratio",
-    labels = comma_format(decimal.mark = intToUtf8("0x00B7"))
+    name = "Hazard ratio"
   )
 } else if (outcome == "stroke") {
   scale_y <- scale_y_continuous(
@@ -101,8 +95,7 @@ if (outcome == "hospitalization_with_heart_failure") {
     trans = "log10",
     breaks = c(.5, .7, 1, 2, 5),
     limits = c(.55, 7.5),
-    name = "Hazard ratio",
-    labels = comma_format(decimal.mark = intToUtf8("0x00B7"))
+    name = "Hazard ratio"
   )
 }
 
@@ -140,13 +133,13 @@ relativePlot <- ggplot(
     ),
     breaks = c(
       "acute myocardial infarction",
-      "hospitalisation with heart failure",
+      "hospitalization with heart failure",
       "stroke",
       "abnormal weight gain",
       "angioedema",
       "cough",
-      "hyperkalaemia",
-      "hypokalaemia",
+      "hyperkalemia",
+      "hypokalemia",
       "hypotension"
     )
   ) +

### 2022-09-18 code/CovariateBalancePlot.R

@@ -3,6 +3,8 @@
 library(tidyverse)
 library(scales)
 
+source("code/helpers/MainCovariateBalancePlot.R")
+
 args <- commandArgs(trailingOnly = TRUE)
 fileType <- as.character(args[1])
 
@@ -20,12 +22,10 @@ for (i in seq_along(databases)) {
     readRDS() %>%
     bind_rows() %>%
     tibble() %>%	
-    CohortMethod::plotCovariateBalanceScatterPlot(
+    plotCovariateBalanceScatterPlot2(
       beforeLabel = "",
-      afterLabel = toupper(databases[i])
-    ) +
-    scale_x_continuous(
-      labels = comma_format(decimal.mark = intToUtf8("0x00B7"))
+      afterLabel = toupper(databases[i]),
+      limits = c(0, .6)
     ) +
     ggplot2::facet_grid(
       ~riskStratum

### 2022-09-18 code/PsDensityPlot.R

@@ -45,8 +45,7 @@ for (i in seq_along(databases)) {
       )
     ) +
     scale_x_continuous(
-      breaks = seq(0, 1, .5),
-      labels = comma_format(decimal.mark = intToUtf8("0x00B7"))
+      breaks = seq(0, 1, .5)
     ) +
     ggplot2::facet_grid(~riskStratum) +
     ggplot2::ylab(

### 2022-09-18 code/helpers/GenerateNcPlot.R

@@ -78,10 +78,9 @@ generateNcPlot <- function(
     ggplot2::geom_hline(yintercept = 0) +
     ggplot2::geom_text(x = log(0.52), y = 0.4, alpha = 1, hjust = "left", ggplot2::aes(label = .data$nLabel), size = 12.5, data = dd) +
     ggplot2::geom_text(x = log(0.52), y = 0.37, alpha = 1, hjust = "left", ggplot2::aes(label = .data$meanLabel), size = 12.5, data = dd) +
-      ggplot2::scale_x_continuous(xLabel, limits = log(c(0.5, 2)), breaks = log(breaks), labels = format(signif(breaks, 3), nsmall = 2, decimal.mark = intToUtf8("0x00B7")))+
+      ggplot2::scale_x_continuous(xLabel, limits = log(c(0.5, 2)), breaks = log(breaks), labels = format(signif(breaks, 3), nsmall = 2)) +
     ggplot2::scale_y_continuous(
-      "Standard Error",
-      labels = scales::comma_format(decimal.mark = intToUtf8("0x00B7"))) +
+      "Standard Error") +
     ggplot2::coord_cartesian(ylim = c(0, .4)) +
     ggplot2::facet_grid(. ~ Group) +
     ggplot2::theme(panel.grid.minor = ggplot2::element_blank(),

### 2022-09-18 code/helpers/MainCovariateBalancePlot.R

@@ -1,0 +1,54 @@
+plotCovariateBalanceScatterPlot2 <- function(balance,
+                                            absolute = TRUE,
+                                            threshold = 0,
+                                            limits,
+                                            title = "Standardized difference of mean",
+                                            fileName = NULL,
+                                            beforeLabel = "Before matching",
+                                            afterLabel = "After matching",
+                                            showCovariateCountLabel = FALSE,
+                                            showMaxLabel = FALSE) {
+  beforeLabel <- as.character(beforeLabel)
+  afterLabel <- as.character(afterLabel)
+  if (absolute) {
+    balance$beforeMatchingStdDiff <- abs(balance$beforeMatchingStdDiff)
+    balance$afterMatchingStdDiff <- abs(balance$afterMatchingStdDiff)
+  }
+  plot <- ggplot2::ggplot(balance,
+                          ggplot2::aes(x = .data$beforeMatchingStdDiff, y = .data$afterMatchingStdDiff)) +
+    ggplot2::geom_point(color = rgb(0, 0, 0.8, alpha = 0.3), shape = 16) +
+    ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+    ggplot2::geom_hline(yintercept = 0) +
+    ggplot2::geom_vline(xintercept = 0) +
+    ggplot2::ggtitle(title) +
+    ggplot2::scale_x_continuous(beforeLabel, limits = limits) +
+    ggplot2::scale_y_continuous(afterLabel, limits = limits)
+  if (threshold != 0) {
+    plot <- plot + ggplot2::geom_hline(yintercept = c(threshold,
+                                                      -threshold), alpha = 0.5, linetype = "dotted")
+  }
+  if (showCovariateCountLabel || showMaxLabel) {
+    labels <- c()
+    if (showCovariateCountLabel) {
+      labels <- c(labels, sprintf("Number of covariates: %s", format(nrow(balance), big.mark = ",", scientific = FALSE)))
+    }
+    if (showMaxLabel) {
+      labels <- c(labels, sprintf("%s max(absolute): %.2f", afterLabel, max(abs(balance$afterMatchingStdDiff), na.rm = TRUE)))
+    }
+    dummy <- data.frame(text = paste(labels, collapse = "\n"))
[…]
